---
title: "Getting Set Up With debiasR"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette shows how to load `debiasR`, access the empirical example data,
and check that the OD matrices have valid support before using them for method
demonstrations.

## Package Setup

Install `debiasR` from the project repository or load it from the current
checkout during development. The empirical examples use the optional
`debiasRdata` package, available from the DEBIAS GitHub organisation:

```r
remotes::install_github("de-bias/debiasRdata")
```

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
load_debiasr_workshop()
```

## Load Empirical Example Data

The vignettes use `debiasR_example_data()` so all inputs are normalised to the
same package schema. The main demonstrations request complete OD support:
matching origin and destination area sets, no duplicated OD pairs, nonnegative
flows, and zero-filled absent pairs.

```{r}
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)

mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od
coverage <- example_data$coverage
covariates <- example_data$covariates
distance <- example_data$distance
```

## Inspect Inputs

```{r}
example_data$metadata
```

```{r}
example_data$od_audit
```

```{r}
tibble::tibble(
  table = c(
    "mpd_od",
    "benchmark_od",
    "coverage",
    "covariates",
    "distance"
  ),
  rows = c(
    nrow(mpd_od),
    nrow(benchmark_od),
    nrow(coverage),
    nrow(covariates),
    nrow(distance)
  ),
  columns = c(
    ncol(mpd_od),
    ncol(benchmark_od),
    ncol(coverage),
    ncol(covariates),
    ncol(distance)
  )
)
```

## First Rows

```{r}
head(mpd_od)
head(benchmark_od)
head(coverage)
head(covariates)
```

The data are described in detail in `08-data.qmd`.
