---
title: "Getting started with debiasR"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette introduces the example data used throughout the `debiasR`
vignettes and shows the first checks needed before measuring bias, adjusting OD flows and validating results.

A typical `debiasR` workflow starts by:

- loading the package and necessary data
- identifying the population count table, OD-flow table and optional benchmark flow table
- checking that the observed and benchmark population and OD-flow tables use compatible areas and valid poulation and flow values

## Loading the package

The previous vignette showed how to install `debiasR` and the optional `debiasRdata` package. Here we load the packages and prepare the shared setup used by the remaining vignettes.

```{r setup}
library(debiasR)
library(debiasRdata)
```

## Example data: Locomizer travel-to-work flows

The vignettes use an empirical example at Local Authority District (LAD) level, combining mobile-phone mobility data provided by Locomizer with benchmark data from the 2021 UK Census. Locomizer is a spatial analytics company that processes anonymised mobile phone location data to produce mobility-related insights. In this example, the observed OD-flow table represents travel-to-work flows inferred from those mobile phone records, while the benchmark OD-flow table comes from the 2021 Census travel-to-work data.

The first step is to identify the core inputs for the `debiasR` workflow, i.e. the observed OD-flow table, the benchmark OD-flow table and the area-level coverage table used to measure population coverage bias. Here we load these directly from the `debiasRdata` package.

```{r}
OD_travel2work <- debiasRdata::lad_OD_travel2work 
census_OD_travel2work <- debiasRdata::census_lad_OD_travel2work
coverage <- debiasRdata::coverage_lad
```

The example data available via `debiasRdata` also includes supporting inputs used in later vignettes. Particularly, `covariates`, which contains area-level characteristics used to explain variation in bias across places, and `distance`, which contains distance between origin and destination areas, used by flow adjustment models.

```{r}
covariates <- debiasRdata::lad_covariates
distance <- debiasRdata::lad_centroids
```

In summary, the main input datasets are:

| Object | Role |
|---|---|
| `OD_travel2work` | observed mobile-phone-derived OD flows |
| `census_OD_travel2work` | benchmark OD flows from the Census |
| `coverage` | area-level benchmark and observed digital trace population counts |
| `covariates` | area-level characteristics used to explain bias |
| `distance` | OD distance used in adjustment models |


## Inspect inputs

Before proceeding to the next steps of the workflow, it is useful to check and understand what each input dataset contains and whether the inputs are valid. 

Below, we produce a compact row and column count, which gives a quick overview of the size of the main datasets. If the workflow includes $N$ areas and within-area flows are included, a complete OD-flow table has $N \times N$ rows: one row for every possible origin-destination pair. If within-area flows are excluded, the complete table has $N \times (N - 1)$ rows. If the table only records observed flows, it can have fewer rows, because OD pairs with no recorded flow may be absent. The `coverage` table should have one row per area, so it should contain $N$ rows.

For the supporting tables, `covariates` table should also have one row per area, with columns describing area-level characteristics used to explain variation in bias. The `distance` table is an OD table, so it should contain one row per origin-destination pair and a distance measure for that pair.

```{r}

nrow(OD_travel2work)
nrow(census_OD_travel2work)
nrow(coverage)
nrow(covariates)
nrow(distance)
```

Next, we can explore the content of each table. We print the column names and the first few rows to show basic structure of the core workflow tables. 

### Observed flows

The `OD_travel2work` table is the observed mobile-phone-derived OD-flow
table. Each row represents a flow from a home area (in this case, a Local Authority District or LAD), stored in `origin`, to a
workplace area, stored in `destination`. The `flow` column gives the observed
number of mobile-phone-derived travel-to-work movements for that OD pair.

Because the example was loaded with `complete_grid = TRUE`, the table also
contains row-status columns. These indicate whether an OD pair was observed in
the original Locomizer-derived data or added during complete-grid construction
with a zero flow. This makes the observed and benchmark OD tables easier to
compare because they share the same set of origin-destination pairs.

The column `mpd_source` identifies the data source. The columns `mpd_observed`,
`mpd_zero_filled` and `mpd_row_status` describe whether the row was present in
the original observed data or added as a zero-filled OD pair. Rows marked as
`mpd_observed = TRUE` were present in the Locomizer OD table; rows marked as
`zero_filled = TRUE` were added so that all possible OD pairs among the selected areas are represented.

```{r}
head(OD_travel2work)
```


###  Benchmark flows 

The `census_OD_travel2work` table is the benchmark OD-flow table. Each row
represents a Census travel-to-work flow from a home area, stored in `origin`, to a workplace area, stored in `destination`. The `flow` column gives the benchmark number of people travelling from the origin to the destination.

As with the observed-flow table, the benchmark table includes row-status columns
because the example was loaded with `complete_grid = TRUE`.
`benchmark_observed`, `benchmark_zero_filled` and `benchmark_row_status` show
whether each OD pair was present in the original Census table or added as a
zero-filled pair to align the benchmark and observed OD grids.

```{r}
head(census_OD_travel2work)
```

### Population counts

The `coverage` dataset is the area-level population count table used to measure bias. Each row represents one spatial unit. The `population` column gives the benchmark population count for the area, taken from the UK 2021 Census. The `user_count` column gives the observed digital trace population count, based on the number of users with detected home locations in that area.

Home locations have been inferred from anonymised mobile-phone location records before movements are aggregated into travel-to-work OD flows. Further details on the home and work detection process are provided by Zhong et al. (2025).

```{r}
head(coverage)
```

The key comparison is between $U_i$, the observed digital trace user count,
and $P_i$, the benchmark population count. The next vignette uses these values
to calculate coverage score $p_i = U_i / P_i$ and coverage bias
$b_i = 1 - p_i$.

