---
title: "Identifying and explaining population coverage bias"
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

## Aim

The previous vignette measured population coverage bias and showed that
some LADs are covered better than others. Now, we ask a second question:
are those differences in coverage related to the characteristics of the
local populations?

The aim is to show how coverage bias can be explored and explained. The
research paper associated with the DEBIAS methodology [@cabrera2025]
uses more flexible machine-learning methods for analysis. Here, we use a
simpler model because it is easier to understand and a useful first step
for package users.

## Load bias and covariates

In this vignette, we illustrate how `debiasR` can be used with example UK data at the local authority district (LAD) level. We start with two tables from debiasRdata: a coverage table and a covariates table. The coverage table contains the benchmark Census population and the observed digital trace user count for each LAD. The covariate table contains selected area-level characteristics from the 2021 Census.

```{r setup}
library(debiasR)
library(debiasRdata)
library(dplyr)
library(ggplot2)
```

The coverage table stores the LAD identifier in a column called `code`.
We rename it to `area` so that it uses the same identifier name as the
covariate table and the rest of the `debiasR` workflow.

```{r}
coverage <- debiasRdata::coverage_lad |>
  rename(area = code)

covariates <- debiasRdata::lad_covariates

head(coverage)
```

The covariates included in the data describe some demographic,
socioeconomic and geographic characteristics of each LAD: percentage of
UK-born residents, percentage of residents aged 20-29, percentage of
residents aged 70 or over, percentage of residents with Level 4
qualifications, percentage of households without central heating,
percentage of residents in routine occupations and percentage of rural land.

```{r}
head(covariates)
```

Next, we calculate the coverage score and coverage bias.

```{r}
bias_lad <- debiasR::measure_bias(coverage)

bias_lad |>
  select(area, name, population, user_count, coverage_score, coverage_bias) |>
  head()
```

The previous vignette introduced the idea of a proportional baseline. If
every LAD were covered at the same overall rate, each local coverage
score would sit close to the national coverage score. Residuals show
where a place departs from that simple expectation.

```{r}
bias_residuals <- debiasR::validate_bias_residual_structure(
  coverage_df = coverage,
  coverage_area_col = "area",
  population_col = "population",
  user_count_col = "user_count"
)

residual_lad <- bias_residuals$area_level |>
  mutate(
    baseline_coverage_bias = 1 - global_coverage_score,
    residual_bias = coverage_bias - baseline_coverage_bias
  ) |>
  left_join(
    coverage |> select(area, name),
    by = "area"
  )

residual_lad |>
  select(area, name, coverage_score, coverage_bias, residual_bias) |>
  head()
```

Positive values of `residual_bias` identify LADs with more coverage bias
than expected under the proportional baseline. Negative values identify
LADs with less coverage bias than expected.

Next, we join these residuals to the area-level covariates.

```{r}
bias_covariates <- residual_lad |>
  left_join(
    covariates |>
      select(
        area,
        per_age_20.29,
        per_age_70plus,
        per_level4,
        per_hh_no_centralheat,
        per_NS_SeC_L13_routine,
        rural_pct
      ),
    by = "area"
  )

bias_covariates |>
  select(
    area,
    name,
    residual_bias,
    per_age_20.29,
    per_age_70plus,
    per_level4,
    rural_pct
  ) |>
  head()
```

## Explore one relationship at a time

Before fitting a model, it is useful to look at simple bivariate
relationships. These plots ask whether residual bias tends to be higher
or lower in areas with different local characteristics.

```{r}
ggplot(bias_covariates, aes(x = per_age_20.29, y = residual_bias)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point(alpha = 0.7, colour = "#356859") +
  geom_smooth(method = "lm", se = FALSE, colour = "#8B3A3A") +
  labs(
    x = "Population aged 20-29 (%)",
    y = "Residual coverage bias",
    title = "Does residual bias vary with the young-adult population share?"
  ) +
  theme_minimal()
```

```{r}
ggplot(bias_covariates, aes(x = rural_pct, y = residual_bias)) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point(alpha = 0.7, colour = "#356859") +
  geom_smooth(method = "lm", se = FALSE, colour = "#8B3A3A") +
  labs(
    x = "Rural population (%)",
    y = "Residual coverage bias",
    title = "Does residual bias vary with rurality?"
  ) +
  theme_minimal()
```

These plots do not prove causality. They help identify whether coverage
bias is patterned rather than randomly distributed across places.

## Fit a simple explanatory model

We now fit a linear regression. The outcome is `residual_bias`, which
measures how much more or less coverage bias an LAD has compared with
the proportional baseline.

The predictors represent broad demographic, socioeconomic,
resource-access and geographic characteristics:

-   `per_age_20.29`: share of the population aged 20-29
-   `per_age_70plus`: share of the population aged 70 or over
-   `per_level4`: share of the population with Level 4 qualifications
-   `per_hh_no_centralheat`: share of households without central heating
-   `per_NS_SeC_L13_routine`: share of the population in routine
    occupations
-   `rural_pct`: rural population share

```{r}
bias_model <- lm(
  residual_bias ~
    per_age_20.29 +
    per_age_70plus +
    per_level4 +
    per_hh_no_centralheat +
    per_NS_SeC_L13_routine +
    rural_pct,
  data = bias_covariates
)

summary(bias_model)
```

For easier reading, we can extract the coefficient table.

```{r}
coefficient_table <- as.data.frame(coef(summary(bias_model)))
coefficient_table$term <- rownames(coefficient_table)
rownames(coefficient_table) <- NULL
names(coefficient_table) <- c("estimate", "std_error", "statistic", "p_value", "term")

coefficient_table |>
  select(term, estimate, std_error, statistic, p_value)
```

The sign of each coefficient gives the direction of the association,
holding the other variables in the model constant. A positive
coefficient means that higher values of the covariate are associated
with more residual coverage bias. A negative coefficient means that
higher values are associated with less residual coverage bias.

## Compare observed and fitted residuals

The fitted values from the model summarise the part of residual bias
that can be described by the selected covariates.

```{r}
bias_covariates <- bias_covariates |>
  mutate(
    fitted_residual_bias = fitted(bias_model),
    unexplained_residual = residual_bias - fitted_residual_bias
  )

ggplot(bias_covariates, aes(x = fitted_residual_bias, y = residual_bias)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_vline(xintercept = 0, colour = "grey80") +
  geom_point(alpha = 0.7, colour = "#356859") +
  geom_smooth(method = "lm", se = FALSE, colour = "#8B3A3A") +
  labs(
    x = "Fitted residual bias",
    y = "Observed residual bias",
    title = "How much residual bias is described by the selected covariates?"
  ) +
  theme_minimal()
```

The largest unexplained residuals are areas where the simple model does
not capture the local pattern well. These areas may be useful to inspect
before choosing an adjustment strategy.

```{r}
bias_covariates |>
  arrange(desc(abs(unexplained_residual))) |>
  select(
    area,
    name,
    residual_bias,
    fitted_residual_bias,
    unexplained_residual
  ) |>
  head(10)
```

## Interpretation

This vignette uses a simple regression to make the idea of explaining
bias concrete. If residual bias is associated with local
characteristics, then the data are not simply noisy. Some types of
places are more or less visible in the digital trace data than expected
under a constant-coverage baseline.

This matters for adjustment. A single global correction assumes that all
places are covered at roughly the same rate. If coverage bias varies
with demographic, socioeconomic or geographic characteristics, then
adjustment methods need to allow bias to vary across places.

The regression used here is deliberately simple and descriptive. The
research paper uses more flexible explainable machine-learning methods
to capture nonlinear and interacting relationships. The purpose of this
vignette is to introduce the same logic in a way that is easy to
inspect, reproduce and adapt.
