---
title: "Identifying and Explaining Bias"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette moves from measuring bias to diagnosing which area characteristics
are associated with geographic differences in mobile-phone-derived coverage.

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
```

## Bias as Spatial and Social Structure

Coverage bias is not only a technical nuisance. If $e_i = p_i - \bar{p}$ is
systematically related to benchmark flows or contextual variables, then the OD
matrix may encode spatial and social selection.

`validate_bias_residual_structure()` provides a reproducible diagnostic layer.
It does not claim causality. It identifies where residual coverage is associated
with benchmark mobility structure or selected covariates.

```{r}
diagnostics <- validate_bias_residual_structure(
  coverage_df = example_data$coverage,
  benchmark_od_df = example_data$census_msoa_OD_travel2work,
  covariate_df = example_data$covariates,
  covariate_col = "census_inflow"
)

diagnostics$summary
```

## Benchmark Flow Associations

```{r}
diagnostics$benchmark_flow_correlation
```

These correlations compare coverage residuals with benchmark origin and
destination totals. They help assess whether areas with large benchmark flows
also have systematically high or low coverage residuals.

## Covariate Associations

```{r}
diagnostics$covariate_correlation
```

The selected covariate can be changed when richer area-level inputs are
available. The example uses `census_inflow` because it is derived directly from
the benchmark OD table and is available in the package example workflow.

## Population-Trend Residual

The population-only residual fits a descriptive model of active users on
benchmark population. It is not a sampling model. It is a diagnostic for areas
where observed users are higher or lower than a simple population trend would
suggest.

```{r}
population_lm_diagnostics <- validate_bias_residual_structure(
  coverage_df = example_data$coverage,
  benchmark_od_df = example_data$census_msoa_OD_travel2work,
  covariate_df = example_data$covariates,
  covariate_col = "census_inflow",
  residual_type = "population_lm"
)

population_lm_diagnostics$population_lm
```

```{r}
population_lm_diagnostics$area_level |>
  arrange(desc(abs(population_lm_residual))) |>
  select(
    area,
    population,
    user_count,
    population_lm_expected_user_count,
    population_lm_residual
  ) |>
  head(10)
```

## Link to Adjustment

The Bayesian model uses these ideas directly. It treats coverage residuals and
area-level characteristics as part of the model for adjusted OD flows, rather
than assuming a single global correction is enough.
