---
title: "Adjusting Biases in OD Flows"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette demonstrates the adjustment functions in `debiasR`. The package
offers a menu of methods for adjusting mobile-phone-derived OD flows. These
methods differ in what they require, what assumptions they make, and how they
should be validated.

The Bayesian multilevel model is the main methodological innovation because it
can represent coverage bias, covariates, distance, and hierarchical structure in
one uncertainty-aware model. The other methods are also useful adjustment
options, especially when users need transparent weighting, benchmark-calibrated
scaling, or margin consistency.

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
```

## Adjustment Method Menu

The available methods answer different adjustment problems. A method should be
chosen based on input data availability, the scale of analysis, interpretability
needs, uncertainty requirements, and validation performance.

```{r}
method_menu <- tibble::tribble(
  ~method, ~method_family, ~main_inputs, ~best_when, ~main_limitation,
  "adjust_inverse_penetration()",
  "Coverage-based reweighting",
  "MPD OD flows; area population; active-user counts",
  "Coverage differences are the main known bias and a transparent correction is needed.",
  "Can over-correct sparse areas and does not model compositional selection.",
  "adjust_selection_rate()",
  "Selection-rate adjustment",
  "MPD OD flows; coverage inputs; covariates; optional benchmark OD flows for calibration",
  "Selection varies systematically with area attributes and calibration is useful.",
  "Depends on covariate quality and calibration choices.",
  "adjust_selection_rate2()",
  "Selection-rate adjustment",
  "MPD OD flows; coverage inputs; optional benchmark OD flows for k calibration",
  "A compact coverage-based correction is preferred.",
  "Less flexible when bias varies strongly across places.",
  "adjust_raking_ratio()",
  "Margin-constrained adjustment",
  "MPD OD flows; benchmark OD flows or trusted origin/destination margins",
  "Matching known origin and destination totals is the priority.",
  "Improves margins but cannot by itself explain local OD residual structure.",
  "adjust_coefficient()",
  "Benchmark-calibrated adjustment",
  "MPD OD flows; benchmark OD flows",
  "A benchmark OD table exists and a direct calibration relationship is acceptable.",
  "A global relationship can miss local heterogeneity.",
  "adjust_multilevel_bayes()",
  "Bayesian multilevel adjustment",
  "MPD OD flows; coverage inputs; covariates; optional distance; optional complete grid",
  "Bias is structured across places and uncertainty-aware partial pooling is needed.",
  "Requires heavier dependencies, runtime checks, and model diagnostics."
)

method_menu
```

## Bayesian Multilevel Adjustment

`adjust_multilevel_bayes()` estimates adjusted OD flows with a multilevel count
model. Unlike fixed-rule adjustment methods, it can pool information across
areas and return model metadata and diagnostics. The core idea is:

$$
\log(\mu_{ij}) =
f(X_i, X_j, d_{ij}, e_i, u)
$$

where $X_i$ and $X_j$ are origin and destination attributes, $d_{ij}$ is
OD distance, $e_i$ is the origin coverage residual or bias term, and $u$ is
an optional hierarchical term. The adjusted flow $F^{adj}_{ij}$ is obtained by
removing the estimated coverage-bias contribution from posterior linear
predictor draws.

The vignette uses a render-feasible square OD subset for the Bayesian model.
Set `DEBIASR_BAYES_MAX_OD_ROWS` to increase or decrease the maximum grid size.

```{r}
bayes_n_areas <- choose_bayes_area_count()
bayes_example <- load_workshop_example(
  n_areas = bayes_n_areas,
  complete_grid = TRUE
)

bayes_example$od_audit
```

```{r}
bayes_ready <- requireNamespace("rstanarm", quietly = TRUE) &&
  nrow(bayes_example$distance) > 0L

adj_bayes <- NULL
if (bayes_ready) {
  adj_bayes <- adjust_multilevel_bayes(
    mpd_od_df = bayes_example$mpd_od,
    coverage_df = bayes_example$coverage,
    covariates_df = bayes_example$covariates,
    distance_df = bayes_example$distance,
    prediction_scope = "complete_grid",
    random_intercept = "origin",
    model_family = "poisson",
    flow_adj_summary = "median",
    iter = 500,
    chains = 2,
    seed = 123,
    refresh = 0
  )

  attr(adj_bayes, "result_metadata")
} else {
  "Bayesian empirical rendering requires `rstanarm` and empirical OD distance. The complete-grid interface is available, but empirical Bayesian examples should be rendered only when the required inputs and runtime budget are available."
}
```

## Coverage-Based Reweighting

`adjust_inverse_penetration()` inflates observed flows using the inverse of the
area coverage score.

$$
w_i = \frac{P_i}{U_i}
$$

$$
F^{adj}_{ij} = F^{mpd}_{ij} w_i
$$

or, when both origin and destination coverage are used:

$$
F^{adj}_{ij} = F^{mpd}_{ij}\sqrt{w_i w_j}
$$

This method is useful when the main concern is uneven coverage and the user has
reliable population and active-user counts. It is simple and interpretable, but
it does not model social or spatial selection beyond the supplied coverage
rates.

## Selection-Rate Adjustment

`adjust_selection_rate()` and `adjust_selection_rate2()` extend coverage-based
adjustment by allowing the correction factor to vary with a selection parameter,
covariates, or calibration against a benchmark.

For `adjust_selection_rate()`, an origin-side weight can be written as:

$$
w_i^{(O)}(r_t) =
\frac{1}{I_i p_i^{(O)} + (1 - I_i)r_t}
$$

and the OD adjustment is:

$$
F^{adj}_{ij} = F^{mpd}_{ij}w_{ij}
$$

These methods are useful when selection bias is thought to vary systematically
with area characteristics. They require more judgement than inverse penetration
because covariate choice and calibration strategy affect the result.

## Margin-Constrained Adjustment

`adjust_raking_ratio()` uses iterative proportional fitting so adjusted OD flows
match trusted origin and destination margins:

$$
\sum_j F^{adj}_{ij} = T^{O}_i,\qquad
\sum_i F^{adj}_{ij} = T^{D}_j
$$

This is useful when consistency with known origin and destination totals is the
main requirement. It is widely used for matrix adjustment, but it should still be
validated at the OD-pair level because matching margins does not guarantee that
the internal destination allocation is correct.

## Benchmark-Calibrated Adjustment

`adjust_coefficient()` estimates a relationship between observed MPD flows and
benchmark flows, then applies that relationship to the MPD table:

$$
E(F^{bench}_{ij}) = \beta F^{mpd}_{ij}
$$

This is useful when a benchmark OD table exists and direct calibration is the
goal. Its strength is that it learns from the benchmark. Its limitation is that
a global coefficient can miss local heterogeneity unless the selected model
family and level capture the relevant structure.

## Fit Adjustment Methods

```{r}
method_results <- fit_adjustment_methods(example_data)

tibble::tibble(
  method = names(method_results),
  rows = vapply(method_results, nrow, integer(1)),
  adjusted_total = vapply(
    method_results,
    function(x) sum(x$flow_adj, na.rm = TRUE),
    numeric(1)
  )
)
```

```{r}
method_results$raking_ratio |>
  summarise(
    raw_total = sum(flow, na.rm = TRUE),
    adjusted_total = sum(flow_adj, na.rm = TRUE),
    median_weight = median(weight_ipf, na.rm = TRUE),
    ipf_converged = attr(method_results$raking_ratio, "ipf_converged"),
    ipf_iterations = attr(method_results$raking_ratio, "ipf_iterations")
  )
```

## Choosing And Validating A Method

No single adjustment method should be treated as universally best. The practical
choice depends on which inputs are available, which assumptions are acceptable,
and whether the adjusted OD matrix improves against the benchmark.

If the Bayesian model is rendered, validate it with the same functions used for
the other adjustment methods in `07-validation.qmd`. If it is not rendered, the
method menu still shows a complete-grid adjustment workflow and makes the
Bayesian preconditions explicit.
