---
title: "Validation"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

Validation asks whether adjusted flows are closer to benchmark OD flows and
where residual problems remain. No single metric is enough; the package returns
complementary diagnostics.

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
```

```{r}
method_results <- fit_adjustment_methods(example_data)
benchmark_od <- example_data$benchmark_od
```

## Overall Fit

`validate_flow_overall()` summarises agreement between $F^{adj}_{ij}$ and
$F^{bench}_{ij}$. Correlations describe concordance; RMSE, MAE, and MAPE
describe error magnitude.

```{r}
overall_metrics <- dplyr::bind_rows(
  lapply(
    names(method_results),
    function(method_name) validation_row(
      method_name,
      method_results[[method_name]],
      benchmark_od
    )
  )
) |>
  arrange(rmse)

overall_metrics
```

## OD-Pair Audit

`validate_flow_pairs()` returns a row-level table for auditing individual OD
pairs.

```{r}
validate_flow_pairs(
  adj_df = method_results$raking_ratio,
  benchmark_od_df = benchmark_od
) |>
  arrange(desc(abs(diff_adj_benchmark))) |>
  select(origin, destination, mpd_flow, adj_flow, benchmark_flow, diff_adj_benchmark) |>
  head(10)
```

## Residual Improvement

`validate_flow_residuals()` shows whether adjustment reduced error relative to
the raw MPD source. The signed residual movement is:

$$
(F^{bench}_{ij} - F^{mpd}_{ij}) -
(F^{bench}_{ij} - F^{adj}_{ij})
$$

The absolute residual-reduction metric is easier for method comparison:
positive values mean the adjusted flow is closer to the benchmark.

```{r}
residuals_raking <- validate_flow_residuals(
  adj_df = method_results$raking_ratio,
  benchmark_od_df = benchmark_od,
  method_name = "raking_ratio",
  top_n = 10
)

residuals_raking$summary
```

```{r}
residuals_raking$top_worst |>
  select(origin, destination, mpd_flow, adj_flow, benchmark_flow, abs_residual_adj_benchmark) |>
  head(10)
```

## Residual Structure

`validate_flow_residual_structure()` checks whether remaining errors are
associated with benchmark flow size or selected covariates.

```{r}
structure_raking <- validate_flow_residual_structure(
  adj_df = method_results$raking_ratio,
  benchmark_od_df = benchmark_od,
  method_name = "raking_ratio",
  covariate_df = example_data$covariates,
  covariate_col = "income_norm"
)

structure_raking$summary
```

## Destination Allocation Fidelity

`validate_flow_distribution()` compares destination-share distributions within
each origin. It reports:

- KL divergence, $KL(F^{bench} || F^{adj})$, a directional allocation metric
- Jensen-Shannon divergence, a symmetric companion metric

Lower values indicate that adjusted flows allocate each origin's flows across
destinations more like the benchmark.

```{r}
distribution_raking <- validate_flow_distribution(
  adj_df = method_results$raking_ratio,
  benchmark_od_df = benchmark_od,
  method_name = "raking_ratio",
  weight_by = "benchmark_origin_total"
)

distribution_raking$summary
```

## Bayesian Validation

When `adjust_multilevel_bayes()` is rendered with real OD distance, validate it
with the same functions used for the other adjustment methods. Bayesian
validation should report both predictive performance and runtime/model
diagnostics attached to the result object, because its main innovation is an
uncertainty-aware multilevel structure rather than a fixed adjustment rule.
