---
title: "Data"
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette describes the empirical data inputs used in the package examples.
The data are supplied by the companion `debiasRdata` package and cover two
geographical scales:

- Local Authority District / Lower Tier Local Authority style areas (`lad`)
- Middle Layer Super Output Areas (`msoa`)

Both scales contain mobile-phone-derived travel-to-work origin-destination (OD)
flows, Census 2021 benchmark OD flows, and Census area-level covariates. The LAD
scale also includes centroid coordinates used to derive selected-area OD
distances for examples.

The aim of this vignette is to help users decide whether the data are suitable
for their analysis: what each table represents, how large it is, what columns it
contains, what checks have been applied, and how the tables should be joined.

```{r setup}
common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
source(common_path)
load_debiasr_workshop()

if (!requireNamespace("debiasRdata", quietly = TRUE)) {
  cat(
    "Install `debiasRdata` to inspect the empirical datasets. ",
    "Use `remotes::install_github(\"de-bias/debiasRdata\")`.\n",
    sep = ""
  )
  knitr::knit_exit()
}

data_objects <- c(
  "lad_OD_travel2work",
  "census_lad_OD_travel2work",
  "lad_covariates",
  "lad_centroids",
  "msoa_OD_travel2work",
  "census_msoa_OD_travel2work",
  "msoa_covariates"
)

debiasrdata_tables <- setNames(vector("list", length(data_objects)), data_objects)
for (object in data_objects) {
  data(list = object, package = "debiasRdata", envir = environment())
  debiasrdata_tables[[object]] <- get(object)
}
```

## Dataset Families

The package is organised around matched datasets by geographical scale.

At the LAD scale:

- `lad_OD_travel2work`: observed MPD travel-to-work OD flows.
- `census_lad_OD_travel2work`: Census 2021 benchmark travel-to-work OD flows.
- `lad_covariates`: Census 2021 area-level covariates.
- `lad_centroids`: LAD centroid coordinates for distance construction.

At the MSOA scale:

- `msoa_OD_travel2work`: observed MPD travel-to-work OD flows.
- `census_msoa_OD_travel2work`: Census 2021 benchmark travel-to-work OD flows.
- `msoa_covariates`: Census 2021 area-level covariates.

The OD tables use the same minimal schema at both scales:

- `origin`: residential origin area code.
- `destination`: workplace destination area code.
- `flow`: nonnegative flow count for the OD pair.

The covariate tables use `area` as the join key. This key corresponds to
`origin` or `destination` in the OD tables, depending on whether the covariate
is being used as an origin-side or destination-side predictor.

## Inventory

```{r inventory}
od_summary <- function(x) {
  if (!all(c("origin", "destination", "flow") %in% names(x))) {
    return(tibble::tibble(
      origin_areas = NA_integer_,
      destination_areas = NA_integer_,
      total_flow = NA_real_,
      min_flow = NA_real_,
      max_flow = NA_real_,
      duplicated_od_pairs = NA_integer_
    ))
  }

  tibble::tibble(
    origin_areas = dplyr::n_distinct(x$origin),
    destination_areas = dplyr::n_distinct(x$destination),
    total_flow = sum(x$flow, na.rm = TRUE),
    min_flow = min(x$flow, na.rm = TRUE),
    max_flow = max(x$flow, na.rm = TRUE),
    duplicated_od_pairs = sum(duplicated(x[c("origin", "destination")]))
  )
}

dataset_inventory <- dplyr::bind_rows(lapply(names(debiasrdata_tables), function(object) {
  x <- debiasrdata_tables[[object]]
  scale <- if (grepl("^msoa|_msoa_", object)) "msoa" else "lad"
  family <- dplyr::case_when(
    grepl("^census_.*OD", object) ~ "benchmark_od",
    grepl("OD_travel2work$", object) ~ "mpd_od",
    grepl("covariates$", object) ~ "covariates",
    grepl("centroids$", object) ~ "centroids",
    TRUE ~ "other"
  )

  dplyr::bind_cols(
    tibble::tibble(
      scale = scale,
      object = object,
      family = family,
      rows = nrow(x),
      columns = ncol(x)
    ),
    od_summary(x)
  )
}))

dataset_inventory
```

The row counts differ sharply by scale. MSOA data are much more spatially
detailed and therefore much larger. They are useful when local spatial
heterogeneity matters, but they are heavier to render, validate, and model.
LAD data are coarser and smaller, which makes them the default scale for
workshop examples and fast vignette rendering.

## Structure

The following output shows the R structure of every packaged table.

```{r structure}
for (object in names(debiasrdata_tables)) {
  cat("\n\n", object, "\n", strrep("-", nchar(object)), "\n", sep = "")
  str(debiasrdata_tables[[object]], give.attr = FALSE)
}
```

## Columns

```{r columns}
column_roles <- tibble::tribble(
  ~column, ~role,
  "origin", "Origin area code in an OD flow table.",
  "destination", "Destination or workplace area code in an OD flow table.",
  "flow", "Nonnegative OD flow count.",
  "area", "Area identifier used to join covariates or centroids to OD origins/destinations.",
  "name", "Human-readable area name.",
  "year", "Reference year for Census covariates.",
  "per_ukborn", "Percentage of residents born in the UK.",
  "per_age_20.29", "Percentage of residents aged 20 to 29.",
  "per_age_70plus", "Percentage of residents aged 70 or over.",
  "per_level4", "Percentage with level 4 qualifications.",
  "per_hh_no_centralheat", "Percentage of households without central heating.",
  "per_NS_SeC_L13_routine", "Percentage in NS-SeC L13 routine occupations.",
  "rural_pct", "Percentage rural.",
  "easting", "British National Grid easting for LAD centroid.",
  "northing", "British National Grid northing for LAD centroid.",
  "longitude", "Longitude of LAD centroid in decimal degrees.",
  "latitude", "Latitude of LAD centroid in decimal degrees."
)

column_inventory <- dplyr::bind_rows(lapply(names(debiasrdata_tables), function(object) {
  x <- debiasrdata_tables[[object]]
  tibble::tibble(
    object = object,
    column = names(x),
    type = vapply(x, function(col) paste(class(col), collapse = ", "), character(1))
  )
})) |>
  dplyr::left_join(column_roles, by = "column")

column_inventory
```

## Example Rows

```{r example-rows}
for (object in names(debiasrdata_tables)) {
  cat("\n\n", object, "\n", strrep("-", nchar(object)), "\n", sep = "")
  print(utils::head(debiasrdata_tables[[object]], 6))
}
```

## Sources And Processing

The MPD OD flow tables are derived from the Zenodo record
`doi:10.5281/zenodo.13327082`. The MSOA MPD table keeps the source MSOA home and
workplace identifiers, normalised to `origin`, `destination`, and `flow`. The
LAD MPD table is aggregated from the MSOA MPD source using the ONS MSOA-to-LAD
lookup, then summed by LAD origin-destination pair.

The Census benchmark OD tables are derived from ONS Census 2021 origin-destination
workplace-flow table `ODWP01EW`. They retain people working in the UK but not
working at or from home, keep area-code-to-area-code OD pairs, normalise
identifiers to `origin` and `destination`, and aggregate duplicated OD pairs if
present.

The covariate tables are selected Census 2021 topic-summary percentages. They
are intended to explain spatial variation in MPD coverage and OD-flow bias. All
covariate percentage columns are numeric and use a 0 to 100 scale.

The LAD centroid table is derived from ONS LAD representative coordinate data.
It is not a full distance matrix. `debiasR_example_data(geography = "lad")`
uses these centroids to compute distances for the selected LAD areas.

## Quality Checks

For OD analysis, the minimum checks are:

- identifiers are present for `origin` and `destination`
- flows are finite and nonnegative
- each OD pair appears at most once
- the MPD and benchmark tables use compatible area identifiers
- the selected example support can be expanded to a square OD grid when
  adjustment or validation requires equal origin and destination sets

```{r quality-checks}
quality_checks <- dplyr::bind_rows(lapply(names(debiasrdata_tables), function(object) {
  x <- debiasrdata_tables[[object]]
  if (all(c("origin", "destination", "flow") %in% names(x))) {
    tibble::tibble(
      object = object,
      missing_origin = sum(is.na(x$origin) | x$origin == ""),
      missing_destination = sum(is.na(x$destination) | x$destination == ""),
      missing_flow = sum(is.na(x$flow)),
      nonfinite_flow = sum(!is.finite(x$flow)),
      negative_flow = sum(x$flow < 0, na.rm = TRUE),
      duplicated_od_pairs = sum(duplicated(x[c("origin", "destination")]))
    )
  } else if ("area" %in% names(x)) {
    tibble::tibble(
      object = object,
      missing_origin = NA_integer_,
      missing_destination = NA_integer_,
      missing_flow = NA_integer_,
      nonfinite_flow = NA_integer_,
      negative_flow = NA_integer_,
      duplicated_od_pairs = sum(duplicated(x$area))
    )
  }
}))

quality_checks
```

These checks do not prove that the MPD data are unbiased. They establish that
the tables are structurally suitable for bias measurement, adjustment, and
validation. Substantive quality is assessed by comparing MPD flows with the
Census benchmark and by modelling how residuals vary across geography and
covariates.

## Using The Data In `debiasR`

`debiasR_example_data()` loads a matched MPD/benchmark/covariate bundle and
normalises it to the package interface. The default scale is LAD because it is
small enough for fast examples.

```{r lad-example}
lad_example <- load_workshop_example(
  n_areas = 25,
  complete_grid = TRUE,
  geography = "lad"
)

names(lad_example)
lad_example$metadata
lad_example$od_audit
```

Use MSOA when the analysis needs finer geography and the runtime cost is
acceptable.

```{r msoa-example}
msoa_example <- load_workshop_example(
  n_areas = 25,
  complete_grid = TRUE,
  geography = "msoa"
)

msoa_example$metadata
msoa_example$od_audit
```

When `complete_grid = TRUE`, absent OD pairs in the selected support are
zero-filled and marked with row-status indicators. These indicators distinguish
observed source rows from rows created to complete the square OD matrix.

```{r complete-grid-status}
names(lad_example$mpd_od)
names(lad_example$benchmark_od)

lad_example$mpd_od |>
  dplyr::count(mpd_row_status)

lad_example$benchmark_od |>
  dplyr::count(benchmark_row_status)
```

## Choosing LAD Or MSOA

Use LAD data when:

- examples need to render quickly
- the analysis is exploratory or national in scope
- the model requires repeated fitting or Bayesian sampling
- coarser geographic interpretation is acceptable

Use MSOA data when:

- local heterogeneity is central to the research question
- the finer Census benchmark is needed
- computational cost is manageable
- there is a clear plan for validating sparse OD cells and zero-filled support

In both cases, MPD flows should be treated as observed, potentially biased
mobility signals. The Census OD tables provide the benchmark used to measure
coverage and representativeness problems, assess adjustment performance, and
decide whether the adjusted matrix is fit for purpose.
