---
title: "Adjusting biases in origin-destination flows"
vignette: >
  %\VignetteIndexEntry{Adjusting biases in origin-destination flows}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette describes the set of adjustment methods included in `debiasR`, to correct mobile-phone-derived origin-destination mobility counts.

The vignette describes how each adjustment method works, associated functions and parameters, and how to implement it.
All the methods in `debiasR` use mobile-phone-derived flows as input data and return an adjusted flow `flow_adj`. However, they differ in fundamental ways, in terms of data input requirements, how they derive adjustment weights and the extent they can adjust biases in the data. These differences are described in more detail below. The practical choice of which methods to use should be guided by what data
are available and validation against a credible benchmark. A key innovation of `debiasR` is the introduction of a Bayesian multilevel modelling approach which we also explain below.


## Intuition

Let us very quickly recall the challenge we face when working with mobile-phone-derived mobility flows. These flows are typically represent an small proportion of the resident population, hence we end up under-counting these populations. Though we can also over-count populations if individuals have multiple mobile devices and they are recorded in the data. So the idea of the methods included in `debiasR` is to adjust these counts to make these counts more representative of the actual observed flow. Formally, let $F^{mpd}_{ij}$ denote the observed mobile-phone-derived flow from origin $i$ to destination $j$, and let $F^{adj}_{ij}$ denote the adjusted flow. If a method treats the observed MPD flow as under-counted, it increases
$F^{adj}_{ij}$. If it treats the observed MPD flow as over-counted, it reduces
$F^{adj}_{ij}$.

Where, let us recall:

| Symbol | Description |
|:--|:--|
| $F^{mpd}_{ij}$ | Observed MPD flow from origin $i$ to destination $j$. |
| $F^{bench}_{ij}$ | Benchmark flow from origin $i$ to destination $j$. |
| $F^{adj}_{ij}$ | Adjusted flow returned by an adjustment method. |
| $P_i, P_j$ | Population at origin and destination. |
| $U_i, U_j$ | Active-user count at origin and destination. |
| $c_i = U_i/P_i$ | Active-user coverage rate. |
| $b_i = 1 - c_i$ | Active-user coverage bias. |
| $d_{ij}$ | Distance between origin $i$ and destination $j$. |
| $X_i, X_j$ | Origin and destination covariates, such as rural share. |

## Methods in `debiasR`

This vignette covers the bias adjustment methods currently available in
`debiasR`. The table below illustrates how they map onto `debiasR` functions and relevant references that have inspire our thinking. Below we explain how each of these methods adjust biases and how you can implement them via `debiasR`.

| Method | Function | Reference |
|:--|:--|:--|
| Inverse Penetration Rate Weights | `adjust_inverse_penetration()` | Chi et al. (2025) |
| Selection Rate | `adjust_selection_rate()` | Chi et al. (2025) |
| Selection Rate II | `adjust_selection_rate2()` | Zagheni and Weber (2012) |
| Raking Ratio (or Iterative proportional fitting) | `adjust_raking_ratio()` | Chi et al. (2025) |
| Coefficient regression modelling | `adjust_coefficient()` | Chi et al. (2025) |
| Bayesian multilevel modelling | `adjust_multilevel_bayes()` | Rowe and Cabrera (in progress); Raymer et al. (2013); Aparicio Castro et al. (2024) |

## Data

We will use data on travel-to-work flows between local authority districts
(LADs) from `debiasRdata`. The data capture origin-destination flows derived
from [Locomizer](https://locomizer.com), paired with Census benchmark flows in
the companion data package.

The README shows how to install `debiasR` and the companion `debiasRdata`
package from GitHub. We start by loading those packages.

```{r load-packages}
library(debiasR)
library(debiasRdata)
```

This vignette also uses helper functions defined in `vignettes/_common.R` to
run the compact all-method comparison near the end. When working from the
package source, you can load them with `source("vignettes/_common.R")`.

```{r load-adjustment-helpers, include=FALSE}
helper_candidates <- c(
  "vignettes/_common.R",
  "_common.R",
  "../vignettes/_common.R",
  "../../vignettes/_common.R"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]
if (is.na(helper_file)) {
  stop(
    "Could not find vignettes/_common.R. ",
    "Set the working directory to the debiasR package root or vignettes folder."
  )
}
source(helper_file)
```

We then load a small LAD travel-to-work example from `debiasRdata` through the
exported `debiasR_example_data()` helper.

```{r load-adjustment-data}
example_data <- debiasR_example_data(n_areas = 25, complete_grid = TRUE)
```

```{r}
od_flow <- example_data$lad_OD_travel2work
head(od_flow)
```

The method examples below use a common comparison view so the reported flow
columns have the same meaning throughout the vignette:

- `flow_mpd_raw`: the raw, unadjusted MPD flow used as input.
- `flow_mpd_adjusted`: the adjusted MPD-derived estimate returned by the
  adjustment method.
- `flow_benchmark`: the observed Census benchmark flow, shown for interpretation
  and validation.

```{r}
benchmark_flow_lookup <- example_data$benchmark_od |>
  dplyr::select(
    origin,
    destination,
    flow_benchmark = flow
  )

show_adjustment_comparison <- function(adjusted_df,
                                       extra_cols = character(),
                                       n = 5) {
  adjusted_df |>
    dplyr::left_join(
      benchmark_flow_lookup,
      by = c("origin", "destination")
    ) |>
    dplyr::select(
      origin,
      destination,
      flow_mpd_raw = flow,
      flow_mpd_adjusted = flow_adj,
      flow_benchmark,
      dplyr::any_of(extra_cols)
    ) |>
    dplyr::slice_head(n = n)
}
```


## Inverse penetration rate

The inverse penetration rate adjusts origin-destination flows by scaling observed mobile-phone-derived flows according to how well the population is represented in the mobile-phone data. For each area, we calculate the active-user coverage rate ($c_i$) as the number of observed active users ($U_i$) divided by the benchmark population ($P_i$). The inverse of this rate becomes the adjustment weight ($w_i$). Areas with lower coverage receive larger weights because their flows are assumed to be more undercounted. For example, if only 5% of residents in an origin area appear in the mobile-phone data, flows from that origin can be scaled up by a factor of 20; that is $1 / 0.05 = 20$, so each observed flow is treated as representing about twenty times as many people in the benchmark population. The adjustment can use origin coverage, destination coverage or a combination of both. This makes the method simple and transparent, but it assumes that bias is mainly proportional to population coverage and does not directly account for differences in who is represented among mobile-phone users. Formally:

*Penetration rate and inverse penetration rate weight*:

$$
c_i = \frac{U_i}{P_i}, \qquad
w_i = \frac{1}{c_i} = \frac{P_i}{U_i}
$$

*Origin-weighted adjustment*:

$$
F^{adj}_{ij} = w_i F^{mpd}_{ij}
$$

*Destination-weighted adjustment*:

$$
F^{adj}_{ij} = w_j F^{mpd}_{ij}
$$

*Both-side adjustment*:

$$
F^{adj}_{ij} = \sqrt{w_iw_j}F^{mpd}_{ij}
$$

### Implementation

To use `adjust_inverse_penetration()`, we input our observed mobile-phone-derived origin-destination flows via `mpd_od_df` and coverage data to `coverage_df` in tabular format. The coverage table should contain the benchmark population and active-user counts needed to calculate how strongly each area is underrepresented. The `weight_by` argument lets you choose where the correction is applied: "origin" adjusts flows using coverage at the origin, "destination" uses coverage at the destination and "both" combines both sides. The result is a data frame that preserves the original observed flow in `flow` and adds `flow_adj`, the adjusted flow after inverse penetration weighting.

If we implement inverse penetration weighting adjusting by origins only, we set `weight_by` to "origin" as done below. The resulting table compares the original observed MPD flow, the inverse-penetration-adjusted estimate and the Census benchmark flow. The difference between the adjusted and benchmark columns gives a quick descriptive view of the adjustment, though we assess it formally in the validation vignette.

```{r}
adj_inverse_penetration <- adjust_inverse_penetration(
  mpd_od_df = example_data$mpd_od,
  coverage_df = example_data$coverage,
  weight_by = "origin"
)

show_adjustment_comparison(adj_inverse_penetration)
```

This method is a useful first adjustment when the main known source of bias is uneven population coverage across areas. It is especially helpful when we have reliable benchmark population counts and active-user counts, but no trusted benchmark origin-destination flow table. In that setting, inverse penetration weighting provides a transparent way to rescale observed MPD flows so that areas with lower mobile-phone coverage contribute more strongly to the adjusted flow estimates. Its main limitation is that it treats coverage as the correction mechanism. It assumes that the observed users in an area are broadly representative of the residents who are missing from the data. It therefore cannot directly account for social, demographic, behavioural or spatial differences in who appears in the MPD source. If mobile-phone users are systematically different from non-users, or if the relationship between coverage and mobility varies across places, this method may reduce undercounting while still leaving important selection bias in the adjusted flows.

## Selection rate

Selection rate adjusts origin-destination flows by allowing the adjustment
weight to depend on both population coverage and a covariate that helps explain
who appears in the mobile-phone-derived data. The intuition is that two areas
can have the same active-user coverage rate but still differ in the kinds of
people represented in the MPD source. The selection-rate method therefore adds
an area characteristic, such as rural share, to the coverage correction. It can also
use benchmark OD flows to calibrate the selection parameter, so that the
adjusted flows are closer to a trusted benchmark when one is available.
This makes the method more flexible than inverse penetration weighting, but it
also makes the result more dependent on the quality of the covariate and the
calibration choice. Formally:

*Origin-side selection weight*:

$$
w_i^{(O)}(r_t) =
\frac{1}{I_i c_i^{(O)} + (1 - I_i) r_t}
$$

*Destination-side selection weight*:

$$
w_j^{(D)}(r_t) =
\frac{1}{I_j c_j^{(D)} + (1 - I_j) r_t}
$$

*Adjusted flow*:

$$
F^{adj}_{ij} = F^{mpd}_{ij}w_{ij}
$$

where $I_i$ and $I_j$ are covariate values; $r_t$ is the selection-rate
parameter; and, $w_{ij}$ is the origin, destination or both-side correction.

### Implementation

To use `adjust_selection_rate()`, we again input the observed MPD origin-destination
flows through `mpd_od_df` and the population coverage table through
`coverage_df`. The `covariates_df` and `covariate_col` arguments tell the
function which area-level characteristic to use when estimating differential
selection. In this example, we use `rural_pct`, the rural land-share covariate
from `debiasRdata`. 
While all other parameters in `adjust_selection_rate()` have defaults, we recommend users to include specify `benchmark_od_df` for a meaningful selection-rate adjustment. This data frame corresponds to a benchmark OD table used to calibrate the selection rate parameter `r` in the equations above. 


::: {.callout-note title="What about if you do not have benchmark flow?"}
 `adjust_selection_rate()` has two no-benchmark routes:

(1) you can provide `r_global` directly based on prior evidence, substantive knowledge or a
 sensitivity-analysis design.

(2) if both `benchmark_od_df` and `r_global` are omitted, `debiasR` uses a fallback based on the overall active-user coverage rate `sum(U) / sum(P)`. This makes the method executable,
 but the result should be interpreted as an exploratory adjustment rather than
 a calibrated estimate.

 *Our recommendation* is to calibrate with a credible
 `benchmark_od_df` whenever one is available. If no benchmark exists, we
 recommend treating `r_global` as a sensitivity parameter. Report the fallback
 value, test a small set of plausible alternatives and validate the adjusted
 flows against any aggregate or external evidence that is available.
:::

To calibrate MPD flows, `calibration_aggregate`is set to origin. This means that MDP flows are calibrated *against* origin-level benchmark totals. This function tries each value in a grid of r values (`r_grid`), adjusts the MPD flows, sums adjusted and benchmark flows by origin, and chooses the `r_t` that minimises the absolute difference in origin totals. `r_grid` is the set of candidate values that `adjust_selection_rate()` tries when it needs to calibrate the global selection parameter `r_t`. If `calibration_aggregate` is set to `od`, it skips that origin aggregation. Instead, for each candidate `r_t`, it compares each adjusted OD cell directly with the benchmark OD cell (as described in the note below), and picks the `r_t` with the smallest total OD-level error. The `weight_by` argument controls whether the correction is applied using origin coverage, destination coverage or both. 


```{r}
adj_selection_rate <- adjust_selection_rate(
  mpd_od_df = example_data$mpd_od,
  coverage_df = example_data$coverage,
  covariates_df = example_data$covariates,
  covariate_col = "rural_pct",
  weight_by = "origin",
  benchmark_od_df = example_data$benchmark_od,
  calibration_aggregate = "origin"
)

show_adjustment_comparison(adj_selection_rate)

tibble::tibble(r_global = attr(adj_selection_rate, 
                               "r_global"))
```

The comparison table reports the original MPD flow, the selection-rate adjusted
estimate and the Census benchmark flow. Because this example uses
`weight_by = "origin"`, all flows leaving a given origin are adjusted using that
origin's coverage and selected covariate value. The reported `r_global`
attribute is the calibrated selection-rate parameter chosen by the function.
This method is useful when coverage alone is too simple and we believe
representation varies with an area characteristic. Its main limitation is that
the chosen covariate must be meaningful: if the covariate is weakly related to
selection into the MPD
source, the adjustment may add complexity without reducing bias. The `weight_missing` column in the `adj_selection_rate` is a diagnostic flag if missing and / or invalid origin coverage data leads to missing and / or invalid weight inputs.

## Selection rate II

Selection rate II adjusts origin-destination flows by applying a compact
coverage correction controlled by a single curvature parameter, $k$. The method
keeps the focus on population coverage, but instead of using the simple inverse
of the coverage rate, it transforms the coverage rate through a nonlinear
correction factor. This gives users a middle ground between inverse penetration
weighting and the fuller covariate-based selection-rate model. When benchmark
OD flows are available, `debiasR` can calibrate $k$ so that the adjusted flows
better match a trusted comparison. Formally:

*Coverage correction factor*:

$$
CF(c; k) =
c\frac{\exp(-k) - 1}{\exp(-kc) - 1}
$$

where $CF(c; k)$ is the coverage correction factor, $c$ is the active-user
coverage rate for the selected origin or destination area, and $k$ is the
curvature parameter that controls how strongly coverage is transformed. 

*Adjusted flow*:

$$
F^{adj}_{ij} = F^{mpd}_{ij}CF(c; k)
$$

The same correction can be applied on the origin side, destination side, or both
sides depending on `weight_by`.

### Implementation

To use `adjust_selection_rate2()`, the strict minimum set of parameter to be specified are MPD flows and coverage data. Input the observed MPD flows to `mpd_od_df`
and the coverage table to `coverage_df`. The `weight_by` argument controls
whether the correction uses origin coverage, destination coverage or both. You
can set `k` directly if you have a chosen correction curvature, or leave it as
`NULL` and provide `benchmark_od_df` so the function calibrates `k` over
`k_grid`. As above, `calibration_aggregate = "origin"` calibrates against
origin-level benchmark totals. The parameter `k` is the analog to `r` in `adjust_selection_rate()`.

```{r}
adj_selection_rate2 <- adjust_selection_rate2(
  mpd_od_df = example_data$mpd_od,
  coverage_df = example_data$coverage,
  weight_by = "origin",
  benchmark_od_df = example_data$benchmark_od,
  calibration_aggregate = "origin"
)

show_adjustment_comparison(adj_selection_rate2)

tibble::tibble(k = attr(adj_selection_rate2, "k"))
```

The comparison table reports the observed MPD count, the adjusted flow and the
Census benchmark flow. In this origin-weighted example, the correction applied to a flow
depends on the origin area's coverage rate and the calibrated $k$ value. The
reported `k` attribute shows the curvature parameter used by the adjustment.
This method is useful when coverage is the dominant concern and a compact
correction is adequate. Its main limitation is that one parameter may be too
simple when selection bias varies strongly across places or population groups.

## Raking ratio

Raking ratio is also known as iterative proportional
fitting (IPF). 
Raking ratio adjusts origin-destination flows by iteratively rescaling the
observed MPD matrix so that its origin and destination totals match trusted
benchmark margins. The intuition is different from coverage weighting. Instead
of calculating a weight from active-user coverage, the method asks the adjusted
OD table to satisfy known aggregate constraints. If we trust total outflows from
each origin and total inflows to each destination, IPF updates the OD cells until the row and column totals agree with those
benchmarks. Formally:

*Margin constraints*:

$$
\sum_j F^{adj}_{ij} = O_i^{bench},
\qquad
\sum_i F^{adj}_{ij} = D_j^{bench}
$$

where $O_i^{bench}$ and $D_j^{bench}$ are benchmark outflow and inflow totals.

### Implementation

To use `adjust_raking_ratio()`, the minimum set of parameter to specify is the MDP flows and set of constraints. Input the observed MPD flows to `mpd_od_df` and
either provide `benchmark_od_df` or supply explicit `origin_targets` and
`destination_targets`. In this vignette, we use `benchmark_od_df`, so `debiasR`
derives the target origin and destination margins from the benchmark OD table.

The `max_iter` and `tol` arguments control the iterative fitting process:
`max_iter` sets the maximum number of update cycles and `tol` sets the
convergence tolerance.

```{r}
adj_raking_ratio <- adjust_raking_ratio(
  mpd_od_df = example_data$mpd_od,
  benchmark_od_df = example_data$benchmark_od
)

show_adjustment_comparison(adj_raking_ratio)

adj_raking_ratio |>
  dplyr::summarise(
    raw_total = sum(flow, na.rm = TRUE),
    adjusted_total = sum(flow_adj, na.rm = TRUE),
    median_weight = median(weight_ipf, na.rm = TRUE),
    ipf_converged = attr(adj_raking_ratio, "ipf_converged"),
    ipf_iterations = attr(adj_raking_ratio, "ipf_iterations")
  )
```

The first table shows how individual observed MPD flows have been converted into
adjusted flows and how those adjusted flows compare with the Census benchmark.
The summary reports whether the
IPF routine converged, how many iterations it used, and how much the total flow
changed after adjustment. Raking is useful when trusted marginal totals are
available and consistency with those totals is a priority. Its main limitation
is that matching origin and destination margins does not guarantee that the
internal allocation between specific origin-destination pairs is correct, so it
should still be validated at the OD-pair level.

## Coefficient regression modelling

Coefficient regression modelling adjusts origin-destination flows by learning a
direct empirical relationship between observed MPD flows and benchmark flows.
The method is useful when a trusted benchmark OD table is available for the
same origin-destination pairs. Instead of using population coverage or margins,
it fits a regression model that estimates how MPD counts map onto benchmark
counts, then applies that fitted relationship to the MPD table. In its simplest
form, this is a single multiplicative coefficient. In count-model versions, the
relationship is estimated through a log link. 
Flow measured as proportions or probability as, for example, the number of outflows over the population at risk could be used as input, as well. 
Formally:

*Core proportional form*:

$$
E(F^{bench}_{ij}) = \beta F^{mpd}_{ij}
$$

*Count-model form*:

$$
\log(\mu_{ij}) = \alpha + \log(F^{mpd}_{ij})
$$

so that:

$$
\mu_{ij} = e^{\alpha}F^{mpd}_{ij},
\qquad
\beta = e^{\alpha}
$$

### Implementation

To use `adjust_coefficient()`, the minimum input requirement are the MPD and benchmark flows. Input the observed MPD flows to `mpd_od_df` and a
benchmark OD table to `benchmark_od_df`. The `model_family` argument controls
the regression family used to learn the MPD-to-benchmark relationship; this
example uses `"ols"` for a simple proportional calibration. The `level`
argument controls whether the coefficient is estimated at the OD, origin or
destination level. Additional options, such as `fit_intercept` and `by_source`, let
users decide whether to estimate an intercept or source-specific coefficients.

```{r}
adj_coefficient <- adjust_coefficient(
  mpd_od_df = example_data$mpd_od,
  benchmark_od_df = example_data$benchmark_od,
  model_family = "ols",
  level = "od"
)

show_adjustment_comparison(adj_coefficient)

attr(adj_coefficient, "coef")
```

The comparison table shows the original observed MPD flow, the regression
adjusted estimate and the Census benchmark flow. The coefficient attribute reports the fitted
scaling relationship used to convert MPD flows into benchmark-like flows. This
method is direct and easy to validate because the benchmark is part of the
model-fitting process. Its main limitation is that benchmark quality is
critical, and a simple coefficient can miss local heterogeneity unless the
chosen model family and estimation level capture the relevant structure.

## Bayesian multilevel modelling

Bayesian multilevel modelling adjusts origin-destination flows by fitting a
structured count model to the observed MPD flows and estimating the underlying
true-flow scale. This vignette focuses on the recommended default
implementation: the coverage-offset true-flow model. This mode treats MPD flows
as coverage-scaled observations of true flows, rather than as flows that are
only adjusted after fitting by switching off a bias covariate. The multilevel
structure can borrow information across origins, destinations and corridors when
those data structures are available. The Bayesian implementation can take time,
so the vignette shows the model code but renders a precomputed example output.
For analysis, run enough iterations and chains for your data and check the
sampler diagnostics before interpreting the posterior summaries.

`adjust_multilevel_bayes()` also includes advanced Bayesian modes for repeated
source/time data and backward compatibility. Those options are described in the
[advanced Bayesian adjustment vignette](v09-advanced-bayesian-adjustment.html);
the practical default for routine adjustment is:

| Target | Recommended setting |
|:--|:--|
| Adjustment scale | `target_scale = "true_flow"` |
| Observation model | `observation_model = "coverage_offset"` |
| Coverage offset | `coverage_scale = "origin"`, `"destination"` or `"both"` |

In coverage-offset true-flow mode, the model can be written as:

*_Level 1: MPD observation model_*

$$
F^{mpd}_{ij} \sim \text{Poisson}(q_{ij} F^{true}_{ij})
$$

$$
\log E(F^{mpd}_{ij}) = \log(q_{ij}) + \eta^{true}_{ij}
$$

where $q_{ij}$ is the observation probability derived from active-user coverage.
With `coverage_scale = "origin"`, $q_{ij} = c_i = U_i/P_i$. The term
$\log(q_{ij})$ is an offset: it fixes the observation-process scale and is not
estimated as a regression coefficient.

*_Level 2: true mobility-flow model_*

$$
\eta^{true}_{ij}
= \alpha + \beta_o X_i + \beta_d X_j + \gamma \log d_{ij}
+ u_i + v_j + w_{ij} + \xi_{ij}
$$

where $X_i$ and $X_j$ are origin and destination covariates, $d_{ij}$ is the
origin-destination distance, and $u_i$, $v_j$ and $w_{ij}$ capture origin,
destination and origin-destination structure.

Intuitively, Level 1 recognises that we observe MPD flows, not the true mobility
flows themselves. The observed MPD flow $F^{mpd}_{ij}$ is treated as a
coverage-scaled measurement of an underlying true flow $F^{true}_{ij}$. If an
origin has 20% active-user coverage, the fitted MPD-scale expectation is the
true-flow expectation multiplied by 0.20 under the origin-offset assumption.
For example, if the model estimates 500 true trips from origin $i$ to
destination $j$ and origin $i$ has 20% active-user coverage, the corresponding
MPD observation-scale prediction is $0.20 \times 500 = 100$ trips.

Level 2 describes the underlying mobility pattern learned from covariates,
distance and origin/destination/corridor structure. It assumes that the true flow
$F^{true}_{ij}$ depends on origin and destination characteristics, distance, and
persistent origin, destination and corridor structure. This gives the model a
way to borrow information across related places instead of treating every OD
corridor as a completely separate unit.

$F^{true}_{ij}$ is not itself a random intercept. It is the estimated true-flow
scale for a specific origin-destination pair, computed from the full true-flow
linear predictor $\eta^{true}_{ij}$. Random intercepts, when included, are
components of that predictor. An origin random intercept $u_i$ shifts all flows
leaving origin $i$ up or down; a destination random intercept $v_j$ shifts flows
arriving at destination $j$; and an OD random intercept $w_{ij}$ captures a
corridor-specific deviation. In the small S1 example below we set
`random_intercept = "none"` because each corridor appears once, so
$F^{true}_{ij}$ is estimated from the fixed-effect predictors and the coverage
offset rather than from random intercepts.

The model adjusts flows by predicting on two scales. The fitted MPD-scale
prediction keeps the coverage offset:

$$
F^{mpd,pred}_{ij} = \exp(\log(q_{ij}) + \eta^{true}_{ij})
$$

The adjusted true-flow prediction removes the observation offset:

$$
F^{adj}_{ij} = F^{true,pred}_{ij} = \exp(\eta^{true}_{ij})
$$

In a Bayesian framework, these predictions are computed draw-by-draw from the
posterior and summarised by mean or median. This means `flow_adj` is a posterior
summary of the true-flow prediction, not a single deterministic rescaling of the
observed MPD count.

### Implementation

To use `adjust_multilevel_bayes()`, input the observed MPD flows to `mpd_od_df`,
the population coverage table to `coverage_df`, area-level covariates to
`covariates_df` and OD distances to `distance_df`. The compact default example
below uses a single source and a single observation period. The
`prediction_scope` argument controls whether the method adjusts only observed
MPD pairs or predicts over a supplied complete OD grid.

The equations above translate into two conceptual formula components.
`mobility_formula` defines the true-flow model; here we use `rural_pct_o`,
`rural_pct_d` and `log_distance` to represent origin covariates, destination
covariates and distance. In the recommended default mode, the observation model
is selected with `target_scale = "true_flow"` and
`observation_model = "coverage_offset"`, while `coverage_scale` selects whether
the offset uses origin, destination or geometric-mean coverage.

For repeated source/time observations, legacy reduced-form behaviour, and the
experimental latent two-level model, see the
[advanced Bayesian adjustment vignette](v09-advanced-bayesian-adjustment.html).
Keeping those choices separate from the default example makes the basic workflow
easier to follow.

The example below therefore uses a Bayesian fixed-effect true-flow model with
an origin coverage offset. The model code is shown for reproducibility, while
the rendered table is loaded from a precomputed package artifact so routine
vignette updates do not rerun MCMC.
Regenerate the artifact explicitly with
`Rscript scripts/precompute_v06_bayesian_example.R`.

```{r eval=FALSE}
mpd_s1 <- example_data$mpd_od |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

coverage_s1 <- example_data$coverage |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

adj_multilevel <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = example_data$covariates,
  distance_df = example_data$distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median",
  include_flow_adj_draws = TRUE,
  iter = 1000,
  chains = 2,
  seed = 123,
  refresh = 0
)
```

The precomputed output below reports both posterior median and posterior mean
summaries from this model fit.

```{r echo=FALSE}
bayesian_example_file <- system.file(
  "extdata",
  "v06-bayesian-s1-comparison.csv",
  package = "debiasR"
)

if (identical(bayesian_example_file, "")) {
  stop(
    "Missing precomputed Bayesian vignette output. ",
    "Regenerate it with scripts/precompute_v06_bayesian_example.R."
  )
}

bayesian_example <- tibble::as_tibble(
  utils::read.csv(bayesian_example_file, check.names = FALSE)
)

knitr::kable(bayesian_example, digits = 2)
```

For the Bayesian model, `flow_mpd_adjusted_median` and
`flow_mpd_adjusted_mean` are posterior summaries of the adjusted MPD-derived
true-flow estimate. The function call sets `flow_adj_summary = "median"`, so
the returned `flow_adj` equals the posterior median by default; the vignette
also reports the posterior mean so readers can see whether the posterior is
skewed. The benchmark column is shown for interpretation and validation, but it
is not used to fit the Bayesian model. The `flow_mpd_pred_*` columns are the
corresponding fitted expected flows on the MPD observation scale, before
removing the coverage offset.

The saved example uses moderate sampler settings for a compact illustration.
For applied analysis, run enough iterations and chains for your data and inspect
the diagnostics:

```{r eval=FALSE}
adj_multilevel_bayes <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = example_data$covariates,
  distance_df = example_data$distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)

attr(adj_multilevel_bayes, "diagnostics")
```

*Additional outputs*

You can inspect the metadata to confirm the scenario and formula fitted.

```{r echo=FALSE}
bayesian_metadata_file <- system.file(
  "extdata",
  "v06-bayesian-s1-metadata.csv",
  package = "debiasR"
)

bayesian_terms_file <- system.file(
  "extdata",
  "v06-bayesian-s1-model-terms.csv",
  package = "debiasR"
)

if (identical(bayesian_metadata_file, "") || identical(bayesian_terms_file, "")) {
  stop(
    "Missing precomputed Bayesian vignette metadata. ",
    "Regenerate it with scripts/precompute_v06_bayesian_example.R."
  )
}

tibble::as_tibble(utils::read.csv(bayesian_metadata_file))
tibble::as_tibble(utils::read.csv(bayesian_terms_file))
```

The output returns the observed MPD flow in `flow`, the fitted MPD-scale
prediction in `flow_mpd_pred`, and the estimated true-flow adjustment in
`flow_true_pred` and `flow_adj`. The source and time columns confirm that this
example is using the S1 contract: one MPD source and one observation period.
The metadata report the engine, target scale, observation model, coverage
offset, resolved scenario, repeated-observation structure, prediction scope and
model terms used to fit the adjustment model.
This method is useful when users need a structured model that can combine
coverage, covariates, distance and hierarchical data structure. Its main
limitation is that it is more data hungry than other methods and requires
significantly more runtime, dependencies and diagnostic checking when fitted
with the full Bayesian engine. The key strength is that the model does not
require a benchmark dataset of flows.

*More complex Bayesian structures*

The example above uses the simplest S1 structure: one source and one observation
period. `adjust_multilevel_bayes()` can also accept S2-S4 source/time
structures through `scenario`, `source_col`, `time_col` and
`repeated_observation`. The
[advanced Bayesian adjustment vignette](v09-advanced-bayesian-adjustment.html)
explains when to keep the default coverage-offset model, when to use the
experimental latent two-level model, and when the legacy reduced-form mode is
useful for backward compatibility.

## Fit all methods

We can also run all six adjustment methods at once on the same empirical LAD
subset. The helper uses the frequentist multilevel engine by default here so the
method-comparison table renders quickly; switch `multilevel_engine` to
`"bayesian"` when you want posterior summaries and have time to inspect sampler
diagnostics.

```{r}
method_results <- fit_adjustment_methods(
  example_data,
  include_multilevel = TRUE,
  multilevel_engine = "frequentist"
)

method_engine <- function(x) {
  engine <- attr(x, "model_engine", exact = TRUE)
  if (is.null(engine)) "deterministic" else engine
}

tibble::tibble(
  method = names(method_results),
  engine = vapply(method_results, method_engine, character(1)),
  rows = vapply(method_results, nrow, integer(1)),
  adjusted_total = vapply(
    method_results,
    function(x) sum(x$flow_adj, na.rm = TRUE),
    numeric(1)
  )
)
```

## Method summary

As described above, each of the adjustment methods implemented via `debiasR` has different assumptions, input data requirements, strengths and weaknesses. The table below offers a summary which can be used as a conceptual
guide helping users identify which method is more appropriate given their purpose and constraints.

```{r results='asis', echo=FALSE}
eq_stack <- function(...) {
  paste0(
    "<div class='eq-stack'>",
    paste0("<div class='eq-line'>", c(...), "</div>", collapse = ""),
    "</div>"
  )
}

method_summary <- tibble::tribble(
  ~Method, ~`Correction<br>level`, ~`How the correction<br>is applied`, ~Rationale, ~Equation, ~`Benchmark<br>population`, ~`Active<br>users`, ~`Benchmark<br>OD flows`, ~Covariates, ~Strengths, ~Limitations,
  "Inverse Penetration Rate Weights",
  "Area-specific coverage correction",
  "Applies origin, destination, or both-side coverage weights using <code>weight_by</code>.",
  "Inflates observed flows using population-to-user ratios where coverage is thin.",
  eq_stack(
    "<i>w</i><sub>i</sub> = P<sub>i</sub> / U<sub>i</sub>",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>i</sub>, <i>w</i><sub>j</sub>, or &radic;(<i>w</i><sub>i</sub><i>w</i><sub>j</sub>)"
  ),
  "Yes", "Yes", "No", "No",
  "Simple, transparent, fast, and does not require benchmark OD flows.",
  "Can overinflate sparse areas and misses compositional selection bias.",

  "Selection Rate",
  "Area-specific selection correction",
  "Applies origin, destination, or both-side selection weights controlled by a global <i>r</i><sub>t</sub> parameter.",
  "Uses coverage, covariates, and optional benchmark calibration to model differential selection.",
  eq_stack(
    "<i>w</i><sub>i</sub>(<i>r</i><sub>t</sub>) = 1 / (I<sub>i</sub><i>c</i><sub>i</sub> + (1 - I<sub>i</sub>)<i>r</i><sub>t</sub>)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>"
  ),
  "Yes", "Yes", "Optional", "Yes",
  "More flexible than inverse penetration and can use benchmark calibration.",
  "Depends on covariate quality, calibration choices, and a meaningful selection parameter.",

  "Selection Rate II",
  "Area-specific nonlinear coverage correction",
  "Applies origin, destination, or both-side coverage correction controlled by a global <code>k</code> parameter.",
  "Applies a compact coverage correction with optional <code>k</code> calibration.",
  eq_stack(
    "CF(<i>c</i>; <i>k</i>) = <i>c</i> &times; (exp(-<i>k</i>) - 1) / (exp(-<i>k</i><i>c</i>) - 1)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; CF(...)"
  ),
  "Yes", "Yes", "Optional", "No",
  "Parsimonious, interpretable, and lighter than a covariate-rich adjustment.",
  "Less flexible when bias varies strongly across areas.",

  "Raking Ratio (or Iterative proportional fitting)",
  "Margin-constrained correction",
  "Iteratively rescales OD cells so adjusted origin and/or destination totals match supplied margins.",
  "Uses iterative proportional fitting so adjusted flows match trusted origin and destination margins.",
  eq_stack(
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>",
    "&Sigma;<sub>j</sub><i>F</i><sup>adj</sup><sub>ij</sub> = O<sub>i</sub> and &Sigma;<sub>i</sub><i>F</i><sup>adj</sup><sub>ij</sub> = D<sub>j</sub>"
  ),
  "Optional as targets", "No", "Yes or user targets", "No",
  "Guarantees consistency with supplied marginal totals.",
  "Matching margins does not guarantee correct destination allocation within each origin.",

  "Coefficient regression modelling",
  "Global coefficient correction",
  "Estimates a multiplicative coefficient from MPD-benchmark overlap and applies it to the MPD table.",
  "Learns a direct mapping from MPD flows to benchmark flows and applies it to the MPD table.",
  eq_stack(
    "E[<i>F</i><sup>bench</sup><sub>ij</sub>] = &beta; <i>F</i><sup>mpd</sup><sub>ij</sub>",
    "log(&mu;<sub>ij</sub>) = &alpha; + log(<i>F</i><sup>mpd</sup><sub>ij</sub>)"
  ),
  "No", "No", "Yes", "No",
  "Direct benchmark-driven calibration with several model families.",
  "Benchmark quality is critical and a global coefficient can miss local heterogeneity.",

  "Bayesian multilevel modelling",
  "Model-based corridor correction",
  "Predicts OD-specific adjusted flows from a true-flow formula and a coverage-offset observation model.",
  "Fits a count model where active-user coverage enters as a fixed observation-process offset, then returns the estimated true-flow scale.",
  eq_stack(
    "log E(<i>F</i><sup>mpd</sup><sub>ij</sub>) = log(<i>q</i><sub>ij</sub>) + &eta;<sup>true</sup><sub>ij</sub>",
    "&eta;<sup>true</sup><sub>ij</sub> = &alpha; + X<sub>i</sub>&beta;<sub>o</sub> + X<sub>j</sub>&beta;<sub>d</sub> + &gamma;log(<i>d</i><sub>ij</sub>)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = exp(&eta;<sup>true</sup><sub>ij</sub>)"
  ),
  "Yes", "Yes", "No", "Yes",
  "Richer structure, no benchmark OD needed for fitting, and a path to uncertainty-aware partial pooling.",
  "Heavier runtime and dependencies; posterior fitting requires sampler diagnostics and careful runtime checks."
)

table_html <- knitr::kable(
  method_summary,
  format = "html",
  escape = FALSE,
  table.attr = "class='table method-summary-table'"
)

cat("<div class='method-table-wrap'>", table_html, "</div>")
```

<style>
main table {
  font-size: 0.9rem;
  line-height: 1.45;
}

main table th,
main table td {
  vertical-align: top;
}

.method-table-wrap {
  overflow-x: auto;
  margin: 1.25rem 0 1.75rem 0;
  padding: 0.6rem 0.8rem;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  background: #fff;
}

.method-summary-table {
  min-width: 1650px;
}

.method-summary-table th,
.method-summary-table td {
  padding: 0.75rem;
  white-space: normal;
}

.eq-stack {
  font-family: "Cambria Math", "STIX Two Text", "Times New Roman", serif;
  line-height: 1.45;
}

.eq-line {
  display: block;
}

.eq-line + .eq-line {
  margin-top: 0.25rem;
}
</style>
