---
title: "Validation"
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

Validation asks whether adjusted mobile-phone-derived flows are closer to
benchmark OD flows and where residual problems remain. The primary comparison
is adjusted versus benchmark flows. Two secondary comparisons help interpret
that result:

- raw mobile-phone-derived flows versus benchmark flows, which describes the
  baseline bias before adjustment
- raw versus adjusted flows, which describes how much each method changes the
  source data

This vignette validates adjustment outputs at three levels: overall flow-table
fit, origin and destination marginals, and individual OD flows. At each level we
combine error metrics, correlation metrics, and visual checks because no single
diagnostic is sufficient.

```{r setup}
input_dir <- dirname(knitr::current_input(dir = TRUE))
debiasr_pkg_root <- normalizePath(file.path(input_dir, ".."), mustWork = FALSE)
common_path <- file.path(input_dir, "_common.R")
if (!file.exists(common_path)) {
  common_path <- if (file.exists("_common.R")) "_common.R" else "vignettes/_common.R"
}
source(common_path)
example_data <- load_workshop_example(n_areas = 25, complete_grid = TRUE)
library(ggplot2)
```

```{r}
method_results <- fit_adjustment_methods(
  example_data,
  include_multilevel = TRUE,
  multilevel_engine = "frequentist"
)
mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od

primary_method <- if ("raking_ratio" %in% names(method_results)) {
  "raking_ratio"
} else {
  names(method_results)[1]
}
```

The examples below use all methods that can be rendered quickly in the vignette.
The multilevel method is included with `model_engine = "frequentist"` so the
validation workflow reflects the formula-aware S1 contract used in the
adjustment vignette, without requiring a slow Bayesian fit during routine
rendering. For detailed scatterplots and row-level diagnostics, the vignette
uses `r primary_method` as a compact example. The same helper calls can be
repeated for any method in `method_results`.

```{r}
safe_cor <- function(x, y, method = "pearson") {
  keep <- is.finite(x) & is.finite(y)
  if (sum(keep) < 2L) {
    return(NA_real_)
  }
  x <- x[keep]
  y <- y[keep]
  if (stats::sd(x) <= 0 || stats::sd(y) <= 0) {
    return(NA_real_)
  }
  suppressWarnings(stats::cor(x, y, method = method))
}

flow_metric_row <- function(data,
                            level,
                            method_name,
                            comparison,
                            estimate_col,
                            reference_col) {
  x <- data[[estimate_col]]
  y <- data[[reference_col]]
  keep <- is.finite(x) & is.finite(y)
  x <- x[keep]
  y <- y[keep]

  if (length(x) == 0L) {
    return(tibble::tibble(
      level = level,
      method = method_name,
      comparison = comparison,
      n = 0L,
      estimate_total = NA_real_,
      reference_total = NA_real_,
      total_difference = NA_real_,
      mean_error = NA_real_,
      mae = NA_real_,
      rmse = NA_real_,
      median_absolute_error = NA_real_,
      mape = NA_real_,
      pearson_r = NA_real_,
      spearman_rho = NA_real_
    ))
  }

  error <- x - y
  denom <- ifelse(y == 0, NA_real_, y)
  mape <- mean(abs(error / denom), na.rm = TRUE)
  if (is.nan(mape)) {
    mape <- NA_real_
  }

  tibble::tibble(
    level = level,
    method = method_name,
    comparison = comparison,
    n = length(error),
    estimate_total = sum(x, na.rm = TRUE),
    reference_total = sum(y, na.rm = TRUE),
    total_difference = sum(x, na.rm = TRUE) - sum(y, na.rm = TRUE),
    mean_error = mean(error, na.rm = TRUE),
    mae = mean(abs(error), na.rm = TRUE),
    rmse = sqrt(mean(error^2, na.rm = TRUE)),
    median_absolute_error = stats::median(abs(error), na.rm = TRUE),
    mape = mape,
    pearson_r = safe_cor(x, y, method = "pearson"),
    spearman_rho = safe_cor(x, y, method = "spearman")
  )
}

comparison_metric_rows <- function(data, level, method_name) {
  dplyr::bind_rows(
    flow_metric_row(
      data = data,
      level = level,
      method_name = method_name,
      comparison = "adjusted_vs_benchmark",
      estimate_col = "flow_adj",
      reference_col = "flow_bench"
    ),
    flow_metric_row(
      data = data,
      level = level,
      method_name = method_name,
      comparison = "raw_vs_benchmark",
      estimate_col = "flow_mpd",
      reference_col = "flow_bench"
    ),
    flow_metric_row(
      data = data,
      level = level,
      method_name = method_name,
      comparison = "raw_vs_adjusted",
      estimate_col = "flow_adj",
      reference_col = "flow_mpd"
    )
  )
}

display_metrics <- function(metrics) {
  metrics |>
    dplyr::mutate(
      dplyr::across(
        dplyr::any_of(c(
          "estimate_total",
          "reference_total",
          "total_difference",
          "mean_error",
          "mae",
          "rmse",
          "median_absolute_error"
        )),
        ~ round(.x, 2)
      ),
      dplyr::across(
        dplyr::any_of(c("mape", "pearson_r", "spearman_rho")),
        ~ round(.x, 3)
      )
    )
}

comparison_palette <- c(
  "Adjusted vs benchmark" = "#2B8CBE",
  "Raw vs benchmark" = "#756BB1",
  "Raw vs adjusted" = "#636363"
)
reference_line_colour <- "#D95F0E"

flow_axis_label <- function(x) {
  dplyr::case_when(
    abs(x) >= 1e6 ~ paste0(round(x / 1e6, 1), "m"),
    abs(x) >= 1e3 ~ paste0(round(x / 1e3, 1), "k"),
    TRUE ~ format(round(x, 0), big.mark = ",", scientific = FALSE, trim = TRUE)
  )
}

theme_validation <- function() {
  theme_minimal(base_size = 11) +
    theme(
      legend.position = "bottom",
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold")
    )
}

build_flow_comparison <- function(method_name, adj_df, mpd_df, benchmark_df) {
  adj_df |>
    dplyr::select(origin, destination, flow_adj) |>
    dplyr::inner_join(
      mpd_df |>
        dplyr::select(origin, destination, flow_mpd = flow),
      by = c("origin", "destination")
    ) |>
    dplyr::inner_join(
      benchmark_df |>
        dplyr::select(origin, destination, flow_bench = flow),
      by = c("origin", "destination")
    ) |>
    dplyr::mutate(method = method_name, .before = 1)
}

build_pair_scatter <- function(data) {
  if (!"level" %in% names(data)) {
    data <- data |>
      dplyr::mutate(level = "individual_flow")
  }

  dplyr::bind_rows(
    data |>
      dplyr::transmute(
        method,
        level,
        comparison = "Adjusted vs benchmark",
        reference_flow = flow_bench,
        compared_flow = flow_adj
      ),
    data |>
      dplyr::transmute(
        method,
        level,
        comparison = "Raw vs benchmark",
        reference_flow = flow_bench,
        compared_flow = flow_mpd
      ),
    data |>
      dplyr::transmute(
        method,
        level,
        comparison = "Raw vs adjusted",
        reference_flow = flow_mpd,
        compared_flow = flow_adj
      )
  ) |>
    dplyr::mutate(
      comparison = factor(
        comparison,
        levels = c(
          "Adjusted vs benchmark",
          "Raw vs benchmark",
          "Raw vs adjusted"
        )
      )
    )
}

plot_pair_scatter <- function(data) {
  ggplot(
    build_pair_scatter(data),
    aes(x = reference_flow, y = compared_flow, colour = comparison)
  ) +
    geom_point(alpha = 0.35, size = 1, na.rm = TRUE) +
    geom_abline(
      intercept = 0,
      slope = 1,
      linetype = "dashed",
      colour = reference_line_colour
    ) +
    facet_wrap(~comparison, scales = "free") +
    scale_colour_manual(values = comparison_palette) +
    scale_x_continuous(labels = flow_axis_label) +
    scale_y_continuous(labels = flow_axis_label) +
    guides(colour = "none") +
    labs(
      x = "Reference flow",
      y = "Compared flow"
    ) +
    theme_validation()
}
```

```{r}
flow_comparison <- dplyr::bind_rows(
  lapply(
    names(method_results),
    function(method_name) {
      build_flow_comparison(
        method_name = method_name,
        adj_df = method_results[[method_name]],
        mpd_df = mpd_od,
        benchmark_df = benchmark_od
      )
    }
  )
)

flow_comparison |>
  dplyr::count(method, name = "od_pairs")
```

```{r}
method_engine <- function(x) {
  engine <- attr(x, "model_engine", exact = TRUE)
  if (is.null(engine)) "deterministic" else engine
}

tibble::tibble(
  method = names(method_results),
  engine = vapply(method_results, method_engine, character(1)),
  rows = vapply(method_results, nrow, integer(1)),
  adjusted_total = vapply(
    method_results,
    function(x) sum(x$flow_adj, na.rm = TRUE),
    numeric(1)
  )
)
```

## Level 1: Overall Validation Metrics

Overall metrics compress each full OD table into a method-level score. Error
metrics answer how far the compared flows are from the reference flows.
Correlation metrics answer whether the compared flows preserve the same
relative pattern across OD pairs.

For `adjusted_vs_benchmark` and `raw_vs_benchmark`, the error sign is compared
against the benchmark. For `raw_vs_adjusted`, the table reports
`adjusted - raw`; this is an adjustment-size diagnostic, not a benchmark
accuracy score.

```{r}
overall_metrics <- dplyr::bind_rows(
  lapply(
    split(flow_comparison, flow_comparison$method),
    function(data) {
      comparison_metric_rows(
        data = data,
        level = "overall",
        method_name = unique(data$method)
      )
    }
  )
)

overall_metrics |>
  dplyr::arrange(comparison, mae) |>
  display_metrics()
```

The package-level helper `validate_flow_overall()` returns the same core
validation family for adjusted-versus-benchmark comparisons. It is the most
direct summary when the question is which adjusted output is closest to the
benchmark.

```{r}
adjusted_benchmark_metrics <- dplyr::bind_rows(
  lapply(
    names(method_results),
    function(method_name) validation_row(
      method_name,
      method_results[[method_name]],
      benchmark_od
    )
  )
) |>
  dplyr::arrange(rmse)

adjusted_benchmark_metrics
```

The figure below compares adjusted and benchmark flows for each method. Points
on the dashed line have equal compared and reference values. Points above the
line are larger than the benchmark; points below it are smaller.

```{r}
flow_comparison |>
  ggplot(aes(x = flow_bench, y = flow_adj)) +
  geom_point(alpha = 0.18, size = 0.7, colour = "#2B8CBE", na.rm = TRUE) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    colour = reference_line_colour
  ) +
  facet_wrap(~method, scales = "free") +
  scale_x_continuous(labels = flow_axis_label) +
  scale_y_continuous(labels = flow_axis_label) +
  labs(
    x = "Benchmark flow",
    y = "Adjusted flow"
  ) +
  theme_validation()
```

This method-level view flags whether each adjustment sits close to the
benchmark as a whole. It does not show where errors concentrate spatially:
a method can have a high overall correlation while still leaving residuals
clustered in particular origins, destinations, or OD pairs. The next two
levels locate that structure.

## Level 2: Origin and Destination Marginals

OD validation can look good overall while still misallocating totals across
origins or destinations. Marginal validation aggregates the OD table first, then
compares those area totals. Origin marginals ask whether total outflows by area
are right. Destination marginals ask whether total inflows by area are right.

```{r}
build_margins <- function(data) {
  dplyr::bind_rows(
    data |>
      dplyr::group_by(method, area = origin) |>
      dplyr::summarise(
        flow_mpd = sum(flow_mpd, na.rm = TRUE),
        flow_adj = sum(flow_adj, na.rm = TRUE),
        flow_bench = sum(flow_bench, na.rm = TRUE),
        .groups = "drop"
      ) |>
      dplyr::mutate(level = "origin_marginal", .after = method),
    data |>
      dplyr::group_by(method, area = destination) |>
      dplyr::summarise(
        flow_mpd = sum(flow_mpd, na.rm = TRUE),
        flow_adj = sum(flow_adj, na.rm = TRUE),
        flow_bench = sum(flow_bench, na.rm = TRUE),
        .groups = "drop"
      ) |>
      dplyr::mutate(level = "destination_marginal", .after = method)
  )
}

marginal_comparison <- build_margins(flow_comparison)

marginal_metrics <- dplyr::bind_rows(
  lapply(
    split(
      marginal_comparison,
      interaction(marginal_comparison$method, marginal_comparison$level, drop = TRUE)
    ),
    function(data) {
      comparison_metric_rows(
        data = data,
        level = unique(data$level),
        method_name = unique(data$method)
      )
    }
  )
)
```

The primary marginal table compares adjusted origin and destination totals to
the benchmark across methods.

```{r}
marginal_metrics |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::arrange(level, mae) |>
  display_metrics()
```

The secondary marginal comparisons for `r primary_method` separate benchmark
accuracy from adjustment magnitude.

```{r}
marginal_metrics |>
  dplyr::filter(method == primary_method) |>
  dplyr::arrange(level, comparison) |>
  display_metrics()
```

The figure below applies the same three comparisons to origin and destination
margins. This makes area-level overestimation or underestimation visible before
drilling into individual OD cells.

```{r}
marginal_comparison |>
  dplyr::filter(method == primary_method) |>
  build_pair_scatter() |>
  ggplot(aes(x = reference_flow, y = compared_flow, colour = comparison)) +
  geom_point(alpha = 0.75, size = 2, na.rm = TRUE) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    colour = reference_line_colour
  ) +
  facet_grid(level ~ comparison, scales = "free") +
  scale_colour_manual(values = comparison_palette) +
  scale_x_continuous(labels = flow_axis_label) +
  scale_y_continuous(labels = flow_axis_label) +
  guides(colour = "none") +
  labs(
    x = "Reference marginal total",
    y = "Compared marginal total"
  ) +
  theme_validation()
```

The table below names the largest remaining marginal residuals for the selected
method. Positive residuals indicate overestimated area totals; negative
residuals indicate underestimated totals. The absolute error reduction column is
positive when adjustment reduced the raw benchmark error for that area.

```{r}
area_names <- example_data$covariates |>
  dplyr::select(area, name)

marginal_patterns <- marginal_comparison |>
  dplyr::filter(method == primary_method) |>
  dplyr::left_join(area_names, by = "area") |>
  dplyr::mutate(
    raw_residual = flow_mpd - flow_bench,
    adjusted_residual = flow_adj - flow_bench,
    abs_adjusted_residual = abs(adjusted_residual),
    abs_error_reduction = abs(raw_residual) - abs(adjusted_residual)
  )

marginal_patterns |>
  dplyr::group_by(level) |>
  dplyr::slice_max(order_by = abs_adjusted_residual, n = 5, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::arrange(level, dplyr::desc(abs_adjusted_residual)) |>
  dplyr::select(
    level,
    area,
    name,
    flow_mpd,
    flow_adj,
    flow_bench,
    adjusted_residual,
    abs_error_reduction
  )
```

These marginal diagnostics are the main place to inspect spatial patterning
without a map. If residuals concentrate in a few origins, the method is not
capturing outflow structure well. If they concentrate in destinations, the
remaining error is more about inflow allocation. When the same areas appear in
both marginal lists, the residual pattern is area-wide rather than only an
origin-side or destination-side issue.

## Level 3: Individual Flow Comparisons

Individual-flow validation audits the OD cells themselves. This level is useful
for identifying large remaining errors, finding flows that changed most during
adjustment, and checking whether those changes moved the OD pair closer to the
benchmark.

```{r}
flow_level_metrics <- overall_metrics |>
  dplyr::filter(method == primary_method) |>
  dplyr::arrange(comparison)

flow_level_metrics |>
  display_metrics()
```

The scatterplot below adapts the workshop comparison logic. Adjusted versus
benchmark shows validation accuracy, raw versus benchmark shows the starting
bias, and raw versus adjusted shows how strongly the method moved the observed
flows.

```{r}
flow_comparison |>
  dplyr::filter(method == primary_method) |>
  plot_pair_scatter()
```

`validate_flow_pairs()` returns the joined OD-pair table for direct inspection.

```{r}
validate_flow_pairs(
  adj_df = method_results[[primary_method]],
  benchmark_od_df = benchmark_od
) |>
  dplyr::arrange(dplyr::desc(abs(diff_adj_benchmark))) |>
  dplyr::select(
    origin,
    destination,
    mpd_flow,
    adj_flow,
    benchmark_flow,
    diff_mpd_benchmark,
    diff_adj_benchmark
  ) |>
  dplyr::slice_head(n = 10)
```

`validate_flow_residuals()` adds signed movement, absolute residual reduction,
and residual outlier diagnostics. The signed residual movement is:

$$
(F^{bench}_{ij} - F^{mpd}_{ij}) -
(F^{bench}_{ij} - F^{adj}_{ij})
$$

The absolute residual-reduction metric is easier for method comparison:
positive values mean the adjusted flow is closer to the benchmark.

```{r}
residuals_primary <- validate_flow_residuals(
  adj_df = method_results[[primary_method]],
  benchmark_od_df = benchmark_od,
  method_name = primary_method,
  top_n = 10
)

residuals_primary$summary
```

```{r}
primary_flow <- flow_comparison |>
  dplyr::filter(method == primary_method) |>
  dplyr::mutate(
    raw_residual = flow_mpd - flow_bench,
    adjusted_residual = flow_adj - flow_bench,
    adjustment = flow_adj - flow_mpd,
    abs_error_reduction = abs(raw_residual) - abs(adjusted_residual),
    movement = dplyr::case_when(
      abs_error_reduction > 0 ~ "closer_to_benchmark",
      abs_error_reduction < 0 ~ "farther_from_benchmark",
      TRUE ~ "unchanged"
    )
  )

primary_flow |>
  dplyr::arrange(dplyr::desc(abs(adjustment))) |>
  dplyr::select(
    origin,
    destination,
    flow_mpd,
    flow_adj,
    flow_bench,
    adjustment,
    raw_residual,
    adjusted_residual,
    abs_error_reduction,
    movement
  ) |>
  dplyr::slice_head(n = 10)
```

The residual histogram below shows the distribution of adjusted-minus-benchmark
errors for the selected method.

```{r}
primary_flow |>
  ggplot(aes(x = adjusted_residual)) +
  geom_vline(xintercept = 0, colour = "grey65") +
  geom_histogram(bins = 30, fill = "#2B8CBE", colour = "white") +
  scale_x_continuous(labels = flow_axis_label) +
  labs(
    x = "Adjusted flow minus benchmark flow",
    y = "OD-pair count"
  ) +
  theme_validation()
```

The residual histogram shows whether remaining errors are centred near zero or
skewed toward systematic overestimation or underestimation. The ranked table
shows which individual flows were most changed by adjustment and whether those
large movements improved benchmark correspondence.

### Residual Structure and Spatial Allocation

The previous sections compare levels of aggregation. Two existing validation
helpers add structure to the interpretation. `validate_flow_residual_structure()`
checks whether remaining flow residuals are associated with benchmark flow size
or with a selected area covariate. `validate_flow_distribution()` checks whether
each origin allocates its destinations similarly to the benchmark.

```{r}
structure_primary <- validate_flow_residual_structure(
  adj_df = method_results[[primary_method]],
  benchmark_od_df = benchmark_od,
  method_name = primary_method,
  covariate_df = example_data$covariates,
  covariate_col = "rural_pct"
)

structure_primary$summary
```

```{r}
distribution_primary <- validate_flow_distribution(
  adj_df = method_results[[primary_method]],
  benchmark_od_df = benchmark_od,
  method_name = primary_method,
  weight_by = "benchmark_origin_total"
)

distribution_primary$summary
```

Lower KL divergence and Jensen-Shannon divergence mean the adjusted flows
allocate each origin's total flow across destinations more like the benchmark.
These allocation metrics complement the marginal checks: a method can match
origin totals while still sending those totals to the wrong destinations.

## Bayesian Validation

When `adjust_multilevel_bayes()` is rendered with real OD distance and a
substantive formula, validate it with the same three-level structure used above.
Bayesian validation should report predictive performance, residual structure,
and runtime or model diagnostics attached to the result object, because its main
innovation is a formula-based, uncertainty-aware multilevel structure rather
than a fixed adjustment rule.
