---
title: "Validation"
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
.template-quarto table {
  font-size: 14.5px;
}
</style>
```

## Aim

This vignette describes the range of validation methods included in `debiasR`, to assess the extent of difference between our adjusted mobile-phone-derived flow estimates and target benchmark flows.

The methods assess differences between:

- adjusted *versus* benchmark flows, to evaluate the final adjustment.
- raw mobile-phone-derived *versus* benchmark flows, to evaluate the
  extent of bias before adjustment.
- raw *versus* adjusted flows, to evaluate the extent of adjustment made to the raw data.

The validation methods included in `debiasR` allows the assessment of adjustment outputs at three levels: overall, OD marginal and individual OD-pair adjustment. Tree type of assessment tools are included:
 error metrics, correlation coefficients and visual tools. These diagnostics provide flexibility to assess particular dimensions of the data which may vary depending on the target use of the adjusted flows.

## Intuition

A practical way to think of assessing the output of the various adjustment methods is to adopt a overall system-wide perspective of the flow system, move to an intermediate and then more granular origin-destination pair level. We use three levels.

The **Level 1** summarises the performance of flow adjustment methods across the whole origin-destination system.

The **Level 2** assesses the performance of flow adjustment methods by origin and destination totals.

The **Level 3** provides the performance of flow adjustment methods between individual origin-destination pairs.

Correlation coefficients offer a measure of correspondence between, for example, adjusted and benchmark flows, and they are better interpreted as a general measure. Error measures provide a measure of the level of difference between these sets of flows. They provide a general measure to assess the precision of the flow adjustment across the whole network of flows, but they also offer a way to measure and assess the precision of the flow adjustment for individual origin-destination pairs.

## Validation measures

Let $F^{adj}_{ij}$ be the adjusted flow from origin $i$ to destination
$j$, and let $F^{bench}_{ij}$ be the benchmark flow for the same OD pair.
Define the error as
$e_{ij} = F^{adj}_{ij} - F^{bench}_{ij}$. Positive errors mean the adjusted
flow is larger than the benchmark. Negative errors mean it is smaller. The
table uses adjusted-versus-benchmark notation.[^validation-measure-reuse]

| Metric | Description | Equation |
|:--|:--|:--|
| Mean error | Average signed error; shows whether flows are overestimated or underestimated on average. | $\frac{1}{n}\sum_{ij} e_{ij}$ |
| Mean absolute error | Average absolute error; shows the typical size of the error, ignoring sign. | $\frac{1}{n}\sum_{ij} \lvert e_{ij} \rvert$ |
| Root mean squared error | Error metric that gives more weight to large errors. | $\sqrt{\frac{1}{n}\sum_{ij} e_{ij}^{2}}$ |
| Median absolute error | Middle absolute error; less affected by a small number of very large errors. | $\mathrm{median}(\lvert e_{ij} \rvert)$ |
| Mean absolute percentage error | Average absolute error as a percentage of the benchmark flow. | $\frac{100}{n}\sum_{ij}\lvert \frac{e_{ij}}{F^{bench}_{ij}} \rvert$ |
| Pearson's r | Linear correlation; shows whether large benchmark flows also tend to be large adjusted flows. | $\mathrm{cor}(F^{adj}_{ij}, F^{bench}_{ij})$ |
| Spearman's rho | Rank correlation; shows whether the ordering of flows is similar. | $\mathrm{cor}(\mathrm{rank}(F^{adj}_{ij}), \mathrm{rank}(F^{bench}_{ij}))$ |

[^validation-measure-reuse]: The equations are written for
    adjusted-versus-benchmark flows. The same formulas can be reused for
    raw-versus-benchmark or raw-versus-adjusted comparisons by replacing
    $F^{adj}_{ij}$ and $F^{bench}_{ij}$ with the two flow columns being
    compared.

## Getting ready

The README shows how to install `debiasR` and the companion `debiasRdata`
package from GitHub. We start by loading those packages.

```{r load-packages}
library(debiasR)
library(debiasRdata)
```

This vignette also uses validation helper functions defined in
`vignettes/_validation_helpers.R` to format tables and plots. When working from
the package source, you can load them with
`source("vignettes/_validation_helpers.R")`.

```{r load-validation-helpers, include=FALSE}
helper_candidates <- c(
  "vignettes/_validation_helpers.R",
  "_validation_helpers.R",
  "../vignettes/_validation_helpers.R",
  "../../vignettes/_validation_helpers.R"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]
if (is.na(helper_file)) {
  stop(
    "Could not find vignettes/_validation_helpers.R. ",
    "Set the working directory to the debiasR package root or vignettes folder."
  )
}
source(helper_file)
```

Finally, we load a small LAD travel-to-work example from `debiasRdata` through
the exported `debiasR_example_data()` helper.

```{r load-validation-data}
example_data <- debiasR_example_data(
  n_areas = 25,
  complete_grid = TRUE
)
```

In this vignette, we will illustrate the validation methods in `debiasR` comparing the two adjustment methods. We compare the inverse penetration and selection rate II methods. First,
we separately organise the raw mobile-phone-derived data and benchmark data. We then apply inverse penetration and selection rate II, and
join the raw, adjusted and benchmark flows into one data frame for comparison.

```{r}
mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od

method_results <- list(
  inverse_penetration = debiasR::adjust_inverse_penetration(
    mpd_od_df = mpd_od,
    coverage_df = example_data$coverage,
    weight_by = "both"
  ),
  selection_rate2 = debiasR::adjust_selection_rate2(
    mpd_od_df = mpd_od,
    coverage_df = example_data$coverage,
    weight_by = "origin",
    benchmark_od_df = benchmark_od,
    calibration_aggregate = "origin"
  )
)

flow_comparison <- validation_build_flow_comparison(
  method_results = method_results,
  mpd_df = mpd_od,
  benchmark_df = benchmark_od
)
```

## Level 1: Overall validation

This level provides a single number summary metric for the whole system of origin-destination flows. Each row in the output table reports a measure for one method and one data-type comparison (e.g. adjusted vs benchmark flows or raw vs adjusted flows).

The main results are the adjusted-versus-benchmark rows. These rows indicate
whether the adjusted flows are close to the benchmark flows. The
raw-versus-benchmark rows show the starting point before adjustment. The
raw-versus-adjusted rows show how much the method adjusted the original data.

```{r}
overall_metrics <- validation_metrics_by_method(
  data = flow_comparison,
  level = "overall"
)

overall_metrics |>
  dplyr::arrange(comparison, mae) |>
  validation_display_metrics()
```

**Interpretation**

*Error metrics*. Lower mean absolute error, RMSE and median
absolute error mean that the set of flows in the comparison are *closer in size*. The sign of mean error indicates the direction of this difference, with a positive score indicating overestimation on average, and
a negative value indicating underestimation on average.

*Correlation coefficients*. A high Pearson correlation means the
method represents the same broad pattern of large and small flows as the benchmark.
A high Spearman correlation indicates that the ranking of flows is similar. A key point here is that the focus of correlation coefficients is to offer an assessment of the overall correspondence or proportionality between two distributions. Hence, a correlation coefficient can be high and still have large errors as correlation may provide a good assessment of the pattern but of the scale of flows.

```{r}
level1_primary <- overall_metrics |>
  dplyr::filter(comparison == "adjusted_vs_benchmark")

level1_lowest_mae <- level1_primary |>
  dplyr::slice_min(mae, n = 1, with_ties = FALSE) |>
  dplyr::pull(method_label)

level1_highest_r <- level1_primary |>
  dplyr::slice_max(pearson_r, n = 1, with_ties = FALSE) |>
  dplyr::pull(method_label)
```

## Level 2: Marginal validation

Level 2 compares geographical area totals across origins and destinations. It complements the level-1 assessment. The adjusted flows can look reasonable overall while still significantly over- or under-estimating specific origins and destinations.
An **origin marginal** adds together all flows
from an individual area. We therefore assess the performance of an adjustment method to produce adjusted flows which are close to the benchmark total outflow. A
**destination marginal** adds together all flows arriving in each area. In this case, we assess the performance of an adjustment method to produce adjusted flows which are close to the benchmark total inflow.

```{r}
marginal_comparison <- validation_build_margins(flow_comparison)
marginal_metrics <- validation_metrics_by_method_level(marginal_comparison)

marginal_metrics |>
  dplyr::arrange(level, comparison, mae) |>
  validation_display_metrics()
```

We can use scatterplots to explore the data underpinning the results. We plotted the three sets of comparisons described at the start of this vignette: adjusted vs benchmark, raw vs benchmark, and raw vs adjusted flows.
Each point represents an area total of inflows or outflows. Points near the dashed line have similar flow values. Points above the line are larger in the
flow dataset on the vertical axis. Points below the line are smaller in that
same dataset.

```{r}
validation_plot_margin_scatter(
  marginal_comparison,
  level = "destination_marginal"
)
```

```{r}
validation_plot_margin_scatter(
  marginal_comparison,
  level = "origin_marginal"
)
```

We can also use the resulting assessment metrics to identify the largest (*remaining*) marginal errors post adjustment. It helps determine
if the remaining error is concentrated in a small number of areas.

```{r}
area_names <- example_data$covariates |>
  dplyr::select(area, name)

marginal_patterns <- marginal_comparison |>
  dplyr::left_join(area_names, by = "area") |>
  dplyr::mutate(
    raw_residual = flow_mpd - flow_bench,
    adjusted_residual = flow_adj - flow_bench,
    abs_adjusted_residual = abs(adjusted_residual),
    abs_error_reduction = abs(raw_residual) - abs(adjusted_residual)
  )

marginal_patterns |>
  dplyr::group_by(method_label, level) |>
  dplyr::slice_max(order_by = abs_adjusted_residual, n = 3, with_ties = FALSE) |>
  dplyr::ungroup() |>
  dplyr::arrange(method_label, level, dplyr::desc(abs_adjusted_residual)) |>
  dplyr::select(
    method = method_label,
    level,
    name,
    flow_adj,
    flow_bench,
    adjusted_residual,
    abs_error_reduction
  )
```

```{r}
level2_best <- marginal_metrics |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::group_by(level) |>
  dplyr::slice_min(mae, n = 1, with_ties = FALSE) |>
  dplyr::ungroup()

level2_origin_best <- level2_best |>
  dplyr::filter(level == "origin_marginal") |>
  dplyr::pull(method_label)

level2_destination_best <- level2_best |>
  dplyr::filter(level == "destination_marginal") |>
  dplyr::pull(method_label)
```

In this example, `r level2_origin_best` is closer to benchmark origin totals by
mean absolute error, while `r level2_destination_best` is closer to benchmark
destination totals. We therefore have clear evidence as the same method is closer at both margins. If the two margins favoured different methods,
the choice would depend on whether origin outflows or destination inflows matter
more for the analysis.

## Level 3: Origin-destination pair validation

Level 3 focuses on comparing individual origin-destination pairs. It zooms in to provide more granular detail and identify specific origin-destination interactions where large local errors are brought to light.

```{r}
flow_level_metrics <- validation_metrics_by_method(
  data = flow_comparison,
  level = "individual_flow"
)

flow_level_metrics |>
  dplyr::arrange(comparison, mae) |>
  validation_display_metrics()
```

*Scatterplots*

We produce a similar scatterplot as we did for marginal totals.
Here each point is a single origin-destination pair. Points close to the dashed line are more similar. Points far from the line are more dissimilar reflecting differences from the reference origin-destination pair.

```{r}
validation_plot_od_scatter(flow_comparison)
```

*Largest residuals and heatmap*

We can also rank individual origin-destination pairs by the largest remaining
adjusted-minus-benchmark residuals. Positive values mean the adjusted estimate
is larger than the benchmark. Negative values mean it is smaller.

```{r}
validation_rank_largest_residuals(flow_comparison, n = 12)
```

More usefully, we can create a compact heatmap to understand and summarise the full set of origin-destination pair residuals by standard
deviation (SD) bands. The residual for each adjustment method is the
adjusted-minus-benchmark difference, $F^{adj}_{ij} - F^{bench}_{ij}$. The SD
bands are defined from the pooled adjusted-minus-benchmark residuals across
the two adjustment methods, so a band means the same thing in each column.

Each cell is the percentage of origin-destination pairs in that residual band. For example, a value of 0.16 means that 0.16% of origin-destination pairs in
that column fall in the residual-size band named on the row. Lower percentages
in the high-SD rows indicate fewer large outliers.

```{r}
residual_heatmap <- validation_build_residual_heatmap(flow_comparison)
validation_plot_residual_heatmap(residual_heatmap)
```

The heatmap also includes the unadjusted raw MPD as a baseline. This baseline uses
the raw-minus-benchmark difference, $F^{mpd}_{ij} - F^{bench}_{ij}$, and is
classified using the same adjusted-residual SD bands. The heatmap offers a way to enable visual
comparison. If an adjustment reduces the share of flows above 2, 3 or 4
standard deviations, it contributes to reducing the number of large origin-destination-pair errors.

```{r}
level3_best <- flow_level_metrics |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::slice_min(mae, n = 1, with_ties = FALSE) |>
  dplyr::pull(method_label)

large_outlier_summary <- residual_heatmap |>
  dplyr::filter(residual_band %in% c("Greater than 4.0 SD", "3.0 to 4.0 SD")) |>
  dplyr::group_by(method_label) |>
  dplyr::summarise(large_outlier_share = sum(share), .groups = "drop") |>
  dplyr::filter(method_label != "Unadjusted") |>
  dplyr::slice_min(large_outlier_share, n = 1, with_ties = FALSE)

level3_outlier_best <- large_outlier_summary |>
  dplyr::pull(method_label)
```

At the individual origin-destination pair level, `r level3_best` has the lower mean absolute error
against the benchmark in this example. Looking only at the largest residual
bands, `r level3_outlier_best` has the smaller share of OD pairs above three
standard deviations.

## Recommendation

Our recommendation is to adopt a sequence approach to validation. First, we suggest to analyse the overall metrics to assess
if the resulting adjusted flows are broadly closer to the benchmark. Second, we recommend evaluating origin
and destination marginal diagnostics to determine the precision of correction for area totals, and identify systematic under- or over-estimation for specific areas. Third,
we suggest inspecting individual origin-destinations pairs to evaluate corridors of special interest, such as origin-destinations pairs that typically have large amounts of interaction.

The validation methods presented here offer quantitative evidence to assist in the selection of which method to use. However, this decision is not binary and depends on a combination of qualitative and quantitative factors:

- *Data availability*. Adjustment methods require different levels of inputs. Simpler deterministic adjustment methods generally require fewer data inputs. Yet, they often require target benchmark flow data, and these are precisely the data inputs that we often do not have. Benchmark flow data often come from censuses, which are considered the gold standard, but these data are generally collected every ten or five years and are outdated when a crisis occurs or a new evaluation of mobility is needed to support humanitarian operations, planning or policy decisions. Hence, using an adjustment method requiring census benchmark data would produce misleading results if it represented the mobility network at the time of the census rather than the mobility system during the period of interest. An adjustment method, such as the proposed Bayesian multilevel modelling approach, that is not reliant on such inputs may be more appropriate in such scenarios.
- *Required measurement precision*. Adjustment methods have varying levels of precision in terms of scale reproducing population numbers. Historically, prior work has focused on satisfying the requirement of proportionality by assessing the linear correlation between raw and/or adjusted mobile-phone-derived flows and benchmark flows. Yet, such assessment does not guarantee that raw or adjusted flows provide an accurate estimate in population numbers. They often still represent number of devices, rather than people. Except for the selection rate II method, all the methods in `debiasR` seek to provide population numbers.
- *Time constraint*. The adjustment methods in `debiasR` may demand more or less time. Such demand will depend on processes such as data gathering, data processing and calibrating adjustment methods. These demands normally intersect and often come into conflict with the speed at which flow estimates are needed. As suggested above, the use of digital trace data is often triggered by the rapid and urgent need to have "rough" population numbers to inform humanitarian, planning or policy decisions, to address a crisis or new intervention.
