---
title: "Visualising origin-destination flows"
description: >
  Prepare origin-destination flows and area centroids for exploratory
  interactive flow visualisation.
vignette: >
  %\VignetteIndexEntry{Visualising origin-destination flows}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Aim

This vignette shows how origin-destination (OD) flows can be prepared for
visual inspection. The aim is not to replace the formal validation diagnostics
in `debiasR`, but to give a quick way to see which origin-destination corridors
carry the largest flows and where those corridors are located.

The examples use benchmark Census travel-to-work flows between Local Authority
Districts (LADs) from `debiasRdata`, together with LAD centroid coordinates.
The same structure can be used for observed mobile-phone-derived flows or
adjusted flows returned by `debiasR`.

## Getting ready

The README shows how to install `debiasR` and the companion `debiasRdata`
package from GitHub. This vignette also uses `dplyr` for data preparation and
the `mapgl` package for interactive visualisation.

```{r load-packages}
library(debiasR)
library(debiasRdata)
```

The interactive map uses the `mapgl` package. You can install it from CRAN with:

```{r eval=FALSE}
install.packages("mapgl")
```

If the package is not installed, the preparation code below still shows the data
structure needed for the map.

```{r mapgl-availability, include=FALSE}
mapgl_available <- requireNamespace("mapgl", quietly = TRUE)
```

We start with the LAD benchmark OD-flow table and the LAD centroid table. The
OD-flow table stores one row per origin-destination pair. The centroid table
stores one coordinate pair per LAD.

```{r load-flow-data}
benchmark_flows <- debiasRdata::census_lad_OD_travel2work
centroids <- debiasRdata::lad_centroids
```

## Identify the largest flows

Flow maps can become difficult to read if every OD pair is plotted at once. A
practical first step is to inspect the largest flows and then pass a bounded
set of between-area flows to the interactive map.

```{r largest-flows}
largest_flows <- benchmark_flows |>
  dplyr::filter(origin != destination) |>
  dplyr::slice_max(order_by = flow, n = 40, with_ties = FALSE)

largest_flows |>
  dplyr::slice_head(n = 10)
```

The table above shows the largest between-area flows. Within-area flows are
excluded here because they have the same origin and destination.

## Prepare map inputs

`mapgl::add_flowmap()` expects two plain data frames:

- `locations`, with one row per area and columns identifying the area,
  longitude and latitude.
- `flows`, with one row per OD pair and columns identifying the origin,
  destination and flow count.

We create those inputs from the LAD centroid and Census OD-flow tables.

```{r prepare-map-inputs}
locations <- data.frame(
  id = centroids$area,
  name = centroids$name,
  lon = centroids$longitude,
  lat = centroids$latitude,
  stringsAsFactors = FALSE
)

flows <- benchmark_flows |>
  dplyr::filter(origin != destination, flow > 0) |>
  dplyr::slice_max(order_by = flow, n = 10000, with_ties = FALSE) |>
  dplyr::left_join(
    centroids |>
      dplyr::select(origin = area, origin_name = name),
    by = "origin"
  ) |>
  dplyr::left_join(
    centroids |>
      dplyr::select(destination = area, dest_name = name),
    by = "destination"
  ) |>
  dplyr::transmute(
    origin = origin,
    dest = destination,
    origin_name = origin_name,
    dest_name = dest_name,
    count = flow
  )

head(locations)
head(flows)
```

The `locations` data frame links LAD codes to coordinates. The `flows` data
frame stores the OD pairs to display. Here we keep the 10,000 largest
between-area Census flows to keep the browser widget responsive.

## Interactive flow map

The map below uses `mapgl::add_flowmap()` to display LAD travel-to-work flows.
The flow layer is interactive: you can zoom, pan and inspect flow corridors in
the rendered HTML page.

```{r interactive-flow-map, eval=mapgl_available}
mapgl::maplibre(
  style = mapgl::carto_style("dark-matter"),
  center = c(-2.0, 54.0),
  zoom = 5
) |>
  mapgl::add_flowmap(
    id = "census-lad-flows",
    locations = locations,
    flows = flows,
    flow_color_scheme = "Teal",
    flow_dark_mode = TRUE,
    flow_lines_rendering_mode = "curved",
    flow_line_thickness_scale = 1.1,
    flow_clustering_enabled = TRUE,
    flow_max_top_flows_display_num = 10000,
    tooltip = list(
      location = "{name}",
      flow = "{origin_name} -> {dest_name}<br>{count}"
    )
  )
```

The map highlights major benchmark travel-to-work corridors. This kind of
visualisation is useful for checking whether a small number of corridors
dominate the OD system, identifying flows that may deserve closer inspection
and comparing raw, adjusted or benchmark flow patterns.

## Visualise other flow columns

The same workflow can be used for observed mobile-phone-derived flows or
adjusted flows. The only requirement is that the table has an origin column, a
destination column and a numeric flow column. For example, after using an
adjustment method from `debiasR`, replace `benchmark_flows` with the adjusted
table and select `flow_adj` as the plotted value.

```{r eval=FALSE}
adjusted_flows <- debiasR::adjust_inverse_penetration(
  mpd_od_df = debiasRdata::lad_OD_travel2work,
  coverage_df = debiasRdata::coverage_lad,
  weight_by = "origin"
)

largest_adjusted_flows <- adjusted_flows |>
  dplyr::filter(origin != destination) |>
  dplyr::slice_max(order_by = flow_adj, n = 40, with_ties = FALSE)
```

In a validation workflow, visualisation is most useful when it is paired with
formal metrics. A map can show where large flows are located, but it does not
quantify whether adjusted flows are closer to a benchmark. For that, use the
validation methods in the validation vignette.

## Alternative visualisation packages

Other packages can also be used for flow visualisation. [`{flowmapper}`](https://cran.r-project.org/package=flowmapper) is a
possible route for static flow maps, while [`{flowmapblue}`](https://flowmapblue.github.io/flowmapblue.R/) provides interactive
flow maps but is less current than the emerging `mapgl` route.
