---
title: "Advanced Bayesian multilevel adjustment"
vignette: >
  %\VignetteIndexEntry{Advanced Bayesian multilevel adjustment}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
main table {
  font-size: 0.9rem;
  line-height: 1.45;
}

main table th,
main table td {
  vertical-align: top;
}
</style>
```

## Aim

This vignette explains the advanced Bayesian options in
`adjust_multilevel_bayes()`. The main adjustment vignette introduces the
recommended default, `observation_model = "coverage_offset"`, because it is the
most practical starting point for routine adjustment. Start with
[Adjusting origin-destination flows](v06-adjusting-biases.html) if you want the
default workflow first. Here we compare all Bayesian observation-model choices
and explain when the experimental latent two-level backend is useful.

The guiding principle is progressive complexity: use the simplest model that
matches the data structure and inferential goal.

## Why more than one Bayesian model?

`adjust_multilevel_bayes()` supports three Bayesian observation-model ideas:

```{r echo=FALSE}
bayesian_modes <- tibble::tribble(
  ~Mode, ~Status, ~Main_use, ~Core_idea,
  "`coverage_offset`",
  "Recommended default",
  "Routine true-flow adjustment and teaching examples",
  "Treat MPD flows as coverage-scaled observations of true flows, with coverage fixed as an offset.",
  "`latent_two_level`",
  "Experimental advanced backend",
  "Repeated source/time data where multiple MPD observations can share a hidden OD or OD-time state",
  "Create a latent-flow state and treat repeated MPD rows as observations of that state.",
  "`reduced_form`",
  "Legacy/backward-compatible",
  "Reproducing earlier analyses or sensitivity checks on the MPD observation scale",
  "Fit observed MPD flows directly and adjust by setting fitted bias terms to zero."
)

knitr::kable(bayesian_modes)
```

The default coverage-offset model and the latent two-level model both aim to
return adjusted flows on a true-flow scale. They differ in what is estimated
inside the model. The coverage-offset model estimates the true-flow scale by
removing a fixed coverage offset from an observed-flow likelihood. The latent
backend estimates a shared latent-flow intensity directly, so it is most useful
when the same underlying OD or OD-time flow is observed by multiple sources.

## Choosing a Bayesian model

```{r echo=FALSE}
decision_table <- tibble::tribble(
  ~Data_or_goal, ~Recommended_mode, ~Reason,
  "One source and one time period, or a small teaching example",
  "`coverage_offset`",
  "The latent split is weakly identified when each OD pair is observed once.",
  "Complete-grid prediction over supplied OD rows",
  "`coverage_offset`",
  "The model can predict true-flow scale for supplied rows using the coverage offset and mobility formula.",
  "Multiple sources observing the same OD pairs",
  "`latent_two_level`",
  "Repeated source observations can share an OD-level latent state.",
  "Multiple sources and multiple time periods",
  "`latent_two_level` with care",
  "Repeated source observations can share OD-time states, but time and source effects need careful interpretation.",
  "Fast design, tests or method comparison",
  "`model_engine = \"frequentist\"` with non-latent modes",
  "The frequentist engine preserves the S1-S4 data contract but does not fit the latent Bayesian backend.",
  "Backward compatibility with earlier reduced-form analyses",
  "`reduced_form`",
  "This preserves the older MPD-scale counterfactual behaviour."
)

knitr::kable(decision_table)
```

As a rule of thumb, start with `coverage_offset`. Move to
`latent_two_level` only when the repeated-observation structure gives the model
real information about the hidden flow state.

## Default coverage-offset model

The default true-flow model is:

$$
F^{mpd}_{ij} \sim \mathrm{Poisson}(q_{ij}F^{true}_{ij})
$$

$$
\log E(F^{mpd}_{ij}) = \log(q_{ij}) + \eta^{true}_{ij}
$$

where $q_{ij}$ is derived from active-user coverage. With
`coverage_scale = "origin"`, $q_{ij} = c_i = U_i / P_i$. The returned
`flow_adj` is a posterior summary of:

$$
F^{true,pred}_{ij} = \exp(\eta^{true}_{ij})
$$

This is the model explained in the main adjustment vignette. It is usually the
best option for S1 examples and for users who want readable, reproducible
Bayesian adjustment without modelling source-specific observation layers.

```{r eval=FALSE}
adj_default <- adjust_multilevel_bayes(
  mpd_od_df = mpd_od,
  coverage_df = coverage,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  model_family = "poisson"
)
```

With richer data, the coverage-offset model can include multilevel structure in
the true-flow formula. Random-effect terms represent origin, destination and
corridor structure: `(1 | origin)` for an origin intercept, `(1 | destination)`
for a destination intercept, and `(1 | od_id)` for an OD-corridor intercept.
Use these terms when the data contain enough repeated information to estimate
the pooling structure.

```{r eval=FALSE}
adj_default_multilevel <- adjust_multilevel_bayes(
  mpd_od_df = mpd_od,
  coverage_df = coverage,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance +
    (1 | origin) + (1 | destination) + (1 | od_id),
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

## Latent two-level backend

The latent two-level backend is for repeated source/time settings. It keeps
the true-flow target but introduces a shared latent-flow state. Conceptually:

$$
Y_{ijst} \sim \mathrm{Poisson}(\mu_{ijst})
$$

$$
\log(\mu_{ijst}) =
\log(q_{ijst}) + \theta_{g(i,j,t)} + z_{ijst}\gamma
$$

where:

- $Y_{ijst}$ is the observed MPD flow for origin $i$, destination $j$, source
  $s$ and time $t$.
- $q_{ijst}$ is the coverage-derived observation probability.
- $\lambda^{true}_{g(i,j,t)}$ is the shared latent true-flow intensity for an
  OD or OD-time group.
- $z_{ijst}\gamma$ represents observation-layer structure such as coverage,
  source and time effects.

In the current experimental backend, `adjust_multilevel_bayes()` represents
this shared state through `latent_flow_id` and estimates it with the custom
`stan_latent` backend. The returned `flow_adj` and `flow_true_pred` summarize
posterior draws of the latent true-flow intensity, while `flow_mpd_pred`
summarizes the source/time-specific MPD observation mean.

```{r eval=FALSE}
adj_latent <- adjust_multilevel_bayes(
  mpd_od_df = mpd_repeated,
  coverage_df = coverage_repeated,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "latent_two_level",
  backend = "auto",
  coverage_scale = "origin",
  latent_flow_unit = "auto",
  model_engine = "bayesian",
  scenario = "s3",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "source",
  prediction_scope = "observed",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 1000,
  chains = 4,
  seed = 123,
  refresh = 0
)
```

The returned table includes `latent_flow_id`, `latent_flow_unit`,
`flow_adj_mean`, `flow_adj_median`, and 95% interval columns for the true-flow
and MPD observation scales. The result metadata also records the custom backend,
the number of latent states, the true-flow and observation formulas, latent
prior/sampler controls, ignored formula random-effect terms, and identifiability
notes.
Because this mode relies on repeated observations, it is most informative for
S3 and S4 designs. In S1 and S2, use it mainly as a sensitivity or compatibility
check.

## Source-time structures

The same S1-S4 scenario language is used across the Bayesian and frequentist
engines.

```{r echo=FALSE}
source_time_table <- tibble::tribble(
  ~Structure, ~Scenario, ~Latent_state, ~Bayesian_guidance,
  "One source, one time",
  "`s1`",
  "OD state, but each state is observed once",
  "Prefer `coverage_offset`; latent mode is weakly identified.",
  "One source, multiple times",
  "`s2`",
  "OD or OD-time state, depending on the research question",
  "Use care: time variation can reflect true mobility change rather than observation bias.",
  "Multiple sources, one time",
  "`s3`",
  "OD state shared across sources",
  "Best first use case for the experimental `latent_two_level` backend.",
  "Multiple sources, multiple times",
  "`s4`",
  "OD-time state shared across sources within each time period",
  "Useful but more demanding; inspect source and time diagnostics carefully."
)

knitr::kable(source_time_table)
```

## Reduced-form compatibility mode

The reduced-form mode preserves the earlier observed-flow workflow:

```{r eval=FALSE}
adj_reduced_form <- adjust_multilevel_bayes(
  mpd_od_df = mpd_od,
  coverage_df = coverage,
  covariates_df = covariates,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "mpd_counterfactual",
  observation_model = "reduced_form",
  model_engine = "bayesian",
  model_family = "poisson"
)
```

Use this mode only when you need to reproduce older analyses or compare the
newer true-flow scale against the earlier MPD-scale counterfactual. For most
new analyses, prefer `coverage_offset`.

## Diagnostics

For any Bayesian fit, inspect:

- sampler diagnostics from `attr(result, "diagnostics")`;
- model metadata from `attr(result, "result_metadata")`;
- model terms from `attr(result, "model_terms")`;
- posterior summaries and, when requested, draw-level outputs;
- whether the source/time structure supplies enough repeated observations for
  the chosen observation model.

```{r eval=FALSE}
attr(adj_latent, "result_metadata")
attr(adj_latent, "diagnostics")
attr(adj_latent, "model_terms")
```

For the custom latent backend, `attr(adj_latent, "diagnostics")$convergence`
reports divergence and treedepth rates, acceptance summaries, effective sample
size, and R-hat for core model parameters. The `latent_*_prior_scale`,
`latent_phi_prior_rate`, `latent_adapt_delta`, `latent_max_treedepth`, and
`latent_rng_eta_max` arguments are available for sensitivity checks when the
default weakly regularizing settings are not appropriate.

The latent backend should be treated as an advanced sensitivity model until
larger repeated-source/time examples, stronger prior checks, and richer
posterior predictive diagnostics are validated.

## Practical recommendation

Use `coverage_offset` as the default Bayesian implementation. It is readable,
fast enough for small workflows, and directly returns a true-flow scale from a
coverage-aware observation model.

Use `latent_two_level` when the data structure justifies it: especially
multiple MPD sources observing the same OD or OD-time state. Keep the older
`reduced_form` mode for reproducibility and backward compatibility.
