---
title: "Stage 3 Measure Bias Review Notebook"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This notebook reviews the Stage 3 measure-bias diagnostics implemented in `validate_bias_residual_structure()`.

It uses the package's deterministic simulated data to inspect:

- `measure_bias()` coverage quantities,
- Stage 3 coverage residual definitions,
- optional Moran's I from a plain neighbour-link table,
- benchmark origin-flow and destination-flow correlations,
- covariate correlations,
- map-ready and plot-ready outputs.

## Setup

```{r}
find_pkg_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "DESCRIPTION"))) {
      return(path)
    }

    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not find package root containing DESCRIPTION.")
    }
    path <- parent
  }
}

pkg_root <- find_pkg_root()
setwd(pkg_root)

if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(debiasR)
}

library(dplyr)
library(tibble)

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) {
  library(ggplot2)
}
```

```{r}
data("simulated_coverage", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_covariates", package = "debiasR")

tibble(
  dataset = c(
    "simulated_coverage",
    "simulated_benchmark.od",
    "simulated_covariates"
  ),
  rows = c(
    nrow(simulated_coverage),
    nrow(simulated_benchmark.od),
    nrow(simulated_covariates)
  ),
  columns = c(
    paste(names(simulated_coverage), collapse = ", "),
    paste(names(simulated_benchmark.od), collapse = ", "),
    paste(names(simulated_covariates), collapse = ", ")
  )
)
```

## Coverage Quantities

`measure_bias()` remains the compact coverage calculator. It returns `coverage_score`, `coverage_bias`, and the backward-compatible `bias` alias.

```{r}
coverage_measures <- measure_bias(simulated_coverage)

coverage_measures |>
  select(origin, population, user_count, coverage_score, coverage_bias, bias) |>
  slice_head(n = 10)
```

```{r}
coverage_measures |>
  summarise(
    areas = n(),
    total_population = sum(population, na.rm = TRUE),
    total_user_count = sum(user_count, na.rm = TRUE),
    global_coverage_score = sum(user_count, na.rm = TRUE) / sum(population, na.rm = TRUE),
    mean_area_coverage_score = mean(coverage_score, na.rm = TRUE),
    sd_area_coverage_score = sd(coverage_score, na.rm = TRUE)
  )
```

## Review Inputs

`validate_bias_residual_structure()` can compute Moran's I when supplied with a plain neighbour-link table. This notebook creates a deterministic ring-neighbour structure from the simulated area names.

It also creates simple circular coordinates so optional map-ready output can be inspected without spatial dependencies.

```{r}
areas <- sort(unique(simulated_coverage$origin))

area_neighbors <- tibble(
  area = areas,
  neighbor = dplyr::lead(areas, default = areas[1])
) |>
  bind_rows(tibble(
    area = areas,
    neighbor = dplyr::lag(areas, default = areas[length(areas)])
  ))

area_coordinates <- tibble(
  area = areas,
  angle = seq(0, 2 * pi, length.out = length(areas) + 1)[seq_along(areas)],
  x = cos(angle),
  y = sin(angle)
) |>
  select(-angle)

area_neighbors |>
  slice_head(n = 10)
```

## Default Stage 3 Diagnostics

The default selected residual is `coverage_score_residual`.

```{r}
bias_structure <- validate_bias_residual_structure(
  coverage_df = simulated_coverage,
  benchmark_od_df = simulated_benchmark.od,
  area_neighbors = area_neighbors,
  covariate_df = simulated_covariates,
  covariate_col = "internet_access",
  geometry_df = area_coordinates,
  x_col = "x",
  y_col = "y",
  make_plots = has_ggplot2
)

bias_structure$summary
```

```{r}
bias_structure$residual_definitions
```

## Area-Level Residuals

Positive residuals indicate areas with higher active-user coverage than expected under the global coverage score. Negative residuals indicate lower coverage than expected.

```{r}
bias_structure$area_level |>
  select(
    area,
    population,
    user_count,
    coverage_score,
    global_coverage_score,
    expected_user_count,
    user_count_residual,
    coverage_score_residual,
    standardized_user_count_residual,
    selected_residual
  ) |>
  arrange(desc(abs(selected_residual))) |>
  slice_head(n = 15)
```

```{r}
bias_structure$area_level |>
  summarise(
    mean_coverage_score_residual = mean(coverage_score_residual, na.rm = TRUE),
    sd_coverage_score_residual = sd(coverage_score_residual, na.rm = TRUE),
    min_coverage_score_residual = min(coverage_score_residual, na.rm = TRUE),
    max_coverage_score_residual = max(coverage_score_residual, na.rm = TRUE),
    mean_user_count_residual = mean(user_count_residual, na.rm = TRUE),
    sum_user_count_residual = sum(user_count_residual, na.rm = TRUE)
  )
```

## Spatial Randomness

The Moran's I calculation uses only the user-supplied neighbour links.

```{r}
bias_structure$moran_i
```

## Benchmark Flow Correlations

Benchmark flows are collapsed separately to origin totals and destination totals before correlation with the selected bias residual.

```{r}
bias_structure$benchmark_flow_correlation
```

```{r}
bias_structure$benchmark_flow_data |>
  arrange(benchmark_flow_role, desc(abs(selected_residual))) |>
  group_by(benchmark_flow_role) |>
  slice_head(n = 8) |>
  ungroup()
```

## Covariate Correlation

The covariate interface mirrors `validate_flow_residual_structure()`: pass a plain area-level table and name the area and covariate columns explicitly.

```{r}
bias_structure$covariate_correlation
```

```{r}
bias_structure$covariate_data |>
  select(area, covariate_value, selected_residual, coverage_score, population, user_count) |>
  arrange(desc(abs(selected_residual))) |>
  slice_head(n = 12)
```

## Alternative Residual Scale

The same helper can use count-scale or standardized count residuals for the selected diagnostic residual.

```{r}
count_residual_structure <- validate_bias_residual_structure(
  coverage_df = simulated_coverage,
  residual_type = "user_count",
  benchmark_od_df = simulated_benchmark.od,
  covariate_df = simulated_covariates,
  covariate_col = "internet_access"
)

standardized_residual_structure <- validate_bias_residual_structure(
  coverage_df = simulated_coverage,
  residual_type = "standardized_user_count",
  benchmark_od_df = simulated_benchmark.od,
  covariate_df = simulated_covariates,
  covariate_col = "internet_access"
)

bind_rows(
  bias_structure$summary,
  count_residual_structure$summary,
  standardized_residual_structure$summary
) |>
  select(
    residual_type,
    selected_residual_col,
    mean_selected_residual,
    sd_selected_residual,
    pearson_bias_benchmark_origin_flow,
    pearson_bias_benchmark_destination_flow,
    pearson_bias_covariate
  )
```

## Optional Plots

Plots are returned only when `ggplot2` is installed and `make_plots = TRUE`.

```{r}
if (has_ggplot2) {
  bias_structure$plots$bias_residual_distribution
}
```

```{r}
if (has_ggplot2) {
  bias_structure$plots$bias_residual_vs_benchmark_flow
}
```

```{r}
if (has_ggplot2) {
  bias_structure$plots$bias_residual_vs_covariate
}
```

```{r}
if (has_ggplot2) {
  bias_structure$plots$bias_residual_map
}
```

## Review Checklist

```{r}
tibble(
  check = c(
    "coverage residuals sum to zero on the count scale",
    "default residual is coverage_score_residual",
    "benchmark origin and destination correlations returned",
    "covariate correlation returned",
    "map-ready data contains coordinates"
  ),
  passed = c(
    abs(sum(bias_structure$area_level$user_count_residual, na.rm = TRUE)) < 1e-8,
    identical(bias_structure$summary$selected_residual_col, "coverage_score_residual"),
    all(c("origin", "destination") %in% bias_structure$benchmark_flow_correlation$benchmark_flow_role),
    is.finite(bias_structure$covariate_correlation$pearson_r),
    all(c("x", "y") %in% names(bias_structure$map_data))
  )
)
```
