---
title: "Stage 2 Validation Review Notebook"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This notebook runs the Stage 2 validation functions on the package's simulated OD data so the implemented behavior can be reviewed in one place.

It covers:

- `validate_flow_overall()`
- `validate_flow_benchmark()` legacy alias
- `validate_flow_pairs()`
- `validate_flow_all()` legacy alias
- `validate_flow_residuals()`
- `validate_flow_residual_structure()`
- `validate_flow_distribution()`

The notebook uses deterministic simulated data and lightweight deterministic adjustment methods. It does not run the Bayesian prototype.

## Setup

```{r}
find_pkg_root <- function(path = getwd()) {
  path <- normalizePath(path, mustWork = TRUE)

  repeat {
    if (file.exists(file.path(path, "DESCRIPTION"))) {
      return(path)
    }

    parent <- dirname(path)
    if (identical(parent, path)) {
      stop("Could not find package root containing DESCRIPTION.")
    }
    path <- parent
  }
}

pkg_root <- find_pkg_root()
setwd(pkg_root)

if (requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(pkg_root, quiet = TRUE)
} else {
  library(debiasR)
}

library(dplyr)
library(tibble)

has_ggplot2 <- requireNamespace("ggplot2", quietly = TRUE)
if (has_ggplot2) {
  library(ggplot2)
}

comparison_palette <- c(
  "Adjusted vs benchmark" = "#2B8CBE",
  "Raw vs benchmark" = "#756BB1",
  "Raw vs adjusted" = "#636363"
)
reference_line_colour <- "#D95F0E"

flow_axis_label <- function(x) {
  dplyr::case_when(
    abs(x) >= 1e6 ~ paste0(round(x / 1e6, 1), "m"),
    abs(x) >= 1e3 ~ paste0(round(x / 1e3, 1), "k"),
    TRUE ~ format(round(x, 0), big.mark = ",", scientific = FALSE, trim = TRUE)
  )
}

theme_validation_review <- function() {
  ggplot2::theme_minimal(base_size = 11) +
    ggplot2::theme(
      legend.position = "bottom",
      panel.grid.minor = ggplot2::element_blank(),
      strip.text = ggplot2::element_text(face = "bold")
    )
}
```

```{r}
data("simulated_mpd.od", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_coverage", package = "debiasR")
data("simulated_covariates", package = "debiasR")

tibble(
  dataset = c(
    "simulated_mpd.od",
    "simulated_benchmark.od",
    "simulated_coverage",
    "simulated_covariates"
  ),
  rows = c(
    nrow(simulated_mpd.od),
    nrow(simulated_benchmark.od),
    nrow(simulated_coverage),
    nrow(simulated_covariates)
  ),
  columns = c(
    paste(names(simulated_mpd.od), collapse = ", "),
    paste(names(simulated_benchmark.od), collapse = ", "),
    paste(names(simulated_coverage), collapse = ", "),
    paste(names(simulated_covariates), collapse = ", ")
  )
)
```

## Simulated Dataset Overview

The review uses four deterministic simulated datasets shipped with `debiasR`:

- `simulated_mpd.od`: observed mobile-phone-derived OD flows. This is the biased flow table that adjustment methods start from.
- `simulated_benchmark.od`: benchmark OD flows used as the validation target.
- `simulated_coverage`: area-level population and active-user counts used by coverage-based adjustment methods.
- `simulated_covariates`: area-level contextual variables used by model-based adjustment and residual-structure diagnostics.

The flow datasets share the same `origin` and `destination` keys. The validation functions compare the observed MPD flow and each adjusted `flow_adj` against the benchmark `flow`.

```{r}
flow_overview <- bind_rows(
  simulated_mpd.od |>
    transmute(series = "Observed MPD", flow = flow),
  simulated_benchmark.od |>
    transmute(series = "Benchmark", flow = flow)
)

flow_overview |>
  group_by(series) |>
  summarise(
    n = n(),
    total_flow = sum(flow, na.rm = TRUE),
    zero_share = mean(flow == 0, na.rm = TRUE),
    median_flow = median(flow, na.rm = TRUE),
    mean_flow = mean(flow, na.rm = TRUE),
    p90_flow = unname(quantile(flow, 0.90, na.rm = TRUE)),
    p99_flow = unname(quantile(flow, 0.99, na.rm = TRUE)),
    .groups = "drop"
  )
```

The figure below compares the simulated raw MPD and benchmark flow
distributions on a log scale.

```{r}
if (has_ggplot2) {
  ggplot2::ggplot(
    flow_overview,
    ggplot2::aes(x = log1p(flow), fill = series)
  ) +
    ggplot2::geom_histogram(
      bins = 40,
      alpha = 0.65,
      position = "identity",
      color = "white"
    ) +
    ggplot2::facet_wrap(ggplot2::vars(series), ncol = 1) +
    ggplot2::scale_fill_manual(
      values = c("Observed MPD" = "#756BB1", "Benchmark" = "#2B8CBE")
    ) +
    ggplot2::labs(
      x = "log1p(flow)",
      y = "OD pairs",
      fill = NULL
    ) +
    theme_validation_review()
} else {
  hist(
    log1p(simulated_mpd.od$flow),
    breaks = 40,
    main = "Observed MPD OD flow distribution",
    xlab = "log1p(flow)"
  )
}
```

## Build Review Inputs

Create a small set of adjusted-flow outputs. These are not intended to pick a winner; they provide method-labelled inputs for reviewing validation behavior.

```{r}
adj_inverse <- adjust_inverse_penetration(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  weight_by = "both"
)

adj_raking <- adjust_raking_ratio(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od
)

adj_coefficient <- adjust_coefficient(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od,
  model_family = "ols",
  level = "od"
)

adjusted_outputs <- list(
  adjust_inverse_penetration = adj_inverse,
  adjust_raking_ratio = adj_raking,
  adjust_coefficient = adj_coefficient
)
```

```{r}
bind_rows(lapply(names(adjusted_outputs), function(method_name) {
  adjusted_outputs[[method_name]] |>
    summarise(
      method = method_name,
      n = n(),
      total_mpd = sum(flow, na.rm = TRUE),
      total_adjusted = sum(flow_adj, na.rm = TRUE),
      min_adjusted = min(flow_adj, na.rm = TRUE),
      max_adjusted = max(flow_adj, na.rm = TRUE)
    )
}))
```

The figure below uses `adjust_raking_ratio` as a compact visual check of the
three validation comparisons: adjusted against benchmark, raw MPD against
benchmark, and adjusted against raw MPD.

```{r}
build_review_scatter <- function(adj_df, method_name) {
  joined <- adj_df |>
    select(origin, destination, flow_mpd = flow, flow_adj) |>
    inner_join(
      simulated_benchmark.od |>
        select(origin, destination, flow_bench = flow),
      by = c("origin", "destination")
    )

  bind_rows(
    joined |>
      transmute(
        method = method_name,
        comparison = "Adjusted vs benchmark",
        reference_flow = flow_bench,
        compared_flow = flow_adj
      ),
    joined |>
      transmute(
        method = method_name,
        comparison = "Raw vs benchmark",
        reference_flow = flow_bench,
        compared_flow = flow_mpd
      ),
    joined |>
      transmute(
        method = method_name,
        comparison = "Raw vs adjusted",
        reference_flow = flow_mpd,
        compared_flow = flow_adj
      )
  ) |>
    mutate(
      comparison = factor(
        comparison,
        levels = c(
          "Adjusted vs benchmark",
          "Raw vs benchmark",
          "Raw vs adjusted"
        )
      )
    )
}

if (has_ggplot2) {
  build_review_scatter(adj_raking, "adjust_raking_ratio") |>
    ggplot(aes(x = reference_flow, y = compared_flow, colour = comparison)) +
    geom_point(alpha = 0.55, size = 1.6, na.rm = TRUE) +
    geom_abline(
      intercept = 0,
      slope = 1,
      linetype = "dashed",
      colour = reference_line_colour
    ) +
    facet_wrap(~comparison, scales = "free") +
    scale_colour_manual(values = comparison_palette) +
    scale_x_continuous(labels = flow_axis_label) +
    scale_y_continuous(labels = flow_axis_label) +
    guides(colour = "none") +
    labs(
      x = "Reference flow",
      y = "Compared flow"
    ) +
    theme_validation_review()
} else {
  "Install ggplot2 to review the three-way validation scatterplot."
}
```

## Helper Inputs For Residual Structure

`validate_flow_residual_structure()` can compute Moran's I if supplied a plain neighbour-link table. This review notebook creates a deterministic ring-neighbour structure from the simulated area names.

It also creates simple circular coordinates so the optional residual map plot can be inspected without adding spatial dependencies.

```{r}
areas <- sort(unique(c(simulated_mpd.od$origin, simulated_mpd.od$destination)))

area_neighbors <- tibble(
  area = areas,
  neighbor = dplyr::lead(areas, default = areas[1])
) |>
  bind_rows(tibble(
    area = areas,
    neighbor = dplyr::lag(areas, default = areas[length(areas)])
  ))

area_coordinates <- tibble(
  area = areas,
  angle = seq(0, 2 * pi, length.out = length(areas) + 1)[seq_along(areas)],
  x = cos(angle),
  y = sin(angle)
) |>
  select(-angle)

area_neighbors |> slice_head(n = 10)
```

## 1. `validate_flow_overall()`

Review method-level fit summaries.

```{r}
overall_results <- bind_rows(lapply(names(adjusted_outputs), function(method_name) {
  res <- validate_flow_overall(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    return_joined = FALSE
  )

  tibble(
    method = res$method,
    n = res$n,
    pearson_r = res$pearson_r,
    spearman_rho = res$spearman_rho,
    rmse = res$rmse,
    mae = res$mae,
    mape = res$mape,
    ols_intercept = res$ols_intercept,
    ols_slope = res$ols_slope,
    r_squared = res$r_squared
  )
}))

overall_results |>
  arrange(desc(pearson_r))
```

## 2. `validate_flow_benchmark()` Legacy Alias

Confirm the legacy alias still delegates to `validate_flow_overall()`.

```{r}
alias_overall <- validate_flow_benchmark(
  adj_df = adj_inverse,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "alias_check",
  return_joined = FALSE
)

tibble(
  alias_method = alias_overall$method,
  alias_n = alias_overall$n,
  alias_pearson_r = alias_overall$pearson_r
)
```

## 3. `validate_flow_pairs()`

Review the joined OD-level comparison table.

```{r}
pairs_inverse <- validate_flow_pairs(
  adj_df = adj_inverse,
  benchmark_od_df = simulated_benchmark.od
)

pairs_inverse |>
  select(
    origin,
    destination,
    mpd_flow,
    adj_flow,
    benchmark_flow,
    diff_mpd_benchmark,
    diff_mpd_adj,
    diff_adj_benchmark
  ) |>
  slice_head(n = 12)
```

## 4. `validate_flow_all()` Legacy Alias

Confirm the legacy row-level alias still delegates to `validate_flow_pairs()`.

```{r}
alias_pairs <- validate_flow_all(
  adj_df = adj_inverse,
  benchmark_od_df = simulated_benchmark.od
)

tibble(
  identical_alias_output = identical(alias_pairs, pairs_inverse),
  alias_rows = nrow(alias_pairs),
  alias_columns = ncol(alias_pairs)
)
```

## 5. `validate_flow_residuals()`

Review OD-level residual reduction, raw-versus-adjusted outlier shares, and top-worst residual rows.

```{r}
residual_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_residuals(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    top_n = 8
  )
})
names(residual_results) <- names(adjusted_outputs)

residual_summary <- bind_rows(lapply(residual_results, `[[`, "summary"))

residual_summary |>
  select(
    method,
    n,
    mean_signed_residual_reduction,
    median_signed_residual_reduction,
    mean_abs_residual_reduction,
    median_abs_residual_reduction,
    share_improved,
    share_worsened,
    share_moved_in_benchmark_direction,
    share_residual_mpd_over_2sd,
    share_residual_adj_over_2sd,
    reduction_share_residual_over_2sd
  ) |>
  arrange(desc(share_improved))
```

```{r}
residual_results$adjust_inverse_penetration$data |>
  select(
    method,
    origin,
    destination,
    mpd_flow,
    adj_flow,
    benchmark_flow,
    signed_residual_reduction,
    abs_residual_reduction,
    moved_in_benchmark_direction,
    improvement_flag
  ) |>
  slice_head(n = 12)
```

```{r}
residual_results$adjust_inverse_penetration$top_worst |>
  select(
    method,
    origin,
    destination,
    mpd_flow,
    adj_flow,
    benchmark_flow,
    abs_residual_adj_benchmark,
    abs_residual_reduction,
    improvement_flag
  )
```

## 6. `validate_flow_residual_structure()`

Review residual structure for one method in detail.

```{r}
structure_inverse <- validate_flow_residual_structure(
  adj_df = adj_inverse,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_inverse_penetration",
  residual_type = "adjusted",
  spatial_role = "origin",
  residual_aggregation = "mean",
  area_neighbors = area_neighbors,
  covariate_df = simulated_covariates,
  covariate_col = "income_norm",
  geometry_df = area_coordinates,
  x_col = "x",
  y_col = "y",
  make_plots = has_ggplot2
)

structure_inverse$summary
```

```{r}
structure_inverse$flow_correlation
```

```{r}
structure_inverse$moran_i
```

```{r}
structure_inverse$covariate_correlation
```

```{r}
structure_inverse$area_level |>
  arrange(desc(abs(selected_residual))) |>
  slice_head(n = 12)
```

```{r}
structure_inverse$map_data |>
  arrange(desc(abs(selected_residual))) |>
  select(area, selected_residual, mean_abs_residual, benchmark_flow_sum, x, y) |>
  slice_head(n = 12)
```

### Optional Residual Structure Plots

These chunks show the plots returned by `validate_flow_residual_structure()` when `ggplot2` is installed.

```{r}
if (has_ggplot2) {
  structure_inverse$plots$residual_reduction_distribution
} else {
  "Install ggplot2 to review the residual-reduction distribution plot."
}
```

```{r}
if (has_ggplot2) {
  structure_inverse$plots$residual_vs_benchmark
} else {
  "Install ggplot2 to review the residual-versus-benchmark plot."
}
```

```{r}
if (has_ggplot2) {
  structure_inverse$plots$residual_vs_covariate
} else {
  "Install ggplot2 to review the residual-versus-covariate plot."
}
```

```{r}
if (has_ggplot2) {
  structure_inverse$plots$residual_map
} else {
  "Install ggplot2 to review the coordinate residual map."
}
```

## 7. `validate_flow_distribution()`

Review origin-conditioned destination-share fidelity with KL divergence and Jensen-Shannon divergence.

```{r}
distribution_results <- lapply(names(adjusted_outputs), function(method_name) {
  validate_flow_distribution(
    adj_df = adjusted_outputs[[method_name]],
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    weight_by = "benchmark_origin_total"
  )
})
names(distribution_results) <- names(adjusted_outputs)

distribution_summary <- bind_rows(lapply(distribution_results, `[[`, "summary"))

distribution_summary |>
  select(
    method,
    n_origins,
    n_origins_used,
    weight_by,
    kl_mean,
    kl_median,
    kl_weighted_mean,
    jsd_mean,
    jsd_median,
    jsd_weighted_mean
  ) |>
  arrange(jsd_weighted_mean)
```

```{r}
distribution_results$adjust_inverse_penetration$origin_level |>
  arrange(desc(jsd_origin)) |>
  slice_head(n = 12)
```

## Combined Review Table

This table joins the main method-level Stage 2 summaries.

```{r}
combined_review <- overall_results |>
  select(method, pearson_r, rmse, mae, mape, r_squared) |>
  left_join(
    residual_summary |>
      select(
        method,
        mean_abs_residual_reduction,
        share_improved,
        reduction_share_residual_over_2sd
      ),
    by = "method"
  ) |>
  left_join(
    distribution_summary |>
      select(method, kl_weighted_mean, jsd_weighted_mean),
    by = "method"
  ) |>
  arrange(desc(pearson_r))

combined_review
```

## Review Checklist

- Does `signed_residual_reduction` read clearly as directional movement rather than guaranteed improvement?
- Is `abs_residual_reduction` the right method-comparison field for "positive means better" residual reduction?
- Are MPD and adjusted outlier shares both useful enough to keep in `validate_flow_residuals()`?
- Is the plain neighbour-link interface for Moran's I transparent enough?
- Should optional plots remain inside `validate_flow_residual_structure()`?
- Are KL divergence and Jensen-Shannon divergence summaries clear enough for method comparison?
