---
title: "debiasR: Conceptual Guide to Adjustment Methods"
format:
  html:
    toc: true
    html-math-method: mathml
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This short notebook explains, intuitively, how each `debiasR` adjustment method works and how it is implemented in the package.

## Setup

```{r}
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

data(simulated_mpd.od, package = "debiasR")
data(simulated_benchmark.od, package = "debiasR")
data(simulated_coverage, package = "debiasR")
data(simulated_covariates, package = "debiasR")
data(simulated_distance, package = "debiasR")
```

<style>
.method-table-wrap {
  overflow-x: auto;
  margin: 1.25rem 0 1.75rem 0;
  padding: 0.6rem 0.8rem;
  border: 1px solid #e5e7eb;
  border-radius: 14px;
  background: #fff;
}

.method-summary-table {
  min-width: 1900px;
  font-size: 0.92rem;
  line-height: 1.45;
}

.method-summary-table th,
.method-summary-table td {
  vertical-align: top;
  padding: 0.8rem 0.75rem;
  white-space: normal;
}

.method-summary-table th {
  font-size: 0.95rem;
}

.method-summary-table td:nth-child(1) {
  min-width: 17rem;
}

.method-summary-table td:nth-child(2) {
  min-width: 17rem;
}

.method-summary-table td:nth-child(3) {
  min-width: 19rem;
}

.eq-stack {
  font-family: "Cambria Math", "STIX Two Text", "Times New Roman", serif;
  line-height: 1.45;
}

.eq-line {
  display: block;
}

.eq-line + .eq-line {
  margin-top: 0.28rem;
}
</style>

## Shared Intuition

All methods start from observed MPD flows (`flow`) and produce adjusted flows (`flow_adj`).

- If a method thinks MPD under-observes a flow, it increases `flow_adj`.
- If it thinks MPD over-observes a flow, it decreases `flow_adj`.

The methods differ in *how* they infer that correction.

### Notation Used

- $F^{mpd}_{ij}$: observed MPD flow from origin $i$ to destination $j$.
- $F^{bench}_{ij}$: benchmark (census-like) flow from $i$ to $j$.
- $F^{adj}_{ij}$: adjusted flow output by a method.
- $P_i, P_j$: population at origin/destination.
- $U_i, U_j$: active user count at origin/destination.
- $p_i = U_i/P_i$: coverage (penetration) rate.
- $d_{ij}$: distance between $i$ and $j$.
- $X_i, X_j$: origin/destination covariates (for example, income).

## Methods In Scope

This vignette covers the six adjustment methods currently implemented in `debiasR`:

- `adjust_inverse_penetration()`
- `adjust_selection_rate()`
- `adjust_selection_rate2()`
- `adjust_raking_ratio()`
- `adjust_coefficient()`
- `adjust_multilevel_bayes()`

Methods discussed in earlier research notes but not currently implemented in the package are intentionally excluded here:

- calibration / transfer frameworks
- difference-in-differences approaches
- Bayesian data fusion across multiple biased sources

## Package Workflow

![Workflow linking clearly identified input data to bias measurement, bias adjustment, and validation.](figures/package-workflow.svg){fig-alt="debiasR workflow showing an input-data panel with coverage inputs, flow inputs, and enrichment inputs feeding three stacked steps: measure bias, adjust bias, and validate bias-adjusted flows."}

## Adjustment Method Landscape

![Available adjustment methods in debiasR, grouped by coverage-based reweighting, benchmark-constrained calibration, and hierarchical Bayesian modelling.](figures/adjustment-methods-overview.svg){fig-alt="Available adjustment methods in debiasR grouped into coverage-based reweighting, benchmark-constrained calibration, and hierarchical Bayesian modelling, with each method card listing its main data requirements."}

## Validation functions

The package exposes five validation functions because they solve different problems:

- `validate_flow_overall()` is the summary scorer. It uses adjusted flows from `adj_df` and benchmark flows from `benchmark_od_df`, then returns method-level fit metrics such as correlation, RMSE, MAE, MAPE, and regression diagnostics.
- `validate_flow_pairs()` is the lower-level audit table. It keeps the core joined columns and raw residual differences, and is useful when you only want the minimal OD-level join without the additional diagnostics.
- `validate_flow_residuals()` is the richer OD-level residual diagnostic. It builds on the same inputs, then adds signed movement, absolute residual reduction, percentage residuals, improvement flags, raw-versus-adjusted standard-deviation threshold flags, and a `top_worst` table of OD pairs with the largest residuals remaining after adjustment.
- `validate_flow_residual_structure()` diagnoses residual structure. It returns residual-versus-benchmark-flow correlation, area-level residual summaries, optional Moran's I from user-supplied neighbour links, optional covariate correlation, map-ready residual data, and optional diagnostic plots.
- `validate_flow_distribution()` checks whether adjusted flows preserve the benchmark destination-share distribution by origin, using `KL(benchmark || adjusted)` and Jensen-Shannon divergence.

For compatibility, the earlier names `validate_flow_benchmark()` and `validate_flow_all()` are still available as aliases for one release cycle, but new code should prefer the current names.

Together they support two complementary checks:

- summary fit or proportionality: are adjusted flows aligned with the benchmark pattern overall?
- residual audit: for each OD pair, by how much do MPD and adjusted flows differ from the benchmark?
- residual structure: are remaining errors spatially patterned or associated with flow size or covariates?
- allocation fidelity: are destinations receiving similar shares of each origin's benchmark flow?

## Implemented Method Summary Table

```{r results='asis', echo=FALSE}
eq_stack <- function(...) {
  paste0(
    "<div class='eq-stack'>",
    paste0("<div class='eq-line'>", c(...), "</div>", collapse = ""),
    "</div>"
  )
}

method_summary <- tibble::tribble(
  ~Method, ~Rationale, ~Equation, ~`Benchmark<br>population`, ~`Active<br>users`, ~`Benchmark<br>OD flows`, ~`Digital<br>OD flows`, ~Covariates, ~Assumptions, ~Strengths, ~Weaknesses, ~`Suitability<br>for OD flows`, ~`Validation<br>strategy`,
  "<code>adjust_inverse_penetration()</code>",
  "Inflate observed flows using population-to-user ratios where coverage is thin.<br/>Good transparent option when bias is mainly under-coverage.",
  eq_stack(
    "<i>w</i><sub>i</sub> = P<sub>i</sub> / U<sub>i</sub>",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>i</sub>, <i>w</i><sub>j</sub>, or &radic;(<i>w</i><sub>i</sub><i>w</i><sub>j</sub>)"
  ),
  "Yes", "Yes", "No", "Yes", "No",
  "Coverage bias is roughly proportional to penetration; origin/destination weights capture most bias.",
  "Simple, transparent, cheap, no benchmark OD needed.",
  "Can overinflate sparse areas; misses compositional and structural selection bias.",
  "Good option for observed OD correction; weaker for nuanced structural bias.",
  "<code>validate_flow_overall()</code> for fit metrics; <code>validate_flow_residuals()</code> to inspect large OD corrections.",

  "<code>adjust_selection_rate()</code>",
  "Blend penetration with an income-like covariate to model differential selection, with optional benchmark calibration.<br/>Useful when simple inverse weighting is too crude.",
  eq_stack(
    "<i>w</i><sub>i</sub>(<i>r</i><sub>t</sub>) = 1 / (I<sub>i</sub><i>p</i><sub>i</sub> + (1 - I<sub>i</sub>)<i>r</i><sub>t</sub>)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>"
  ),
  "Yes", "Yes", "Optional", "Yes", "Yes",
  "Income-like covariate is informative; global <code>r_t</code> is meaningful; calibration benchmark is representative if used.",
  "More flexible than inverse penetration; can incorporate covariates and benchmark calibration.",
  "More tuning and interpretation burden; depends on covariate quality and calibration choice.",
  "Good for OD flows where selection varies systematically across places.",
  "Compare calibrated and uncalibrated outputs with <code>validate_flow_overall()</code>; use <code>validate_flow_residuals()</code> for OD residuals.",

  "<code>adjust_selection_rate2()</code>",
  "Apply a closed-form penetration correction with optional <code>k</code> calibration.<br/>Useful as a parsimonious selection-style alternative.",
  eq_stack(
    "CF(<i>p</i>; <i>k</i>) = <i>p</i> &times; (exp(-<i>k</i>) - 1) / (exp(-<i>k</i><i>p</i>) - 1)",
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; CF(...)"
  ),
  "Yes", "Yes", "Optional", "Yes", "No",
  "A single curvature parameter <code>k</code> can summarize selection bias; penetration is measured well.",
  "Compact, interpretable, lighter than method 2, can be calibrated.",
  "Less flexible than method 2; may underfit heterogeneous bias.",
  "Good lightweight OD correction when coverage is the dominant concern.",
  "Use <code>validate_flow_overall()</code> for summary performance and <code>validate_flow_residuals()</code> for OD-level residual patterns.",

  "<code>adjust_raking_ratio()</code>",
  "Force adjusted OD flows to match trusted origin and/or destination margins.<br/>Best when consistency with benchmark totals is the main priority.",
  eq_stack(
    "<i>F</i><sup>adj</sup><sub>ij</sub> = <i>F</i><sup>mpd</sup><sub>ij</sub> &times; <i>w</i><sub>ij</sub>",
    "&Sigma;<sub>j</sub><i>F</i><sup>adj</sup><sub>ij</sub> = O<sub>i</sub> and/or &Sigma;<sub>i</sub><i>F</i><sup>adj</sup><sub>ij</sub> = D<sub>j</sub>"
  ),
  "Optional (as targets)", "No", "Yes or user targets", "Yes", "No",
  "Benchmark marginals are correct; observed OD support is adequate; zero cells need not be created.",
  "Ensures consistency with trusted marginals; clear and widely understood.",
  "Cannot create missing OD support; may improve totals more than internal OD realism.",
  "Very good for OD tables when marginal consistency matters most.",
  "Check convergence and margin matching, then use <code>validate_flow_overall()</code> and <code>validate_flow_residuals()</code>.",

  "<code>adjust_coefficient()</code>",
  "Learn a global mapping from MPD flows to benchmark OD flows, then apply it back to MPD.<br/>Best when benchmark OD is available and calibration is the main goal.",
  eq_stack(
    "E[<i>F</i><sup>bench</sup><sub>ij</sub>] = &beta; <i>F</i><sup>mpd</sup><sub>ij</sub>",
    "log(&mu;<sub>ij</sub>) = &alpha; + log(<i>F</i><sup>mpd</sup><sub>ij</sub>)"
  ),
  "No", "No", "Yes", "Yes", "No",
  "Relationship between MPD and benchmark flows is stable enough to estimate; overlapping positive support exists.",
  "Direct benchmark-driven calibration; multiple model families available.",
  "Benchmark quality is critical; a global coefficient may miss local heterogeneity.",
  "Strong when benchmark OD exists and a simple calibrated mapping is acceptable.",
  "Use <code>validate_flow_overall()</code> across families and levels; use <code>validate_flow_residuals()</code> for systematic residuals.",

  "<code>adjust_multilevel_bayes()</code>",
  "Fit a multilevel count model with coverage bias, covariates, and optional distance, then remove the estimated coverage-bias contribution from model fixed-effect predictors.<br/>Current S1-S4 development uses the frequentist engine as a fast placeholder while Bayesian transfer is completed.",
  eq_stack(
    "log(&mu;<sup>obs</sup><sub>ij</sub>) = &alpha; + &beta;<sub>o</sub>X<sub>i</sub> + &beta;<sub>d</sub>X<sub>j</sub> + &gamma; log(<i>d</i><sub>ij</sub>) + &omega;e<sub>i</sub> + <i>u</i>",
    "flow<sub>adj</sub> = summary(exp(&eta;<sup>fix</sup><sub>ij</sub> - &omega;e<sub>i</sub>))"
  ),
  "Yes", "Yes", "No", "Yes", "Yes",
  "Model family and hierarchy are well specified; coverage bias is captured by the fitted bias term; S1-S4 source/time structure is declared explicitly.",
  "Richer structure, no benchmark OD needed for fitting; frequentist engine gives a fast contract check before Bayesian sampling.",
  "Bayesian S2-S4 implementation is deferred; frequentist output is a development placeholder, not final uncertainty-aware inference.",
  "Promising for noisy or sparse observed OD correction; not yet ready for full OD completion.",
  "Posterior predictive checks plus <code>validate_flow_overall()</code> on observed OD rows; use <code>validate_flow_residuals()</code> for row-level audit."
)

table_html <- knitr::kable(
  method_summary,
  format = "html",
  escape = FALSE,
  table.attr = "class='table method-summary-table'"
)

cat("<div class='method-table-wrap'>", table_html, "</div>")
```

Key contrast:
- `adjust_coefficient()` directly learns from benchmark flows.
- `adjust_multilevel_bayes()` does **not** use benchmark flows in model fitting; it adjusts MPD using coverage/covariates/distance/hierarchical structure. It keeps observed OD behavior as the default and now also supports a complete-grid prediction scope for square OD inputs.
- During S1-S4 scenario development, use `model_engine = "frequentist"` as the fast placeholder. The final inferential target remains the Bayesian engine.

## Method 1: `adjust_inverse_penetration()`

### Concept

If only a fraction of the population is observed in MPD, inflate flows by the inverse of that fraction.

- Coverage at area `i`: `user_count_i / population_i`
- Inverse penetration weight: `population_i / user_count_i`

### Equation

$$
p_i = \frac{U_i}{P_i}, \quad w_i = \frac{1}{p_i} = \frac{P_i}{U_i}
$$

Origin-weighted version:
$$
F^{adj}_{ij} = w_i F^{mpd}_{ij}
$$

Destination-weighted version:
$$
F^{adj}_{ij} = w_j F^{mpd}_{ij}
$$

Both:
$$
F^{adj}_{ij} = \sqrt{w_i w_j} \, F^{mpd}_{ij}
$$

### Implementation in `debiasR`

- Function: `adjust_inverse_penetration()`
- Uses `coverage_df` columns `origin`, `population`, `user_count` (and optionally `mpd_source`).
- Applies weights by origin, destination, or both (`weight_by`).

```{r}
adj_m1 <- adjust_inverse_penetration(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  weight_by = "both"
)

adj_m1 |> select(origin, destination, flow, flow_adj) |> slice_head(n = 5)
```

## Method 2: `adjust_selection_rate()`

### Concept

Adjust with a selection-rate style factor that combines:

- coverage terms (population/users),
- covariates (for example, income),
- optional calibration to benchmark totals.

### Equation

Origin-side weight:
$$
w_i^{(O)}(r_t) = \frac{1}{I_i p_i^{(O)} + (1 - I_i) r_t}
$$

Destination-side weight:
$$
w_j^{(D)}(r_t) = \frac{1}{I_j p_j^{(D)} + (1 - I_j) r_t}
$$

Adjustment:
$$
F^{adj}_{ij} = F^{mpd}_{ij} \times w_{ij}
$$

where $w_{ij}$ is origin-weighted, destination-weighted, or the geometric mean of the two.

### Implementation in `debiasR`

- Function: `adjust_selection_rate()`
- Requires `covariates_df` and coverage inputs.
- Can calibrate globally/origin/destination using `benchmark_od_df`.

```{r}
adj_m2 <- adjust_selection_rate(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  covariates_df = simulated_covariates,
  income_col = "income_norm",
  weight_by = "origin",
  benchmark_od_df = simulated_benchmark.od,
  calibration_aggregate = "origin"
)

attr(adj_m2, "r_global")
```

## Method 3: `adjust_selection_rate2()`

### Concept

A simplified/alternative selection-rate adjustment that often behaves like a multiplicative scaling of MPD flows.

### Equation

$$
CF(p; k) = p \frac{e^{-k} - 1}{e^{-k p} - 1}
$$

with:
$$
F^{adj}_{ij} = F^{mpd}_{ij} \times CF(p; k)
$$

or the corresponding destination-side / both-side version depending on `weight_by`.

### Implementation in `debiasR`

- Function: `adjust_selection_rate2()`
- Uses coverage-derived correction and optional calibration constant `k`.
- Often yields near-linear `flow_adj` vs `flow`.

```{r}
adj_m3 <- adjust_selection_rate2(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  weight_by = "origin",
  benchmark_od_df = simulated_benchmark.od,
  calibration_aggregate = "origin"
)

attr(adj_m3, "k")
```

## Method 4: `adjust_raking_ratio()`

### Concept

Use iterative proportional fitting (IPF/raking) so adjusted OD flows match known marginals (from benchmark).

### Equation (constraint form)

Find $F^{adj}_{ij}$ close to $F^{mpd}_{ij}$ such that:
$$
\sum_j F^{adj}_{ij} = O_i^{bench}, \quad
\sum_i F^{adj}_{ij} = D_j^{bench}
$$

where $O_i^{bench}$ and $D_j^{bench}$ are benchmark outflow/inflow marginals.
IPF alternates row and column scaling until constraints are met.

### Implementation in `debiasR`

- Function: `adjust_raking_ratio()`
- Reweights the OD matrix iteratively to satisfy margin constraints.
- Good when matching aggregate totals is the priority.

```{r}
adj_m4 <- adjust_raking_ratio(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od
)

list(
  ipf_converged = attr(adj_m4, "ipf_converged"),
  ipf_iterations = attr(adj_m4, "ipf_iterations")
)
```

## Method 5: `adjust_coefficient()`

### Concept

Learn a mapping from MPD flows to benchmark flows via a regression coefficient model, then apply that mapping.

### Equation

Core proportional form:
$$
E[F^{bench}_{ij}] = \beta F^{mpd}_{ij}
$$

For count families in the package:
$$
\log(\mu_{ij}) = \alpha + \log(F^{mpd}_{ij})
$$

so that:
$$
\mu_{ij} = e^{\alpha} F^{mpd}_{ij}, \quad \beta = e^{\alpha}
$$

### Implementation in `debiasR`

- Function: `adjust_coefficient()`
- Supports model families (for example, OLS).
- Can fit at OD level and project to adjusted flows.

```{r}
adj_m5 <- adjust_coefficient(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od,
  model_family = "ols",
  level = "od"
)

attr(adj_m5, "coef")
```

## Method 6: `adjust_multilevel_bayes()`

### Concept

Fit a Bayesian multilevel count model to observed MPD flows:

- fixed effects for origin/destination predictors and distance,
- explicit bias term from coverage (`bias_e_origin`),
- random intercept structure (for example, origin).

Then remove the estimated coverage-bias contribution from the model fixed-effect linear predictor to obtain `flow_adj`.

### Equation (prototype in package)

Observation model:
$$
F^{mpd}_{ij} \sim \text{Poisson}(\mu^{obs}_{ij})
$$

Observed-flow model:
$$
\log \mu^{obs}_{ij}
= \alpha + \beta_o X_i + \beta_d X_j + \gamma \log d_{ij} + \omega e_i + u_i
$$

with coverage-bias term:
$$
e_i = 1 - \frac{U_i}{P_i}
$$

Stage-1 adjusted fixed-effect predictor:
$$
\eta^{adj}_{ij} = \eta^{fix}_{ij} - \omega e_i
$$

and:
$$
F^{adj}_{ij} = \exp(\eta^{adj}_{ij})
$$

In the Bayesian implementation, this is done draw-by-draw from the posterior and summarized by mean/median. While the S1-S4 Bayesian transfer is being completed, the same function can be run with `model_engine = "frequentist"` to test the data contract, formula, and output shape quickly.

### Implementation in `debiasR`

- Function: `adjust_multilevel_bayes()`
- Development engine: `model_engine = "frequentist"` for fast Poisson/negative-binomial GLM or GLMM checks.
- Bayesian engine: `model_engine = "bayesian"` for the existing Stage-1 S1 path; S2-S4 Bayesian fitting is deferred until the frequentist contract is stable.
- Returns explicit scenario, formula-term, model, and diagnostic metadata in attributes.

The example below constructs the simplest S1 data structure: one MPD data source
and one observation period. We include constant `mpd_source` and `mpd_time`
columns even though S1 can also be inferred without them, because naming those
columns makes the transition to S2-S4 explicit.

```{r}
mpd_s1 <- simulated_mpd.od |>
  mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

coverage_s1 <- simulated_coverage |>
  mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

adj_m6 <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = simulated_covariates,
  distance_df = simulated_distance,
  formula = flow ~ income_norm_o + income_norm_d + log_distance + bias_e_origin,
  model_engine = "frequentist",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median"
)

adj_m6 |>
  select(origin, destination, mpd_source, mpd_time, flow, flow_adj) |>
  slice_head(n = 5)
```

Inspect the metadata to confirm this was treated as S1 and to see the resolved
formula terms used by the placeholder engine.

```{r}
attr(adj_m6, "result_metadata")[c(
  "model_engine",
  "scenario",
  "repeated_observation",
  "n_sources",
  "n_time_periods"
)]

attr(adj_m6, "model_terms")
```

To fit alternative data structures, keep the same function and change the
scenario columns and repeated-observation contract:

```{r}
tibble::tribble(
  ~Data_structure, ~scenario, ~source_col, ~time_col, ~repeated_observation, ~typical_random_intercept,
  "S1: one source, one time", "s1", "optional constant source column", "optional constant time column", "none", "none or origin",
  "S2: one source, multiple times", "s2", "optional constant source column", "time-period column", "time", "time or origin",
  "S3: multiple sources, one time", "s3", "source column", "optional constant time column", "source", "source or origin",
  "S4: multiple sources, multiple times", "s4", "source column", "time-period column", "source_time", "source_time, source, time, or origin"
)
```

## How to Read Outputs

For all methods:

- Primary output: tibble with `origin`, `destination`, `flow`, `flow_adj`.
- Compare methods with:
  - `validate_flow_overall()` for fit/proportionality metrics
  - `validate_flow_residuals()` for richer OD-level residual inspection
  - `validate_flow_pairs()` for the lower-level joined audit table
  - scatterplots (`flow_adj` vs benchmark, `flow_adj` vs MPD).

```{r}
val_m1 <- validate_flow_overall(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_inverse_penetration",
  return_joined = FALSE
)

tibble::tibble(
  method = val_m1$method,
  n = val_m1$n,
  pearson_r = val_m1$pearson_r,
  rmse = val_m1$rmse,
  mae = val_m1$mae,
  r_squared = val_m1$r_squared
)
```

```{r}
audit_m1 <- validate_flow_pairs(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od
)

audit_m1 |>
  select(origin, destination, mpd_flow, adj_flow, benchmark_flow,
         diff_mpd_benchmark, diff_adj_benchmark) |>
  slice_head(n = 5)
```

```{r}
resid_m1 <- validate_flow_residuals(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od,
  top_n = 5
)

resid_m1$summary
```

```{r}
resid_m1$top_worst |>
  select(origin, destination, benchmark_flow, adj_flow,
         residual_adj_benchmark, abs_residual_adj_benchmark,
         residual_adj_over_1sd, residual_adj_over_2sd, residual_adj_over_3sd,
         improvement_flag) |>
  slice_head(n = 5)
```

## Practical Takeaway

- Need fast transparent scaling: start with Method 1 or 3.
- Need margin consistency: Method 4.
- Need model-based mapping: Method 5.
- Need exploratory uncertainty + hierarchical bias structure with observed or complete-grid OD prediction: Method 6.
