---
title: "debiasR: Method Comparison Walkthrough"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This vignette compares the implemented adjustment methods in `debiasR` and
includes the multilevel helper as the main experimental prototype:

1. `adjust_inverse_penetration`
2. `adjust_selection_rate`
3. `adjust_selection_rate2`
4. `adjust_raking_ratio`
5. `adjust_coefficient`
6. `adjust_multilevel_bayes` with `model_engine = "frequentist"` (temporary S1 placeholder for the Bayesian model)

It uses a fast but fair comparison protocol:

1. same stratified OD subset for all methods per replicate,
2. repeated over multiple seeds,
3. Method 6 fit with the frequentist development engine while S1-S4 Bayesian implementation is completed.

`adjust_multilevel_bayes()` is the main methodological innovation. Treat Method
6 results here as a fast scenario-contract check, not final Bayesian inference.
The example below uses S1 data: one MPD source observed during one time unit.

## Setup

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)
library(ggplot2)
library(forcats)
library(tidyr)
library(readr)
library(stringr)
library(purrr)

if (!requireNamespace("patchwork", quietly = TRUE)) {
  stop("Install 'patchwork' to render the multi-panel comparison plots.")
}
```

## Data

```{r}
data("simulated_mpd.od", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_coverage", package = "debiasR")
data("simulated_covariates", package = "debiasR")
data("simulated_distance", package = "debiasR")

tibble::tibble(
  dataset = c(
    "simulated_mpd.od",
    "simulated_benchmark.od",
    "simulated_coverage",
    "simulated_covariates",
    "simulated_distance"
  ),
  rows = c(
    nrow(simulated_mpd.od),
    nrow(simulated_benchmark.od),
    nrow(simulated_coverage),
    nrow(simulated_covariates),
    nrow(simulated_distance)
  )
)

# coverage_df schema used by methods in this walkthrough
names(simulated_coverage)
```

## Simulation Fidelity Check

```{r}
std_area <- function(x) stringr::str_squish(as.character(x))
std_source <- function(x) {
  x <- tolower(stringr::str_squish(as.character(x)))
  dplyr::case_when(
    stringr::str_detect(x, "facebook|fb") ~ "facebook",
    stringr::str_detect(x, "twitter| x ") ~ "twitter",
    stringr::str_detect(x, "multi.?app.?1") ~ "multiapp1",
    stringr::str_detect(x, "multi.?app.?2") ~ "multiapp2",
    TRUE ~ x
  )
}
safe_max <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  if (!any(is.finite(x))) return(NA_real_)
  max(x, na.rm = TRUE)
}

aup_raw <- readr::read_csv(file.path(pkg_root, "data-raw", "active-user-count_data.csv"), show_col_types = FALSE)
pop_raw <- readr::read_csv(file.path(pkg_root, "data-raw", "benchmark-population_data.csv"), show_col_types = FALSE)
mpd_raw <- readr::read_csv(file.path(pkg_root, "data-raw", "od_df_mar2020_feb2021.csv"), show_col_types = FALSE)
bench_raw <- readr::read_csv(file.path(pkg_root, "data-raw", "internal-migration-benchmark_data.csv"), show_col_types = FALSE)

users_wide <- aup_raw |>
  dplyr::transmute(
    area = std_area(.data$name),
    user_count_raw = suppressWarnings(as.numeric(.data$origin_users)),
    mpd_source = std_source(.data$source_mpd)
  ) |>
  dplyr::group_by(.data$area, .data$mpd_source) |>
  dplyr::summarise(user_count = safe_max(.data$user_count_raw), .groups = "drop") |>
  dplyr::filter(is.finite(.data$user_count), .data$user_count > 0)

pop_wide <- pop_raw |>
  dplyr::transmute(
    area = std_area(.data$lad_name),
    population_raw = suppressWarnings(as.numeric(.data$origin_population))
  ) |>
  dplyr::group_by(.data$area) |>
  dplyr::summarise(population = safe_max(.data$population_raw), .groups = "drop") |>
  dplyr::filter(is.finite(.data$population), .data$population > 0)

bench_norm <- bench_raw |>
  dplyr::transmute(
    origin = std_area(.data$lad_name_2020),
    destination = std_area(.data$lad_name_2021),
    flow = suppressWarnings(as.numeric(.data$Count))
  ) |>
  dplyr::filter(is.finite(.data$flow), .data$flow >= 0) |>
  dplyr::group_by(.data$origin, .data$destination) |>
  dplyr::summarise(flow = sum(.data$flow), .groups = "drop")

mpd_norm <- mpd_raw |>
  dplyr::transmute(
    origin = std_area(.data$origin),
    destination = std_area(.data$destination),
    flow = suppressWarnings(as.numeric(.data$total_flow)),
    mpd_source = std_source(.data$source_mpd)
  ) |>
  dplyr::filter(is.finite(.data$flow), .data$flow >= 0) |>
  dplyr::group_by(.data$origin, .data$destination, .data$mpd_source) |>
  dplyr::summarise(flow = sum(.data$flow), .groups = "drop")

top_source <- mpd_norm |>
  dplyr::group_by(.data$mpd_source) |>
  dplyr::summarise(total = sum(.data$flow), .groups = "drop") |>
  dplyr::arrange(dplyr::desc(.data$total)) |>
  dplyr::slice(1) |>
  dplyr::pull(.data$mpd_source)

S <- Reduce(
  intersect,
  list(
    unique(bench_norm$origin),
    unique(bench_norm$destination),
    unique(pop_wide$area),
    unique(users_wide$area)
  )
) |>
  sort()

bench_emp <- tidyr::crossing(origin = S, destination = S) |>
  dplyr::left_join(
    bench_norm |>
      dplyr::filter(.data$origin %in% S, .data$destination %in% S) |>
      dplyr::select(.data$origin, .data$destination, .data$flow),
    by = c("origin", "destination")
  ) |>
  dplyr::mutate(flow = tidyr::replace_na(.data$flow, 0)) |>
  dplyr::filter(.data$origin != .data$destination)

mpd_emp <- tidyr::crossing(origin = S, destination = S) |>
  dplyr::left_join(
    mpd_norm |>
      dplyr::filter(.data$origin %in% S, .data$destination %in% S, .data$mpd_source == top_source) |>
      dplyr::select(.data$origin, .data$destination, .data$flow),
    by = c("origin", "destination")
  ) |>
  dplyr::mutate(flow = tidyr::replace_na(.data$flow, 0)) |>
  dplyr::filter(.data$origin != .data$destination)

cov_emp <- pop_wide |>
  dplyr::filter(.data$area %in% S) |>
  dplyr::left_join(
    users_wide |>
      dplyr::filter(.data$mpd_source == top_source) |>
      dplyr::select(.data$area, .data$user_count),
    by = "area"
  ) |>
  dplyr::mutate(
    user_count = ifelse(is.finite(.data$user_count), .data$user_count, .data$population * 0.6),
    coverage_rate = .data$user_count / .data$population,
    coverage_bias = 1 - .data$coverage_rate
  )

cov_sim <- simulated_coverage |>
  dplyr::transmute(
    area = .data$origin,
    population = .data$population,
    user_count = .data$user_count,
    coverage_rate = .data$user_count / .data$population,
    coverage_bias = 1 - .data$coverage_rate
  )

compare_distribution <- function(x, y, label) {
  qx <- stats::quantile(x, probs = c(0.1, 0.5, 0.9), na.rm = TRUE)
  qy <- stats::quantile(y, probs = c(0.1, 0.5, 0.9), na.rm = TRUE)
  ks <- suppressWarnings(stats::ks.test(x, y))

  tibble::tibble(
    series = label,
    n_emp = length(x),
    n_sim = length(y),
    mean_emp = mean(x, na.rm = TRUE),
    mean_sim = mean(y, na.rm = TRUE),
    q10_emp = qx[[1]],
    q10_sim = qy[[1]],
    q50_emp = qx[[2]],
    q50_sim = qy[[2]],
    q90_emp = qx[[3]],
    q90_sim = qy[[3]],
    ks_D = as.numeric(ks$statistic),
    ks_p = ks$p.value
  )
}

fidelity_tbl <- dplyr::bind_rows(
  compare_distribution(log1p(bench_emp$flow), log1p(simulated_benchmark.od$flow), "benchmark_flow_log1p"),
  compare_distribution(log1p(mpd_emp$flow), log1p(simulated_mpd.od$flow), "mpd_flow_log1p"),
  compare_distribution(cov_emp$coverage_rate, cov_sim$coverage_rate, "coverage_rate"),
  compare_distribution(cov_emp$coverage_bias, cov_sim$coverage_bias, "coverage_bias")
)

fidelity_tbl
```

```{r}
plot_fidelity_ecdf <- function(emp, sim, xlab, title) {
  df <- dplyr::bind_rows(
    tibble::tibble(value = emp, source = "Empirical"),
    tibble::tibble(value = sim, source = "Simulated")
  )

  ggplot(df, aes(x = .data$value, color = .data$source)) +
    stat_ecdf(linewidth = 1) +
    labs(title = title, x = xlab, y = "ECDF", color = NULL)
}

p_fid_1 <- plot_fidelity_ecdf(
  log1p(bench_emp$flow),
  log1p(simulated_benchmark.od$flow),
  xlab = "log1p(flow)",
  title = "Benchmark Flow Distribution"
)

p_fid_2 <- plot_fidelity_ecdf(
  log1p(mpd_emp$flow),
  log1p(simulated_mpd.od$flow),
  xlab = "log1p(flow)",
  title = "MPD Flow Distribution"
)

p_fid_3 <- plot_fidelity_ecdf(
  cov_emp$coverage_bias,
  cov_sim$coverage_bias,
  xlab = "coverage bias = 1 - user_count/population",
  title = "Coverage-Bias Distribution"
)

patchwork::wrap_plots(p_fid_1, p_fid_2, p_fid_3, nrow = 1)
```

## Fair-Lite Settings

```{r}
n_sub <- 12000
fair_seeds <- c(123, 456, 789)

make_flow_strata <- function(flow_vec) {
  pos <- flow_vec[flow_vec > 0]
  q50 <- stats::quantile(pos, 0.50, na.rm = TRUE)
  q90 <- stats::quantile(pos, 0.90, na.rm = TRUE)

  dplyr::case_when(
    flow_vec <= 0 ~ "zero",
    flow_vec <= q50 ~ "low",
    flow_vec <= q90 ~ "mid",
    TRUE ~ "high"
  )
}

draw_fair_subset <- function(mpd_df, n_target, seed) {
  set.seed(seed)
  tmp <- mpd_df |>
    dplyr::mutate(stratum = make_flow_strata(.data$flow))

  counts <- tmp |>
    dplyr::count(.data$stratum, name = "n_stratum") |>
    dplyr::mutate(
      n_take = pmax(1L, round(n_target * .data$n_stratum / sum(.data$n_stratum)))
    )

  # Force exact sample size by adjusting the largest stratum.
  diff_n <- n_target - sum(counts$n_take)
  if (diff_n != 0) {
    idx <- which.max(counts$n_stratum)
    counts$n_take[idx] <- pmax(1L, counts$n_take[idx] + diff_n)
  }

  sampled <- lapply(counts$stratum, function(s) {
    df_s <- tmp[tmp$stratum == s, , drop = FALSE]
    n_take_s <- counts$n_take[counts$stratum == s][1]
    n_take_s <- min(n_take_s, nrow(df_s))
    df_s[sample.int(nrow(df_s), size = n_take_s), , drop = FALSE]
  }) |>
    dplyr::bind_rows()

  sampled |>
    dplyr::select(-.data$stratum)
}
```

## Validation functions

```{r}
collect_metrics <- function(method_name, adj_df, benchmark_df, seed_id) {
  v <- validate_flow_overall(
    adj_df = adj_df,
    benchmark_od_df = benchmark_df,
    method_name = method_name,
    return_joined = FALSE
  )

  tibble::tibble(
    seed = seed_id,
    method = v$method,
    n = v$n,
    pearson_r = v$pearson_r,
    spearman_rho = v$spearman_rho,
    rmse = v$rmse,
    mae = v$mae,
    mape = v$mape,
    r_squared = v$r_squared
  )
}

build_compare_df <- function(adj_df, mpd_df, benchmark_df) {
  adj_df |>
    dplyr::select(origin, destination, flow_adj) |>
    dplyr::inner_join(
      mpd_df |>
        dplyr::select(origin, destination, flow_mpd = flow),
      by = c("origin", "destination")
    ) |>
    dplyr::inner_join(
      benchmark_df |>
        dplyr::select(origin, destination, flow_bench = flow),
      by = c("origin", "destination")
    )
}

plot_method_triptych <- function(method_label, adj_df, mpd_df, benchmark_df) {
  df <- build_compare_df(adj_df, mpd_df = mpd_df, benchmark_df = benchmark_df)

  p1 <- ggplot(df, aes(x = flow_mpd, y = flow_adj)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#2B8CBE") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "Adjusted vs MPD",
      x = "Mobile-phone-derived flow",
      y = "Adjusted flow"
    )

  p2 <- ggplot(df, aes(x = flow_bench, y = flow_adj)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#41AE76") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "Adjusted vs Benchmark",
      x = "Benchmark flow",
      y = "Adjusted flow"
    )

  p3 <- ggplot(df, aes(x = flow_bench, y = flow_mpd)) +
    geom_point(alpha = 0.2, size = 0.8, color = "#756BB1") +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#D95F0E") +
    labs(
      title = "MPD vs Benchmark",
      x = "Benchmark flow",
      y = "Mobile-phone-derived flow"
    )

  patchwork::wrap_plots(p1, p2, p3, nrow = 1) +
    patchwork::plot_annotation(title = method_label)
}
```

## Run Methods (Fair-Lite)

```{r}
run_methods_once <- function(seed_id) {
  mpd_sub <- draw_fair_subset(simulated_mpd.od, n_target = n_sub, seed = seed_id)
  benchmark_sub <- simulated_benchmark.od |>
    dplyr::inner_join(
      mpd_sub |> dplyr::select(origin, destination),
      by = c("origin", "destination")
    )

  distance_sub <- simulated_distance |>
    dplyr::inner_join(
      mpd_sub |> dplyr::select(origin, destination),
      by = c("origin", "destination")
    )

  adj_m1 <- adjust_inverse_penetration(
    mpd_od_df = mpd_sub,
    coverage_df = simulated_coverage,
    weight_by = "both"
  )

  adj_m2 <- adjust_selection_rate(
    mpd_od_df = mpd_sub,
    coverage_df = simulated_coverage,
    covariates_df = simulated_covariates,
    income_col = "income_norm",
    weight_by = "origin",
    benchmark_od_df = benchmark_sub,
    calibration_aggregate = "origin"
  )

  adj_m3 <- adjust_selection_rate2(
    mpd_od_df = mpd_sub,
    coverage_df = simulated_coverage,
    weight_by = "origin",
    benchmark_od_df = benchmark_sub,
    calibration_aggregate = "origin"
  )

  adj_m4 <- adjust_raking_ratio(
    mpd_od_df = mpd_sub,
    benchmark_od_df = benchmark_sub
  )

  adj_m5 <- adjust_coefficient(
    mpd_od_df = mpd_sub,
    benchmark_od_df = benchmark_sub,
    model_family = "ols",
    level = "od"
  )

  mpd_s1 <- mpd_sub |>
    dplyr::mutate(
      mpd_source = "operator_a",
      mpd_time = "2021_q1"
    )
  coverage_s1 <- simulated_coverage |>
    dplyr::mutate(
      mpd_source = "operator_a",
      mpd_time = "2021_q1"
    )

  adj_m6 <- adjust_multilevel_bayes(
    mpd_od_df = mpd_s1,
    coverage_df = coverage_s1,
    covariates_df = simulated_covariates,
    distance_df = distance_sub,
    formula = flow ~ income_norm_o + income_norm_d + log_distance + bias_e_origin,
    model_engine = "frequentist",
    scenario = "s1",
    source_col = "mpd_source",
    time_col = "mpd_time",
    repeated_observation = "none",
    random_intercept = "none",
    model_family = "poisson",
    flow_adj_summary = "median"
  )

  metrics <- dplyr::bind_rows(
    collect_metrics("adjust_inverse_penetration", adj_m1, benchmark_sub, seed_id),
    collect_metrics("adjust_selection_rate", adj_m2, benchmark_sub, seed_id),
    collect_metrics("adjust_selection_rate2", adj_m3, benchmark_sub, seed_id),
    collect_metrics("adjust_raking_ratio", adj_m4, benchmark_sub, seed_id),
    collect_metrics("adjust_coefficient", adj_m5, benchmark_sub, seed_id),
    collect_metrics("adjust_multilevel_bayes", adj_m6, benchmark_sub, seed_id)
  )

  list(
    mpd_sub = mpd_sub,
    benchmark_sub = benchmark_sub,
    adj = list(
      adjust_inverse_penetration = adj_m1,
      adjust_selection_rate = adj_m2,
      adjust_selection_rate2 = adj_m3,
      adjust_raking_ratio = adj_m4,
      adjust_coefficient = adj_m5,
      adjust_multilevel_bayes = adj_m6
    ),
    metrics = metrics
  )
}

res_list <- purrr::map(fair_seeds, run_methods_once)
metrics_tbl <- dplyr::bind_rows(purrr::map(res_list, "metrics"))

metrics_summary <- metrics_tbl |>
  dplyr::group_by(.data$method) |>
  dplyr::summarise(
    n_mean = mean(.data$n),
    pearson_r_mean = mean(.data$pearson_r),
    pearson_r_sd = stats::sd(.data$pearson_r),
    rmse_mean = mean(.data$rmse),
    rmse_sd = stats::sd(.data$rmse),
    r_squared_mean = mean(.data$r_squared),
    r_squared_sd = stats::sd(.data$r_squared),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(.data$pearson_r_mean))

metrics_summary
```

## Replicate-Level Metrics

```{r}
metrics_tbl
```

## Scatterplot Rows By Method (Seed 123)

```{r}
vis <- res_list[[1]]
```

```{r}
plot_method_triptych(
  "adjust_inverse_penetration",
  vis$adj$adjust_inverse_penetration,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_selection_rate`

```{r}
plot_method_triptych(
  "adjust_selection_rate",
  vis$adj$adjust_selection_rate,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_selection_rate2`

```{r}
plot_method_triptych(
  "adjust_selection_rate2",
  vis$adj$adjust_selection_rate2,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_raking_ratio`

```{r}
plot_method_triptych(
  "adjust_raking_ratio",
  vis$adj$adjust_raking_ratio,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_coefficient`

```{r}
plot_method_triptych(
  "adjust_coefficient",
  vis$adj$adjust_coefficient,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

### `adjust_multilevel_bayes` frequentist placeholder

```{r}
plot_method_triptych(
  "adjust_multilevel_bayes (frequentist placeholder)",
  vis$adj$adjust_multilevel_bayes,
  mpd_df = vis$mpd_sub,
  benchmark_df = vis$benchmark_sub
)
```

## Plot Comparison (Mean Across Seeds)

```{r}
metrics_summary |>
  dplyr::mutate(method = forcats::fct_reorder(method, pearson_r_mean)) |>
  ggplot(aes(x = method, y = pearson_r_mean)) +
  geom_col(fill = "#2B8CBE", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Pearson Correlation by Method (Mean Across Seeds)",
    x = "Method",
    y = "Pearson r"
  )
```

```{r}
metrics_summary |>
  dplyr::mutate(method = forcats::fct_reorder(method, r_squared_mean)) |>
  ggplot(aes(x = method, y = r_squared_mean)) +
  geom_col(fill = "#41AE76", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "R-squared by Method (Mean Across Seeds)",
    x = "Method",
    y = "R-squared"
  )
```

## Inspect Method 6 Attributes (Seed 123)

```{r}
m6_vis <- vis$adj$adjust_multilevel_bayes
head(attr(m6_vis, "coefficients"))
attr(m6_vis, "formula")
attr(m6_vis, "backend")
attr(m6_vis, "model_engine")
attr(m6_vis, "scenario")
attr(m6_vis, "model_family")
```

The example above is S1: one source and one time unit. For repeated
observations, keep the same function but change `scenario`,
`source_col`, `time_col`, and `repeated_observation`: use `s2` and
`repeated_observation = "time"` for one source over multiple periods, `s3` and
`"source"` for multiple sources in one period, and `s4` plus `"source_time"`
when both source and time vary.

## Render

```bash
quarto render vignettes/method-comparison.qmd
```
