---
title: "Frequentist Placeholder for Bayesian Multilevel Adjustment"
format: html
editor: visual
---

# Purpose

This walkthrough uses `adjust_multilevel_bayes()` with
`model_engine = "frequentist"` as a fast placeholder while the full S1-S4
Bayesian implementation is completed. The fitted example is S1: one
mobile-phone data source observed during one time unit.

# Package and data

```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

data(simulated_mpd.od, package = "debiasR")
data(simulated_benchmark.od, package = "debiasR")
data(simulated_coverage, package = "debiasR")
data(simulated_covariates, package = "debiasR")
data(simulated_distance, package = "debiasR")
```

# Build an S1 source/time structure

S1 means one source and one time unit. The source and time columns can be absent
for S1, but adding constant columns makes the scenario contract visible and
keeps the example close to the S2-S4 data structure.

```{r}
example_mpd_od <- simulated_mpd.od |>
  mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

example_coverage <- simulated_coverage |>
  mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

example_mpd_od |>
  select(origin, destination, mpd_source, mpd_time, flow) |>
  head(5)
```

# Fit the frequentist placeholder

```{r}
adj <- adjust_multilevel_bayes(
  mpd_od_df = example_mpd_od,
  coverage_df = example_coverage,
  covariates_df = simulated_covariates,
  distance_df = simulated_distance,
  formula = flow ~ income_norm_o + income_norm_d + log_distance + bias_e_origin,
  model_engine = "frequentist",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median"
)

attr(adj, "result_metadata")[c(
  "model_engine",
  "scenario",
  "repeated_observation",
  "n_sources",
  "n_time_periods"
)]
```

The placeholder uses a fast GLM/GLMM engine to check the formula and output
contract. It is not a substitute for final Bayesian inference.

```{r}
attr(adj, "model_terms")
attr(adj, "coefficients") |>
  select(any_of(c("term", "mean", "sd", "z value", "Pr(>|z|)"))) |>
  head(10)
```

# Compare raw vs adjusted against the benchmark

```{r}
raw_as_adj <- example_mpd_od |>
  transmute(origin, destination, flow_adj = flow)

val_raw <- validate_flow_overall(
  adj_df = raw_as_adj,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "raw_mpd",
  return_joined = FALSE
)

val_adj <- validate_flow_overall(
  adj_df = adj,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_multilevel_bayes_frequentist_placeholder",
  return_joined = FALSE
)

tibble(
  method = c(val_raw$method, val_adj$method),
  pearson_r = c(val_raw$pearson_r, val_adj$pearson_r),
  rmse = c(val_raw$rmse, val_adj$rmse),
  mae = c(val_raw$mae, val_adj$mae),
  r_squared = c(val_raw$r_squared, val_adj$r_squared)
)
```

# Understand row-level changes

```{r}
adj |>
  mutate(
    delta = .data$flow_adj - .data$flow,
    pct_change = 100 * (.data$flow_adj - .data$flow) / pmax(.data$flow, 1e-8)
  ) |>
  select(
    origin,
    destination,
    mpd_source,
    mpd_time,
    flow,
    flow_adj,
    bias_e_origin,
    log_distance,
    delta,
    pct_change
  ) |>
  arrange(desc(abs(.data$delta))) |>
  head(15)
```

Large positive `delta` means the placeholder model increases the MPD flow for
that OD pair. Large negative `delta` means it decreases the MPD flow. The
`bias_e_origin` column shows the origin-side coverage-bias term used by the
model.

# Parameters for alternative data structures

The same function handles the four source/time scenarios by changing the
scenario metadata and the repeated-observation structure.

```{r}
tibble::tribble(
  ~Scenario, ~Data_structure, ~Required_columns, ~Key_parameters,
  "S1", "one source, one time", "source/time columns optional or constant", "scenario = 's1', repeated_observation = 'none'",
  "S2", "one source, multiple times", "time column", "scenario = 's2', time_col = '<time>', repeated_observation = 'time'",
  "S3", "multiple sources, one time", "source column", "scenario = 's3', source_col = '<source>', repeated_observation = 'source'",
  "S4", "multiple sources, multiple times", "source and time columns", "scenario = 's4', source_col = '<source>', time_col = '<time>', repeated_observation = 'source_time'"
)
```

Use `model_engine = "frequentist"` for these S1-S4 examples until the complete
Bayesian scenario implementation is transferred and validated.
