---
title: "debiasR: Simulated Methods Walkthrough"
format:
  html:
    toc: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This walkthrough runs and visualises the implemented `debiasR` correction methods using simulated package datasets.
The current simulated MPD data are calibrated to empirical OD support to match
zero-mass and upper-tail quantiles more closely.

## Setup
```{r}
candidate_roots <- c(".", "..", "../..")
pkg_root <- candidate_roots[file.exists(file.path(candidate_roots, "DESCRIPTION"))][1]
if (is.na(pkg_root)) {
  stop("Could not locate the debiasR package root.")
}

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(tidyverse)

theme_path <- file.path(pkg_root, "style", "data-visualisation_theme.R")
if (file.exists(theme_path)) {
  source(theme_path)
}

"validate_flow_overall" %in% getNamespaceExports("debiasR")
```

```{r}
data("simulated_mpd.od", package = "debiasR")
data("simulated_benchmark.od", package = "debiasR")
data("simulated_coverage", package = "debiasR")
data("simulated_covariates", package = "debiasR")
data("simulated_distance", package = "debiasR")
data("simulated_active.users", package = "debiasR")
data("simulated_pop", package = "debiasR")

tibble(
  dataset = c(
    "simulated_mpd.od",
    "simulated_benchmark.od",
    "simulated_coverage",
    "simulated_covariates",
    "simulated_distance",
    "simulated_active.users",
    "simulated_pop"
  ),
  rows = c(
    nrow(simulated_mpd.od),
    nrow(simulated_benchmark.od),
    nrow(simulated_coverage),
    nrow(simulated_covariates),
    nrow(simulated_distance),
    nrow(simulated_active.users),
    nrow(simulated_pop)
  )
)
```

## Simulated Distribution Check

```{r}
tibble(
  series = c("benchmark_flow", "mpd_flow"),
  zero_share = c(
    mean(simulated_benchmark.od$flow == 0),
    mean(simulated_mpd.od$flow == 0)
  ),
  q50_log1p = c(
    quantile(log1p(simulated_benchmark.od$flow), 0.5),
    quantile(log1p(simulated_mpd.od$flow), 0.5)
  ),
  q90_log1p = c(
    quantile(log1p(simulated_benchmark.od$flow), 0.9),
    quantile(log1p(simulated_mpd.od$flow), 0.9)
  ),
  q99_log1p = c(
    quantile(log1p(simulated_benchmark.od$flow), 0.99),
    quantile(log1p(simulated_mpd.od$flow), 0.99)
  )
)
```

## Plot Helpers

```{r}
plot_origin_totals <- function(df, flow_col = "flow", title_text) {
  df |>
    group_by(origin) |>
    summarise(total_flow = sum(.data[[flow_col]], na.rm = TRUE), .groups = "drop") |>
    ggplot(aes(x = reorder(origin, total_flow), y = total_flow)) +
    geom_col(fill = "#121282", alpha = 0.9) +
    coord_flip() +
    labs(
      title = title_text,
      x = "Origin",
      y = "Total flow"
    ) +
    theme_plot_tufte()
}

plot_adj_vs_benchmark <- function(adj_df, benchmark_df, title_text) {
  cmp <- adj_df |>
    select(origin, destination, flow_adj) |>
    inner_join(
      benchmark_df |>
        select(origin, destination, flow_bench = flow),
      by = c("origin", "destination")
    )

  ggplot(cmp, aes(x = flow_bench, y = flow_adj)) +
    geom_point(color = "#121282", alpha = 0.75, size = 2) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "#FE7105") +
    labs(
      title = title_text,
      x = "Benchmark flow",
      y = "Adjusted flow"
    ) +
    theme_plot_tufte()
}

collect_metrics <- function(method_name, adj_df) {
  v <- validate_flow_overall(
    adj_df = adj_df,
    benchmark_od_df = simulated_benchmark.od,
    method_name = method_name,
    return_joined = FALSE
  )

  tibble(
    method = v$method,
    n = v$n,
    pearson_r = v$pearson_r,
    spearman_rho = v$spearman_rho,
    rmse = v$rmse,
    mae = v$mae,
    mape = v$mape,
    r_squared = v$r_squared
  )
}
```

## Baseline Visuals

```{r}
plot_origin_totals(simulated_mpd.od, "flow", "Observed MPD Flows by Origin")
```

```{r}
plot_origin_totals(simulated_benchmark.od, "flow", "Benchmark Flows by Origin")
```

## 1) `adjust_inverse_penetration`

```{r}
adj_m1 <- adjust_inverse_penetration(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  weight_by = "both"
)

collect_metrics("adjust_inverse_penetration", adj_m1)
```

```{r}
plot_adj_vs_benchmark(adj_m1, simulated_benchmark.od, "Method 1: Adjusted vs Benchmark")
```

## 2) `adjust_selection_rate`

```{r}
adj_m2 <- adjust_selection_rate(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  covariates_df = simulated_covariates,
  income_col = "income_norm",
  weight_by = "origin",
  benchmark_od_df = simulated_benchmark.od,
  calibration_aggregate = "origin"
)

list(r_global = attr(adj_m2, "r_global"))
```

```{r}
plot_adj_vs_benchmark(adj_m2, simulated_benchmark.od, "Method 2: Adjusted vs Benchmark")
```

## 3) `adjust_selection_rate2`

```{r}
adj_m3 <- adjust_selection_rate2(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  weight_by = "origin",
  benchmark_od_df = simulated_benchmark.od,
  calibration_aggregate = "origin"
)

list(k = attr(adj_m3, "k"))
```

```{r}
plot_adj_vs_benchmark(adj_m3, simulated_benchmark.od, "Method 3: Adjusted vs Benchmark")
```

## 4) `adjust_raking_ratio`

```{r}
adj_m4 <- adjust_raking_ratio(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od
)

list(
  ipf_converged = attr(adj_m4, "ipf_converged"),
  ipf_iterations = attr(adj_m4, "ipf_iterations")
)
```

```{r}
plot_adj_vs_benchmark(adj_m4, simulated_benchmark.od, "Method 4: Adjusted vs Benchmark")
```

## 5) `adjust_coefficient`

```{r}
adj_m5 <- adjust_coefficient(
  mpd_od_df = simulated_mpd.od,
  benchmark_od_df = simulated_benchmark.od,
  model_family = "ols",
  level = "od"
)

list(coef = attr(adj_m5, "coef"))
```


```{r}
plot_adj_vs_benchmark(adj_m5, simulated_benchmark.od, "Method 5: Adjusted vs Benchmark")
```

## 6) `adjust_multilevel_bayes` (frequentist placeholder)

While the full S1-S4 Bayesian implementation is being completed, this example
uses the frequentist development engine as a fast placeholder. The data below
represent S1: one MPD source observed in one time period.

```{r}
mpd_s1 <- simulated_mpd.od |>
  mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

coverage_s1 <- simulated_coverage |>
  mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

adj_m6 <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = simulated_covariates,
  distance_df = simulated_distance,
  formula = flow ~ income_norm_o + income_norm_d + log_distance + bias_e_origin,
  model_engine = "frequentist",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median"
)

attr(adj_m6, "result_metadata")[c("model_engine", "scenario", "repeated_observation")]
attr(adj_m6, "model_terms")
```

```{r}
plot_adj_vs_benchmark(adj_m6, simulated_benchmark.od, "Method 6: Adjusted vs Benchmark")
```

Use the same function for other repeated-observation structures by changing the
scenario metadata:

```{r}
tibble::tribble(
  ~scenario, ~data_structure, ~source_col, ~time_col, ~repeated_observation,
  "s1", "one source, one time", "optional constant", "optional constant", "none",
  "s2", "one source, multiple times", "optional constant", "time column", "time",
  "s3", "multiple sources, one time", "source column", "optional constant", "source",
  "s4", "multiple sources, multiple times", "source column", "time column", "source_time"
)
```

## Compare Validation Metrics Across Methods

```{r}
metrics_tbl <- bind_rows(
  collect_metrics("adjust_inverse_penetration", adj_m1),
  collect_metrics("adjust_selection_rate", adj_m2),
  collect_metrics("adjust_selection_rate2", adj_m3),
  collect_metrics("adjust_raking_ratio", adj_m4),
  collect_metrics("adjust_coefficient", adj_m5),
  collect_metrics("adjust_multilevel_bayes_frequentist_placeholder", adj_m6)
) |>
  arrange(desc(pearson_r))

metrics_tbl
```

```{r}
metrics_tbl |>
  mutate(method = fct_reorder(method, pearson_r)) |>
  ggplot(aes(x = method, y = pearson_r)) +
  geom_col(fill = "#41AE76", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Pearson Correlation by Method",
    x = "Method",
    y = "Pearson r"
  ) +
  theme_plot_tufte()
```

## Validation Helper Examples

### `validate_flow_overall`

```{r}
val_benchmark_m1 <- validate_flow_overall(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_inverse_penetration",
  return_joined = FALSE
)

val_benchmark_m1
```

### `validate_flow_pairs`

```{r}
val_all_m1 <- validate_flow_pairs(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od
)

val_all_m1 |>
  select(origin, destination, mpd_flow, adj_flow, benchmark_flow, diff_mpd_adj, diff_adj_benchmark) |>
  slice_head(n = 10)
```

### `validate_flow_residuals`

```{r}
residual_m1 <- validate_flow_residuals(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_inverse_penetration"
)

residual_m1$summary
```

### `validate_flow_residual_structure`

```{r}
structure_m1 <- validate_flow_residual_structure(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_inverse_penetration",
  covariate_df = simulated_covariates,
  covariate_col = "income_norm"
)

structure_m1$summary
```

### `validate_flow_distribution`

```{r}
distribution_m1 <- validate_flow_distribution(
  adj_df = adj_m1,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_inverse_penetration",
  weight_by = "benchmark_origin_total"
)

distribution_m1$summary
```

## `measure_bias` Example

```{r}
coverage_for_bias <- simulated_coverage |>
  select(origin, population, user_count)

bias_df <- measure_bias(coverage_for_bias)
bias_df
```

```{r}
bias_df |>
  ggplot(aes(x = reorder(origin, coverage_score), y = coverage_score)) +
  geom_col(fill = "#2B8CBE", alpha = 0.9) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "#D95F0E") +
  coord_flip() +
  labs(
    title = "Coverage Score by Origin",
    x = "Origin",
    y = "coverage_score = user_count / population"
  ) +
  theme_plot_tufte()
```

## Stage 3 Bias Residual Diagnostics

```{r}
areas <- sort(unique(simulated_coverage$origin))

area_neighbors <- tibble(
  area = areas,
  neighbor = dplyr::lead(areas, default = areas[1])
) |>
  bind_rows(tibble(
    area = areas,
    neighbor = dplyr::lag(areas, default = areas[length(areas)])
  ))

bias_structure <- validate_bias_residual_structure(
  coverage_df = simulated_coverage,
  benchmark_od_df = simulated_benchmark.od,
  area_neighbors = area_neighbors,
  covariate_df = simulated_covariates,
  covariate_col = "internet_access"
)

bias_structure$summary
```

```{r}
bias_structure$area_level |>
  select(
    area,
    population,
    user_count,
    coverage_score,
    expected_user_count,
    user_count_residual,
    coverage_score_residual,
    selected_residual
  ) |>
  arrange(desc(abs(selected_residual))) |>
  slice_head(n = 10)
```

```{r}
bias_structure$benchmark_flow_correlation
```

## Render

```bash
quarto render vignettes/simulated-methods-walkthrough.qmd
```
