---
title: "Bayesian crash"
format: html
editor: visual
---

# Package and data
```{r}
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."

if (!requireNamespace("debiasR", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    stop("Install either `debiasR` or `devtools` to load the local package.")
  }
  devtools::load_all(pkg_root, quiet = TRUE)
}

library(debiasR)
library(dplyr)

has_rstanarm <- requireNamespace("rstanarm", quietly = TRUE)
if (!has_rstanarm) {
  cat("This vignette requires the optional 'rstanarm' package. Install it to run the Bayesian walkthrough.\n")
  knitr::knit_exit()
}

library(rstanarm)

data(simulated_mpd.od)
data(simulated_coverage)
data(simulated_covariates)
data(simulated_distance)
data(simulated_benchmark.od)

names(simulated_coverage)

tibble::tibble(
  series = c("benchmark_flow", "mpd_flow"),
  zero_share = c(
    mean(simulated_benchmark.od$flow == 0),
    mean(simulated_mpd.od$flow == 0)
  ),
  q90_log1p = c(
    quantile(log1p(simulated_benchmark.od$flow), 0.9),
    quantile(log1p(simulated_mpd.od$flow), 0.9)
  ),
  q99_log1p = c(
    quantile(log1p(simulated_benchmark.od$flow), 0.99),
    quantile(log1p(simulated_mpd.od$flow), 0.99)
  )
)
```



Run model 
```{r}
adj <- adjust_multilevel_bayes(
  mpd_od_df = simulated_mpd.od,
  coverage_df = simulated_coverage,
  covariates_df = simulated_covariates,
  distance_df = simulated_distance,
  random_intercept = "origin",
  model_family = "poisson",
  flow_adj_summary = "median",
  iter = 500,
  chains = 2,
  seed = 123,
  refresh = 0
)

m <- attr(adj, "model")
coef_tbl <- attr(adj, "coefficients")

```
# Read posterior summaries (fixed effects)

```{r}
coef_tbl %>%
  select(any_of(c("term", "mean", "sd", "q2.5", "q97.5"))) %>%
  print(n = Inf)

```

> Interpretation:

- mean: posterior central estimate.
- q2.5 to q97.5: 95% credible interval.
- For bias_e_origin (omega): if interval mostly above 0, higher bias term is associated with higher observed flow on log scale (conditional on other terms); if below 0, opposite.

#  Direction probabilities from posterior draws
```{r}
draws <- as.matrix(m)

# Helper to robustly pick a parameter by pattern
pick <- function(pattern) {
  nm <- grep(pattern, colnames(draws), value = TRUE)
  nm[1]
}

p_bias <- pick("^bias_e_origin$")
p_dist <- pick("^log_distance$")
p_inc_o <- pick("^income_o$")
p_inc_d <- pick("^income_d$")

c(
  Pr_bias_gt_0 = mean(draws[, p_bias] > 0),
  Pr_dist_lt_0 = mean(draws[, p_dist] < 0),
  Pr_inc_o_gt_0 = mean(draws[, p_inc_o] > 0),
  Pr_inc_d_gt_0 = mean(draws[, p_inc_d] > 0)
)

```

> Interpretation:

These are posterior probabilities of effect direction.
Example: Pr_dist_lt_0 = 0.98 means strong evidence that distance reduces flows.

# Posterior predictive uncertainty
```{r}
# Posterior predictive draws for observed-data scale counts
pp <- rstanarm::posterior_predict(m)  # draws x N

pred_med <- apply(pp, 2, median)
pred_lo  <- apply(pp, 2, quantile, 0.05)
pred_hi  <- apply(pp, 2, quantile, 0.95)

obs <- model.response(model.frame(m))

pred_check <- tibble(
  obs = obs,
  pred_med = pred_med,
  pred_lo = pred_lo,
  pred_hi = pred_hi
)

head(pred_check, 10)
mean(pred_check$obs >= pred_check$pred_lo & pred_check$obs <= pred_check$pred_hi)

```
> Interpretation:

pred_lo/pred_hi are uncertainty bands for observed counts.
Coverage proportion gives quick calibration feel (not formal validation alone).

# Compare raw vs adjusted against benchmark
```{r}
raw_as_adj <- simulated_mpd.od %>%
  transmute(origin, destination, flow_adj = flow)

val_raw <- debiasR::validate_flow_overall(
  adj_df = raw_as_adj,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "raw_mpd",
  return_joined = FALSE
)

val_adj <- debiasR::validate_flow_overall(
  adj_df = adj,
  benchmark_od_df = simulated_benchmark.od,
  method_name = "adjust_multilevel_bayes",
  return_joined = FALSE
)


```

```{r}
tibble(
  method = c("raw_mpd", "adjust_multilevel_bayes"),
  pearson_r = c(val_raw$pearson_r, val_adj$pearson_r),
  rmse = c(val_raw$rmse, val_adj$rmse),
  mae = c(val_raw$mae, val_adj$mae),
  r_squared = c(val_raw$r_squared, val_adj$r_squared)
)

```

> Interpretation:

If adjusted method is working, you typically want higher correlation and lower error than raw MPD.

# Understand what flow_adj is doing row-by-row
```{r}
adj %>%
  mutate(delta = flow_adj - flow,
         pct_change = 100 * (flow_adj - flow) / pmax(flow, 1e-8)) %>%
  select(origin, destination, flow, flow_adj, bias_e_origin, log_distance, delta, pct_change) %>%
  arrange(desc(abs(delta))) %>%
  head(15)

```

> Interpretation:

Large positive delta: model believes MPD undercounts that OD pair.
Large negative delta: model believes MPD overcounts that OD pair.
bias_e_origin helps explain where correction pressure comes from.
